### 2011-09-20 soulgem.tex
First commit to GitHub

\documentclass[10pt, conference, compsocconf]{IEEEtran}

\begin{document}
\title{The Soulgem Paper}

\author{\IEEEauthorblockN{Jorge Aranda, Adrian Schr\"{o}ter, Daniela Damian, and Margaret-Anne Storey}
\IEEEauthorblockA{Department of Computer Science\\
University of Victoria\\
Victoria, Canada\\
Email: \{jaranda, schadr, danielad, mstorey\}@uvic.ca}
}

\maketitle

\begin{abstract}

I'm leaving the abstract until later.

\end{abstract}

\begin{IEEEkeywords}
pink robots;
\end{IEEEkeywords}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

People develop software, but the activities that each of them perform to do so differ from those of their colleagues: they divide their labour. For instance, some may write production code, others test it, and still others try to ensure that the product satisfies a real need. In order to talk about this division of labour, we abstract their responsibilities into a number of \emph{roles} (such as ``developer,'' ``tester,'' and ``product manager'') and we use these abstractions as labels to refer to the set of activities that we expect them to carry out. These labels appear everywhere in our academic literature, as well as in books for practitioners, human resources postings, blog posts, and everyday conversation.

Despite this commonplace use, we have no clear indication that the people using these labels use them to refer to the same meanings. For instance, what does a ``developer'' actually need to do in order to fulfill her responsibilities? What should she avoid doing? When should we judge that she has failed to fulfill her responsibilities, and would the answer be generalizable throughout the many domains in which the software industry functions?

In more mature domains, especially in those of critical importance, the division of labour is fairly formal and clear. Surgeons, anesthesiologists, nurses, and patients generally know what to expect from each other, in and out of the operating room. Knowing that someone is ``a nurse'' will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board. In our own community we use role labels just as directly and unthinkingly as in the medical domain, but without realizing (or such is the premise of our paper) that we may be talking about very different things, based on our varying assumptions, experiences, and cultures.

This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has unexplored repercussions for researchers and practitioners alike. Researchers may assume coordination dynamics between, for instance, analysts, designers, and developers that have no connection with the interactions between those groups in reality. Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him. For an abstraction as heavily used as role definitions in software organization, a closer look should prove beneficial.

In this paper we did not attempt to provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success. Instead, (a) we examine the concepts of roles, positions, and expectations in software development groups, providing a theoretical framework based on sociological and organizational research, (b) we report the findings from a case study at two software organizations in which we used this theoretical framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Theoretical Framework}

Sociologists and organizational scientists have studied the concept of roles in social groups over decades. We take their findings as the starting points for the discussion below, acknowledging them where appropriate.

\subsection{Roles, positions, and labels}

A \textbf{\emph{role}} is a set of expectations of behaviour placed on a person given their position, status, history, or preference \cite{Turner1956,Gordon1976,Rizzo1970}. This entails that roles are \emph{social constructions}: in a real sense, a ``developer'' is whatever the relevant social group determines that a developer should be. There is no archetypal developer. Upon learning that a colleague is a developer, all the members of the organization place a number of expectations on said colleague. Perhaps they expect the developer to spend as much time as possible writing production code, to be responsive to user requests for assistance, to validate the code produced by other developers, or to self-manage. Some colleagues may have no expectations other than that the developer should strive for the well-being of the organization. Other colleagues, especially if they interact with the developer on a regular basis, may have many more expectations, and these may be far more concrete. People outside the formal organization, but with connections or interactions to it, may hold other expectations on her behaviour. And the developer herself, the target of all these expectations, likely has a number of additional expectations on her own conduct: the activities she feels she should be doing to play her part in the organization.

These expectations, taken together as a collection, conform the role. The resulting set may be straightforward or conflictive (a phenomenon called \emph{role strain} \cite{Goode1960}); it may even be impossible to satisfy in full. And since the expectations arise from the members of the group, there is no guarantee that they will match those of other groups: the meaning of a ``developer'' role in one organization may differ from that in another organization, and even within an organization the semantic differences may be profound.

A \textbf{\emph{position}} or a \textbf{\emph{title}}, in turn, is an official statement of a person's goals and responsibilities. A role is not a position \cite{Turner1956}, though they often carry the same labels: one can play an architectural role and hold the position of Software Architect. A person's title plays a part in determining what she will feel compelled to do, and what she believes she is expected to do, but it is only one element of her role. Roles are enacted, positions are occupied (several people could fulfill the same role, despite their different positions; a person can fulfill several roles while holding a single position). For instance, one could carry an architectural role without holding an Software Architect position, and one could hold a Software Architect position without performing any of the activities that others would expect from an architect.

We can see the positions in an organization, and their accompanying lists of responsibilities, as the explicit manifestations of what the fraction of the members of the organization that have the power to set those positions and responsibilities expect from their colleagues. In that sense they are a part of the total set of expectations placed upon organizational roles; a powerful part (especially if a position's responsibilities are tied to incentives and job evaluations), but still only a part.

As indicated above, we often use \textbf{\emph{labels}} to refer to roles. A fitting label helps us convey the essence of the role to our interlocutors. But as we will see some roles have ill-fitting labels, and others have no labels or easy ways to refer to them in concrete terms (in many organizations there are people who are expected to ``keep the place together,'' for example). We must then remember that neither the label nor the formal position make the role.

\subsection{Roles and culture}

Of course, the larger culture plays an important part in setting and steering our expectations. In some domains the culture is sufficiently strong to establish an interpretation of a role throughout the population. As we mention in the Introduction, in the medical domain, a combination of intensive training, codes of ethics, professional societies, and familiarization in person and through popular culture leads us to expect certain behaviours and activities from a doctor, and it leads the doctor to become attuned to the expectations placed upon her. We find it shocking if a doctor deviates from these expectations significantly. In our own domain, a similar combination of forces (education, past experiences, popular culture) allows us to make some general assumptions about software development roles that we can reasonably expect will hold in most cases. For example, we would be surprised if a tester did not spend at least part of his time preparing or running tests. And so, thanks to this larger culture, a newcomer to the test department will likely assume that he is expected to prepare and run tests, and his peers and supervisors will indeed have these expectations placed upon him, and this is what he will spend his time doing, thereby reinforcing the general idea of what testers do.

However, in our domain, the forces from the larger culture have not yet enforced a singular view of what the different roles in a software organization consist of in any useful detail, or even which roles exist. As the evidence of our case study shows below, significant disagreements exist even within the same software organizations. Determining the reasons for this disagreement is beyond the scope of this paper; we speculate that the relative immaturity of the software industry, the wide variety of fields in which it functions, and the low barriers of entry to the profession all contribute to the phenomenon. But whatever the reason, the lack of conceptual uniformity in software development roles has implications for both research and practice, which we discuss in the coming pages.

\subsection{Dysfunction and role fluctuation}

In theory, we can use the definition of roles as sets of expectations of behaviour to build a conceptual map of roles in an organization: we find what every member of the organization expects from everyone else, and the aggregate of these expectations determines the roles that all members enact. But the members of the organization do not always act according to the expectations placed on them. The disagreement between expectation and actual behaviour is a major source of dysfunction in the organization.

Several reasons explain these disagreements between expectation and performance. The expectation can be unrealistic (``testers should find all the bugs in our code''), the performance may fall short of reasonable demands due to poor skill or carelessness, the recipient of the expectation may be unaware of it (a failure to coordinate or communicate), or the expectation could be in conflict with other expectations placed on the same role or on the same person.

Poor performance in the satisfaction of expectations is commonplace, but so is a performance that goes beyond the original expectation. If these disagreements recur over time, we undergo a shift in our expectations: we learn to expect the performance that is common to our experience, our ``new normal.'' And therefore, since roles are conformed by sets of expectations, and these expectations change over time, role definitions fluctuate in any social group based on past performance.

\subsection{Summary}

The following points summarize the main elements of the previous discussion.

\begin{itemize}
\item Roles are \emph{sets of expectations of behaviour} placed on a person given their position, status, history, or preference.

\item Therefore, roles are \emph{social constructs}. There is no official definition of what a certain role is.

\item A set of expectations (a role) does not need to be satisfiable. When it is not, the individual enacting the role experiences \emph{role strain}.

\item \emph{Positions} (or \emph{titles}) are official statements of a person's goals and responsibilities. Positions are not roles, though they often carry the same labels.

\item We often use \emph{labels} to refer to roles, but some roles have ill-fitting or no labels in common use.

\item Expectations are partly set by the larger \emph{culture}. However, in our domain, the culture has not enforced a singular view of the roles in software organizations.

\item Disagreements between expectation and performance are a source of organizational \emph{dysfunction}.

\item Performance that diverges from expectations, either positively or negatively, affects future expectations. Therefore, past performance in an organization causes \emph{fluctuations} in its role definitions.
\end{itemize}

Although most of these points are well established in the sociological literature, to our knowledge they had not been previously examined under the particular characteristics of software organizations. We provide empirical evidence to support them in the following pages.

% TODO: Add something about specialization with growth in this section?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related Work}

Sociologists and organizational scientists have explored the topic of professional and social roles for many years. Biddle \cite{Biddle1986} provides a good point of entry into the theoretical work in this area, as it teases out the confusion and the overloading of the concept of roles. His overview covers five main variations of role theory: \emph{functional} role theory (roles are functions in a stable social system), \emph{symbolic interactionist} role theory (roles evolve through social interaction, norms only provide broad imperatives), \emph{structural} role theory (roles are patterns of behaviour shared among people in similar positions in the social structure), \emph{organizational} role theory (which assumes that roles are mainly set by social positions in preplanned, rational settings), and \emph{cognitive} role theory (which focuses on determining which conditions give rise to expectations, how to measure expectations, and what impact do they have on social conduct). Our theoretical framework above draws mainly from the symbolic interactionist theory as described by Biddle, but although there are significant philosophical and methodological differences between role theories, they are compatible on many essential points. As Biddle points out, ``most versions of role theory presume that expectations are the major generators of roles, that expectations are learned through experience, and that persons are aware of the expectations they hold.''

A similar conceptualization comes from Turner \cite{Turner1956}, who states that roles are sets of norms, where a norm is an expected or appropriate behaviour that should be consistent. A role ``is made up of all those norms which are thought to apply to a person occupying a given position.''

Although there are roles that everyone in an organization is expected to enact (such as being a good team member), the roles that are useful in most analyses are those that allow us to differentiate between people. As Ashforth and Mael \cite{Ashforth1989} point out, we define ourselves partly in opposition to what others do. These distinctions generate both conflicts and opportunities for the members of a social group. Goode \cite{Goode1960} talks about the stresses related with role strain, argues that in general an individual's role obligations are overdemanding, and discusses coping mechanisms commonly used. In contrast, Sieber \cite{Sieber1974} claims that the additional demands caused by accumulating roles also bring rewards, and that these rewards may well compensate for the corresponding strain in many cases.

A good analysis of role fluctuation comes from Nicholson \cite{Nicholson1984}. He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).

% TODO: Check if Indi's papers, Fernando's paper, or those of Andy Begel's, are appropriate in this discussion.

Researchers discuss roles frequently, in both prescriptive and descriptive efforts, although prescriptions appear to outnumber descriptions. Chief Programmer Teams \cite{Baker1972} are an early example in our literature; they are modelled after surgical teams, with a Chief Programmer assisted by a group of professionals. Process definitions tend to include role prescriptions: Scrum \cite{Schwaber2001}, for instance, calls for a Scrum Master, a Product Owner, and a number of Developers. Practices also prescribe roles, but in a more minute detail and, usually, for a shorter term. Thus in a Fagan inspection \cite{Fagan1976} we have a Moderator, a Reader, a Coder, and several Reviewers.

Recent work in this community regularly discusses roles in interesting ways, but without a solid or consistent theoretical framework. Hoda \emph{et al.}\ \cite{Hoda2010} identified six ``informal roles'' that team members adopt in self-organizing teams. These roles (Mentor, Co-ordinator, Translator, Champion, Promoter, and Terminator) are separate from the people and positions that enact them (Agile Coaches, Developers, Business Analysts), a distinction we advocate for in this paper. But several of the ``roles'' they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of ``expectations'' from our perspective; expectations that are bundled with others to conform a richer, multi-faceted role. Moe \emph{et al.} \cite{Moe2010} use Dickinson and McIntyre's \cite{Dickinson1997} teamwork model to provide such a set of expectations for members of self-managing teams. While useful, this model is overly generic for our purposes, as it does not allow us to study how or why differences in expectations and specializations within teams arise.

Other researchers have attempted to study roles structurally, as they arise by social network interactions. For instance, Toral \emph{et al.}\ \cite{Toral2010} studied ``middle-men'' or ``brokers,'' and Ricca and Marchetto \cite{Ricca2010} examine the role of ``heroes,'' in open source projects. Their structural analysis allows them to scale to large projects and electronic repositories at the expense of detail. To our knowledge, more sophisticated analyses of roles in social networks, such as the blockmodelling work of Breiger and Pattison \cite{Breiger1986}, has not been carried out in our domain.

We can also find rich descriptions of some roles in software development. Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature. F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role fluctuation in a case in which developers attempted to rotate the job of customer support.

Finally, Litecky \emph{et al.}\ \cite{Litecky2010} report on the titles offered and the skills required in the job market in computing. Their analysis is useful for a study of roles in software organizations, as these titles and skills give us an indication of the roles that managers and human resources professionals perceive as important for their organizations.

% TODO: Add a mention to Sabrina's work

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Method and Study Execution}

\subsection{Study Design}

We performed a qualitative empirical study based on the theoretical framework presented above to improve our understanding of roles in software organizations. Since there is already a significant amount of theoretical work on the concept of roles in organizations, we did not need a grounded theory approach. We therefore chose to follow Yin's case study methodology \cite{Yin2003}. We executed an explanatory two-case case study with two embedded units of analysis (software organizations and roles).

We formulated eight research questions to guide our data collection and analysis. For each of them, we also formulated a corresponding proposition based on the theoretical framework we delineate above.

\textbf{\emph{RQ1:}} Do roles vary considerably between and within software organizations?

\emph{Proposition:} Roles vary considerably both between software organizations and within them. The sets of expectations placed on their members are different in significant ways.

\textbf{\emph{RQ2:}} What common roles are there in software organizations? Of what expectations are they built from? Can we provide a simple classification of roles?

\emph{Proposition:} We do not expect to be able to provide a simple classification of roles. However, we hypothesize that some expectations often come together, and that these patterns will allow us to form a partial picture of roles in software development.

\textbf{\emph{RQ3:}} Why are expectations clustered in roles in the way they are? That is, what is the logic, if any, behind organizing expectations in the particular ways in which they have been, and not in others?

\emph{Proposition:} Expectations are clustered in roles based on a combination of institutional, historical, structural, and personal factors. Institutional factors refer to the larger culture: what we generally expect a person fulfilling a certain role should be doing. Historical factors consist of past experiences in the organization: what people fulfilling this role or similar roles have typically done in the past. Structural factors refer to the power and affordances available to a person by virtue of their position in the social structure. Personal factors include the preferences and background of the agents enacting the role.

\textbf{\emph{RQ4:}} Is there a significant divergence between roles and positions in software organizations? If so, why, and what are its consequences?

\emph{Proposition:} There is a significant divergence between roles and positions, and the reason is that position is a structural factor, as specified above, and there are many other factors that contribute to the shaping of a role.

\textbf{\emph{RQ5:}} Why are some expectations consistently unsatisfied, and what consequences does such dissatisfaction bring?

\emph{Proposition:} Expectations may be consistently unsatisfied due to (a) poor performance, (b) unreasonable demands, (c) role strain, or (d) coordination problems.

\textbf{\emph{RQ6:}} Are there purely technical roles? Can we disassociate organizational and social concerns in technical role descriptions or prescriptions?

\emph{Proposition:} There are no purely technical roles. There are organizational and social concerns intertwined with the technical expectations placed on all members of the organization.

\textbf{\emph{RQ7:}} How do roles change over time in a software organization?

\emph{Proposition:} There are two main drivers of role fluctuation: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.

\textbf{\emph{RQ8:}} What media do people enacting similar roles use to communicate to each other? What media do they use to communicate with others?

\emph{Proposition:} People enacting similar roles use similar communication media. They communicate with people enacting different but interfacing roles through other media, which act as a boundary object between them \cite{Bowker1999}.

\subsection{Study Execution and Analysis}

To answer our research questions, we collected field data from two software organizations in the area of Victoria, Canada. Our case selection was opportunistic: we wanted groups that develop software as their main activity, and whose headquarters were located in our vicinity. We contacted managers in both organizations and negotiated access to their sites in return for feedback on their development practices.

We carried out our case study during the Winter and Spring of 2011. The first author spent about six weeks performing observations, three weeks at each of the two sites, sitting at a desk in their development areas, and attending to stand-up meetings, product meetings, all-hands meetings, and several other gatherings and social events. He also gained user access to the issue tracking and project estimation systems at both sites, and manually examined several issues, bugs, and milestones.

The first author also interviewed people spanning most areas of both organizations. He performed 36 interviews, most of them lasting almost one hour. In both sites the number of interviewed corresponded to about 50\% of the total number of organization members at the time.

Following Yin's \cite{Yin2003} suggestions, we analyzed all the observation and interview data examining whether it supported or challenged our study propositions. Whenever possible we triangulated our evidence. We were also particularly interested in evidence that did not fit with our theoretical framework, in order to refine the latter for future studies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Details from the Field Sites}

\subsection{Gallium}

Gallium was the first organization we studied.\footnote{For confidentiality reasons, the names of all organizations and people in this discussion are fictional, and some details not central to our findings have been obfuscated.} It is a biotechnology software organization headquartered in Victoria; it offers specialized software products commonly used by academic groups, pharmaceutical companies, and corporations in the biotechnology sector. At the time of our observations it employed about 45 people. Most of them worked at the Victoria office, while several members of the management team had their offices in the American West Coast, and the company also had a few agents, mostly in charge of sales, in Europe.

The development group was formed by 17 people, including the Director of Software Development, two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster. The development group was divided in four teams, although during our observations one of the teams was being phased out and its members were transitioning into another team. The largest team had eight people (a Team Lead, five Developers, and two QA), and was in charge of maintaining and iterating on the main product of the organization. The second largest had five people, counting the transitioning employees (a Team Lead, the two Tech Leads, one Developer, and one QA), and was in charge of recreating and streamlining the workflow and user interface of the main product, without being dragged by the burden of the product's legacy code. The third team (consisting of a Team Lead and a Developer) focused on generating custom scripts and solutions for particular customers, using their own product's API. The fourth team (its members already accounted for above) was wrapping up on its customer obligations for a secondary product. Organizationally, the Buildmaster was not associated with any one team, although he interacted mostly with the first and largest team. Everyone in the development group, except for its Director, worked in the same shared area: an open space with cubicles arranged in such a way as to demarcate smaller areas functioning as team rooms.

Several groups interacted with or supported the development group. There was a Product Definition group formed by three people (two Product Managers and a Product Design Lead), a Program Manager, a Project Management group (one Project Management Lead and two Project Managers), a Customer Support group (one Lead and two to three Customer Support Analysts), several Field Application and Field Implementation Specialists (numbers varied during our observations, but were never more than four), a Systems Administrator, several Sales Executives, a Graphics Designer, an Office Administrator, and eight other high-ranking staff, including the CEO and four VPs. Among these people most of those who worked in the Victoria headquarters did so in personal offices along the edges of the development group's cubicle area, or in a second cubicle area connected to the first. An exception was one of the Product Managers, who sat with a development team to be able to provide feedback more rapidly.

Our observations began at an unusual point in the organization's history, as it had recently gone through a round of layoffs. While this was an unfortunate situation for the organization, it allowed us to explore in greater detail the mechanics of shifting expectations across the organizational structure, and the employees' negotiation of the ``new normal'' at their offices.

The development group followed a lifecycle process loosely based on Scrum \cite{Schwaber2001}. The largest development team has a six-week release cycle; the second largest structures its work on shorter (two-week) cycles, but for internal purposes only, as it had not yet made any releases of its software by the time our observations ended. All teams had a daily stand-up meeting, and the Team and Tech Leads, the Buildmaster, and the Director had a second daily stand-up meeting. The Team Leads functioned as the Scrum Masters, but the Product Owner was not defined consistently. The group also deviated from the simple backlogs and story cards practices (requirements were often specified in greater detail, and they were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support), but it paid attention to Agile metrics such as burndown and velocity.

\subsection{Cobalt}

Cobalt was the second organization we studied. In contrast to Gallium, the group is publicly funded. It is a computing and data processing and storage unit servicing Canadian and international scientists in one particular field. It is part of a research institute, but except for the sharing of support and hardware staff, it has a limited interaction with the rest of the institute. At the time of our observations the group consisted of about 22 people, but the figure cannot be exact as membership is fluid. Our figure includes visiting scientists that collaborate in some sense with the group's software development activities. All of Cobalt's employees worked at their Victoria offices.

Cobalt is divided in three subgroups: Software, Operations, and Science. The Software group has 9 members, including the head of the group. Again in contrast with Gallium, the developers in Cobalt are not divided in teams. Nominally at least, all may work and request assistance for any of the projects of the group (as we will see, there was an explicit push to achieve this organizational structure in response to concerns with the more compartmentalized structure of the past). However, every project has a Team Lead, and the group has five Subsystem Leads---a subsystem being an area of interest and expertise for the group, such as Web Services or User Interface---that coordinate the work of those subsystems for each project. There are no QA specialists in this group; developers are expected to test their work and that of their peers.

The Operations group supports Software in coordinating, configuring, and ensuring the appropriate execution of its projects and processes. It includes the head of the group, a database administrator, and two software specialists. The Science group is formed by researchers: government scientists, visiting scientists, and postdoctoral fellows. There were seven at the time of our observations. As the name indicates, they are tasked with doing scientific research, not with writing production code. However, they often depend on the work of the Software group, and they have a significant say in the direction that the Software group takes. Although officially the Science group, like the other two, has a head, his functions are more those of a coordinator than a leader. The members of the Science group have considerably more autonomy in determining their activities. There are two staff employees in the group that do not form part of any team in particular: the Group Leader, and a Project Administrator.

The layout of the group's offices presented a mix of private offices and shared spaces with cubicles. Seniority seemed to be the main criterion in assigning private offices: more junior members of all three groups, and our observer, sat in the shared offices.

Our observations at Cobalt happened during a period of few changes or upsets. There had been two major restructures of the group (first, when the Software, Operations, and Science divisions were created; later, when the Software group broke its compartmentalization), but they had both taken place years before our visit.

For a few months since and including the period of our observations, some group members had been experimenting with some Agile practices. They introduced daily stand-up meetings in their projects, as well as continuous integration testing, but have had only partial success in getting test-driven development or code reviews adopted in the group. And although they work in one-week iterations, there is no push for a release at the end of each cycle; the iterations are merely used as an opportunity to set the priorities for the next week.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Findings}

We present our findings organized by the research question they address.

\subsection{Role Variation (RQ1)}

We found overwhelming evidence of variation in roles between and within organizations. In fact, it was difficult to find people in any site with roughly the same role. We offer several examples of this phenomenon, as well as the few counterexamples we found.

\emph{Product Managers, Gallium:} There are mainly three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers. These are the two Product Managers and the Product Design Lead. Some Gallium employees see the work of the three as interchangeable: one of them will be in charge of the high-level decisions of the product they're working on, and to some extent, they believe that these three people are in control of the product's direction. Others (typically those who have a closer interaction with this trio) assign different expectations to each, and a closer look reveals that there are indeed significant differences between their activities. The Product Design Lead is concerned with documenting the requirements of new functionality, he fulfills a ``historian'' responsibility (sharing design rationale with developers and explaining the pitfalls in some approaches), and he is active in shaping the organizational structure, having helped create some groups and practices throughout his time in the organization. Of the two Product Managers, one has an inside-looking approach, coordinating the development of a product and determining what requirements will make it into the product's roadmap, and when. The other Product Manager has a more outside-looking approach. He has a Ph.D.\ in the organization's area of expertise, and his presence at client meetings, trade shows, and conferences legitimizes the organization. He also understands how the users would interact with the product, and provides feedback on how to improve the product's workflow and user interface to bring it closer to the users' mental model. Additionally, we could argue that other people in the organization also performed some Product Management activities unofficially: A Customer Services Representative or a Field Specialist could persuade developers to focus on the features that cause them the most pain, for example. In sum, although all three employees have similar goals, the detailed expectations placed on them vary considerably, and are not interchangeable. Furthermore, people without a Product Management position also have some unofficial Product Management expectations.

\emph{Team Leads, Gallium and Cobalt:} Every team in the sites we studied had a Team Lead, but this leadership role was tinged by the characteristics of the team and project in question. For instance, in Cobalt, team membership is fluid---every developer may be expected to work on any project, according to need---and some Team Leads performed coordination tasks, others had a dormant leadership position, waiting for new issues or concerns in their projects to arise, and yet others were more like ``first responders'' than coordinators, requesting for assistance if the work load for their projects became too much for a single person. In Gallium, there were three Team Leads at the time of our observations. They shared a few characteristics: they were expected to develop software, and they reported to their Director. But their roles were quite different. The Team Lead of the largest team would spend about half of his time developing software; among his lead activities he meets with customer service representatives, project and product managers, and he performs estimation tasks and analyzes the performance of his team. A second Team Lead is in charge of a product about to be discontinued, and that responsibility takes a small part of his time. He is also associated with another team, but since that team has an abundance of ``leads'' (it includes the two Tech Leads of the organization), his coordination and reporting expectations are minimal. The third Team Lead has one other team member, a newcomer to the organization, and his leadership expectations become mostly mentoring and training rather than coordination and performance analysis.

\emph{Developers, Gallium and Cobalt:} The quintessential role in the software development literature was one of the hardest to pin down. A developer, of course, is expected to write production code---but what else? There is a multitude of answers to that question: some developers are expected to test their own code, some are expected to perform code reviews, some are expected to provide architectural input, some are expected to carry their own requirements analysis, some are expected to socialize, some are expected to interface with customers. None in our sample is expected to perform all of the above.

\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization.\footnote{Note that, as opposed to the previous examples, this role has no clear label to refer to it.} But the kind of stuff that falls through the cracks was different in both sites, and therefore the expectations placed on this ``glue'' role were different. In Cobalt, much of this referred to sheltering the organization from bureaucratic processes and constraints. In Gallium, it partly meant helping create and shape new and needed groups, and partly to perform activities that nobody else was tasked with doing, such as some kinds of documentation.

\emph{Project Managers, Gallium:} These project managers are an interesting counterexample to our first finding, to some extent. Although there are significant distinctions between them (one has a project management leader role, another creates templates and institutes new processes, and a third and more junior project manager implements them), the similarities in their roles seem to outweigh the differences. The main reason seems to be that these Project Managers bring a fairly specific and standardized way of doing things with them, in the form of a Project Management Body of Knowledge and certification. Part of their strategy appears to be to define their role with sufficient precision, and to enforce their definition in the rest of the organization. No other role that we examined had the emphasis on role definition and consistency that the Gallium Project Managers had.

In summary, as proposed by our theoretical framework, \textbf{we found a considerable inconsistency in role definitions both within and between organizations}. Roles (sets of expectations) are socially constructed: a developer here is not the same as a developer there, since what we expect of developers here and there varies. Part of the social construction occurs at the societal level, so that we generally expect testers to test, developers to develop, and managers to manage. In one case, such as that of Project Managers at Gallium, this broader influence is strong enough to shape most aspects of the role. In most roles in the software organizations we studied, however, this is not the case, and most of the relevant details of what it actually means to test, develop, or manage is defined by the group.

% Common roles? common expectations?

% There is a confusion between roles and positions: We have positions that attempt to approximate roles, but they're not the same thing. A role is a set of responsibilities that people place on a person by virtue of his or her background, group membership, experience, position in the organization, cultural expectations, and stated commitments.

% The fact that expectations are placed does not mean that they are satisfied. (The expectee may not know or understand the expectation; the expectation may be beyond her means; the expectation may cause a role stress, etc.)

% Therefore a dynamic of expectations that works in theory may not work in practice. Similarly, a dynamic that *doesn't* work in theory may work in practice, due to the initiative of some people who contribute beyond what the rest expects of them. However, in both cases, over time the organization grows to expect the typical performance of others (of the under/over performers), and a new understanding (a new expectation,  a new norm) takes hold.

% Roles, even technical roles, include a number of organizational and social expectations that its enactors should satisfy. No role is purely technical, and often the organizational / social / political is a prime concern.

% On growth

% As an organization grows, labour is divided further, roles become more specialized. Some expectations that took a fraction of someone's time now are more demanding, hence the tendency to formalize them as an important part of a new role. But note that organization members can fight the specialization trend.

% Some expectations arise or are solidified due to the weight of the organization or to deal with some sort of dysfunction or structural situation.

% On media

% People enacting different (but connected) roles communicate through boundary objects. Within roles, however, the communication media could be different. Since (a) communication media vary across roles, and (b) most or many roles are essential to the functioning of the organization and the satisfaction of its goals, studying/mining a single medium will miss important people and interactions.

% On the expectations that occur over and over (move to earlier?)

% In a software organization, some expectations arise over and over. We can consider them a necessity for these organizations. A list will be arbitrary perhaps (we will almost surely miss some important ones?) but listing some key expectations is still useful.
% - Develop production code
% - Explore / scout development
% - Develop configuration code
% - Develop low level (unit) tests
% - Develop high level (integration) tests
% - Assign tasks
% - Review progress
% - Set direction, vision goals
% - Resolve interpersonal conflicts
% - Represent users
% - Take care of what falls between the cracks
% - and so on...
% I think it will be worth listing all the important ones. Perhaps more than a list? The key ideas here are that (a) these expectations *need* to be satisfied, or the performance of the organization will suffer, (b) some of them are conceptually closer to each other, and having them be carried out by the same person or persons should be more effective from a coordination perspective (see Proportionality, Maturity), and (c) understanding these patterns of responsibility pairing should yield efficiency gains more generalized.

\section{Discussion}

% First a brief summary of the findings

% Second, state that the main theoretical foundations held, and were improved upon

% Implications for research: on media, important one. On better role definitions.

% Implications for practice: an awareness that role specialization is a trend that's hard to fight, and a role that makes sense in the short term to deal with a dysfunction may become a department later on that is too hard to control. Also an awareness that what people nominally do does not translate into the benefit that they bring to organizations, as their formal position does not match the expectations that people have of them.

% Anything else?

\section{Conclusion}

% Brief summary of findings and implications.

% Future work.

\section*{Acknowledgments}
Thanks for all the fish.

\bibliographystyle{IEEEtran}
\bibliography{biblio}

\end{document}

### 2011-09-21 soulgem.tex
Clarification on descriptions of sites

@@
 
-The development group was formed by 17 people, including the Director of Software Development, two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster.
+The development group was formed by 17 people, including the Director of Software Development,\footnote{All of the labels uppercased in this discussion refer to the positions that these individuals hold, not to the roles they enact.} two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster.
 The development group was divided in four teams, although during our observations one of the teams was being phased out and its members were transitioning into another team.

### 2011-09-21 expectationPeopleTable.tex
first table version

\begingroup
\newcommand{\add}{\hspace{0pt}}

\definecolor{white}{rgb}{1,1,1}
\definecolor{mygray}{rgb}{0.7,0.7,0.7}
\definecolor{yourgray}{rgb}{0.4,0.4,0.4}
\definecolor{black}{rgb}{0.0,0.0,0.0}

\newdimen\qdx
\newdimen\qda
\newdimen\qdb
\newdimen\qd
\def\rrrr#1#2#3#4#5#6{\qd=#4 % length of bar for 1.0
\qdx=\qd\multiply\qdx by 5\divide\qdx by 4
\qda=\qd
\qdb=\qd
\multiply\qda by #1\divide\qda by #3\multiply\qdb by #2\divide\qdb by #3\advance\qdb by -\qda
    \leavevmode\hbox to \qdx{\hfil\vbox{%
    \hbox{\vrule\vbox{\hrule\hbox to 1\qd
            {\vrule depth0pt height#6 width \qda#5\vrule depth0pt height#6 width \qdb\hfill}\hrule}\vrule}
    }\hfil}}
\def\rrr#1#2#3#4{\rrrr{#1}{#2}{#3}{0.19cm}{#4}{1.5ex}}

\def\w{\rrr{0}{1}{1}{\color{white}}}
\def\l{\rrr{0}{1}{1}{\color{mygray}}}
\def\g{\rrr{0}{1}{1}{\color{yourgray}}}
\def\b{\rrr{0}{1}{1}{\color{black}}}

\begin{table*}[tb]
\centering
%\footnotesize
%\scriptsize
\begin{tabular}{l@{\hspace{-3cm}}r@{\hspace{2pt}}c}
\toprule
& Expectations &People\\
\midrule
%&&& $p_1$ & $p_2$& $p_3$& $p_4$& $p_5$& $p_6$& $p_7$& $p_8$& $p_9$& $p_{10}$& $p_{11}$& $p_{12}$& $p_{13}$& $p_{14}$& $p_{15}$& $p_{16}$& $p_{17}$& $p_{18}$& $p_{19}$& $p_{20}$& $p_{21}$& $p_{22}$& $p_{23}$& $p_{24}$& $p_{25}$& $p_{26}$& $p_{27}$& $p_{28}$& $p_{29}$& $p_{30}$& $p_{31}$& $p_{32}$& $p_{33}$& $p_{34}$& $p_{35}$& $p_{36}$\\
%&&&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p&p\\
%\midrule
\textbf{Technical - Core} 
&Write bug-fix code&  \w\w\l\w\g\b\b\b\g\w\w\w\w\w\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l \\
&Maintain legacy code&  \w\w\l\w\g\b\b\b\g\w\w\w\w\w\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l \\
&Write new-feature code&\\
&Write customization (or script) code&\\
%& \\
%
%
\midrule
\textbf{Technical - Periphery}
& Write build code&\\
& Design user interface&\\
& Write user documentation&\\
& Bundle and prepare a release&\\
& Write technical documentation&\\
&Write low-level (or unit) test code&\\
& Write experimental (or exploratory) code&\\
& Write high-level (or integration) test code&\\
& Write configuration and implementation code&\\
& Write and maintain tools to aid in development&\\
%& \\
%
%
\midrule
\textbf{Technical - High level}
& Elicit requirements&\\
& Document requirements&\\
& Design system architecture&\\
& Determine if product meets quality goals&\\
& Set roadmap for system (prioritize requirements)&\\
& Provide domain context to non-domain experts (user proxy)&\\
%& \\
%
%
\midrule
\textbf{Technical - Expertise}
& Tackle the toughest technical issues and bugs&\\
& Provide historical perspective on design decisions&\\
& Provide guidance (mentorship) to newcomers on technical issues&\\
%& \\
%
%
\midrule
\textbf{Technical - Management}
& Monitor low-level task (ticket) progress&\\
& Monitor high-level task (project) progress&\\
& Determine lifecycle process and practices&\\
& Document (notify) on one's own progress&\\
& Assign low-level tasks (tickets) to individuals&\\
& Assign high-level tasks (projects) to individuals&\\
& Coordinate a product release or implementation&\\
& Provide estimates on remaining or potential tasks&\\
& Connect an issue with the person with appropriate expertise&\\
& Handle obstacles and interruptions for technical teams (PoC)&\\
%& \\
%
%
\midrule
\textbf{Organizational - External Interface}
& Obtain new users&\\
& Service current users&\\
& Receive and respond to user feedback&\\
& Obtain and maintain sources of funding&\\
& Provide domain legitimacy to organization&\\
& Provide technical legitimacy to organization&\\
& Advocate for the resolution of user/customer problems&\\
& Represent the organization's interests to funders/customers&\\
%& \\
%
%
\midrule
\textbf{Organizational - Management}
& Deal with clerical issues&\\
& Deal with financial issues&\\
& Deal with personnel issues&\\
& Deal with otherwise unattended issues&\\
& Deal with system administration issues&\\
& Set up and modify organizational structure&\\
& Communicate organizational status and news&\\
%& \\
%
%
\midrule
\textbf{Organizational - Strategic}
& Determine long-term vision for products&\\
& Determine emphasis on new market areas&\\
%& \\
\bottomrule
\end{tabular}
\end{table*}
\endgroup

### 2011-09-22 soulgem.tex
added some packages
A few more paragraphs in the Findings section

@@
 \documentclass[10pt, conference, compsocconf]{IEEEtran}
+
+\usepackage{comment}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{array}
+\usepackage{booktabs}
+\usepackage{capt-of}
+\usepackage{colortbl}
+\usepackage{graphicx}
+\usepackage{multirow}
+\usepackage{subfig}
+%\DeclareCaptionType{copyrightbox}
+\usepackage{pifont}
+\usepackage[latin1]{inputenc}
+\usepackage{times}
+\usepackage{url}
+\usepackage{boxedminipage}
+\usepackage{xspace}
+\usepackage{sepnum}
+\usepackage{cite}
+\usepackage{fancyhdr}
 
@@
 
+\input{expectationPeopleTable}
+
 \section{Method and Study Execution}
@@
 
-In summary, as proposed by our theoretical framework, \textbf{we found a considerable inconsistency in role definitions both within and between organizations}.
+In summary, as proposed by our theoretical framework, \textbf{we found inconsistent role definitions both within and between organizations}.
 Roles (sets of expectations) are socially constructed: a developer here is not the same as a developer there, since what we expect of developers here and there varies.
@@
 
-% Common roles? common expectations?
+\subsection{Role Commonalities and Expectations (RQ2)}
+
+As expected, we could not provide a simple classification of roles. However, we believe that, should we continue studying more organizations, some patterns in role definitions would begin to emerge.
+
+We can see the beginnings of such patterns in Table X, which shows the expectations placed on the people we interviewed (but not on all the members of the organizations we studied).
+The columns in the table are arranged by positions, so that people with roughly similar positions\footnote{An exact matching of people to positions was impractical.} in the two organizations are grouped together.
+The rows in the table represent common or significant expectations.
+The reader should note that this list of expectations is necessarily incomplete for the general case and somewhat subjective, as expectations in our interviews and observations were not always clearly articulated and our participants did not always agree on their expectation assignments.
+
+In the case of experienced developers, for instance, we found a shifting emphasis towards providing guidance to newcomers, helping team members understand the rationale behind design decisions in legacy code, experimenting with novel approaches and techniques, having a greater say in architectural decisions, and providing the organization with technical authority in its meetings with current or potential customers.
+In other words, they transition towards expectations that are centred on their experience and expertise.
+
+% TODO: Revisit this. Do we have something to say about other role flavours here?
+
+\subsection{Expectation Clustering (RQ3)}
+
+We proposed that expectations are clustered in roles based on a combination of institutional, historical, structural, and personal factors.
+Our empirical evidence provided support for this proposition, as detailed below.
+
+\emph{Institutional factors:} The larger culture plays an important part in shaping the expectations of members of software organizations.
+For example, when Gallium hired one of its Product Managers, he had significant domain expertise, but no software development experience.
+He initially struggled to satisfy the expectations placed on him.
+(...)
+
+\emph{Historical factors:}
+
+\emph{Structural factors:}
+
+\emph{Personal factors:}
+
+\subsection{Divergence between Roles and Positions (RQ4)}
 

### 2011-09-22 soulgem.tex
Findings - description of reasons for clustering of expectations.

@@
 Our empirical evidence provided support for this proposition, as detailed below.
+For space reasons we cannot summarize more than one example of each kind.
 
-\emph{Institutional factors:} The larger culture plays an important part in shaping the expectations of members of software organizations.
-For example, when Gallium hired one of its Product Managers, he had significant domain expertise, but no software development experience.
-He initially struggled to satisfy the expectations placed on him.
-(...)
+\emph{Institutional factors:} Although the software industry's culture still leaves plenty of room for role variability, it contributes to shape role definitions.
+We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital, it quickly expanded and began distributing its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them up with newcomers with the goal of taking care of that aspect of the business.
+Before this influx, the division of labour was more flexible and organic: most organization members did a bit of everything.
+After the influx, role definitions shifted towards a more stereotypical differentiation of roles and groups.
 
-\emph{Historical factors:}
+\emph{Historical factors:} Previous events in the history of the organization shape its current clustering of expectations, in the same way that alterations early in the life of other organic systems shape their later structure.
+Some developers that have been part of the organization for a long time are not just developers: they are also mentors, historians, and political players, despite what the organizational chart says about them.
+This affects the expectations that people place on them.
+During the Gallium downsizing, some people were demoted to keep the company afloat.
+One of them was the previous head of the QA department (which was dispersed) and, later, Program Manager.
+At the time of our observations he was a QA Analyst, yet he was still involved (and expected to be involved) in many strategic activities that do not match the institutional view of what a QA Analyst is supposed to do.
 
-\emph{Structural factors:}
+\emph{Structural factors:} A person's position in the organizational structure gives her a certain power and awareness that others may not have.
+Her structural position leads others to expect from her behaviours relating to its characteristics: she may be expected to play a bridging, directing, or even confrontational role.
+Many of the expectations placed on the head of the Operations group at Cobalt spawn from his structural position.
+He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from the System Administration group of the Institute that Cobalt is a part of, and he is expected to control the gateway that determines whether a product is ready to be released or not, that is, to assess the operational quality of the product.
+These expectations largely spawn from his position in the organization; it would be hard for someone without his position to satisfy them.
 
-\emph{Personal factors:}
+\emph{Personal factors:} Finally, a person's background, preferences, and personality shapes the expectations she receives from others.
+Perhaps the person's background could be deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from him.
+Alternatively, a person's traits (curiosity, proactivity, a talent) may lead him to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors.
+For example, the Project Administrator at Cobalt was hired to perform mainly clerical tasks, but his own preference led him to gradually take more and greater responsibilities.
+The organization learned to be dependent on such behaviour.
+When the Project Administrator temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement, and its projects suffered quickly and significantly, as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him.
+We should also note that at the time of our observations the Project Administrator position was undergoing an official redefinition in order to match real expectations better.
+In this way, personal factors were leading to historical and structural changes in the organization, so that future enactors of the role played by the Project Administrator would be affected by them.
 

### 2011-09-22 soulgem.tex
Nearly done with first draft of Findings section --just one and a half subsections left.

@@
 
-I'm leaving the abstract until later.
+Let me not to the marriage of true minds / Admit impediments, love is not love / Which alters when it alteration finds, / Or bends with the remover to remove.
 
+O no, it is an ever fixed mark / That looks on tempests and is never shaken; / It is the star to every wand'ring bark, / Whose worth's unknown although his height be taken.
+
+Love's not time's fool, though rosy lips and cheeks / Within his bending sickle's compass come, / Love alters not with his brief hours and weeks, / But bears it out even to the edge of doom:
+
+If this be error and upon me proved, / I never writ, nor no man ever loved.
 \end{abstract}
@@
 
-\subsection{Dysfunction and role fluctuation}
+\subsection{Dysfunction and role evolution}
 
@@
 \item Performance that diverges from expectations, either positively or negatively, affects future expectations.
-Therefore, past performance in an organization causes \emph{fluctuations} in its role definitions.
+Therefore, past performance in an organization causes \emph{evolutions} in its role definitions.
 \end{itemize}
@@
 
-A good analysis of role fluctuation comes from Nicholson \cite{Nicholson1984}.
+A good analysis of role evolution comes from Nicholson \cite{Nicholson1984}.
 He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).
@@
 Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature.
-F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role fluctuation in a case in which developers attempted to rotate the job of customer support.
+F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role evolution in a case in which developers attempted to rotate the job of customer support.
 
@@
 
-\textbf{\emph{RQ5:}} Why are some expectations consistently unsatisfied, and what consequences does such dissatisfaction bring?
+\textbf{\emph{RQ5:}} How do roles change over time in a software organization?
 
-\emph{Proposition:} Expectations may be consistently unsatisfied due to (a) poor performance, (b) unreasonable demands, (c) role strain, or (d) coordination problems.
+\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.
 
-\textbf{\emph{RQ6:}} Are there purely technical roles? Can we disassociate organizational and social concerns in technical role descriptions or prescriptions?
-
-\emph{Proposition:} There are no purely technical roles. There are organizational and social concerns intertwined with the technical expectations placed on all members of the organization.
-
-\textbf{\emph{RQ7:}} How do roles change over time in a software organization?
-
-\emph{Proposition:} There are two main drivers of role fluctuation: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.
-
-\textbf{\emph{RQ8:}} What media do people enacting similar roles use to communicate to each other? What media do they use to communicate with others?
+\textbf{\emph{RQ6:}} What media do people enacting similar roles use to communicate to each other? What media do they use to communicate with others?
 
@@
 
-% There is a confusion between roles and positions: We have positions that attempt to approximate roles, but they're not the same thing.
-A role is a set of responsibilities that people place on a person by virtue of his or her background, group membership, experience, position in the organization, cultural expectations, and stated commitments.
+As Table X shows, we found a considerable divergence between people's positions and the sets of expectations placed on them---their roles.
+However, there is an overlap: one can reasonably expect a Developer to write code of some kind, a QA Analyst to be concerned with software quality in one way or another, and a Team Lead to monitor and report on the team's progress.
+Table X provides an abstracted view of this finding, which also holds throughout our more detailed qualitative data: while there was no full agreement on expectations between people holding similar positions, some patterns and tendencies become apparent.
+We believe that this partial agreement comes from the fact that a person's position is, first, a statement about her placement in the organizational structure, and second, a partial statement of the institutional expectations placed on her.
+That is, those in the organization with the requisite authority give her a label that approximates their expectations (though perhaps not those of the rest of the organization) and place her in a position in the structure that should help her satisfy them.
 
-% The fact that expectations are placed does not mean that they are satisfied.
-(The expectee may not know or understand the expectation; the expectation may be beyond her means; the expectation may cause a role stress, etc.)
+However, a person's position has limited power in defining the person's role.
+As we have seen above, the expectations placed on a person also come from historical and personal factors.
+An example out of several in our observations comes from one of the Gallium Product Managers.
+He was hired as a Product Manager to steer and coordinate the development of the company's products, but he had no experience in software development, thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization he struggled to understand the level of control and direction he was expected to have.
+Over time he learned to fill the void of control that he was expected to fill, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of his peers with the same or similar positions.
 
-% Therefore a dynamic of expectations that works in theory may not work in practice.
-Similarly, a dynamic that *doesn't* work in theory may work in practice, due to the initiative of some people who contribute beyond what the rest expects of them.
-However, in both cases, over time the organization grows to expect the typical performance of others (of the under/over performers), and a new understanding (a new expectation,  a new norm) takes hold.
+\subsection{Role Evolution (RQ5)}
 
-% Roles, even technical roles, include a number of organizational and social expectations that its enactors should satisfy.
-No role is purely technical, and often the organizational / social / political is a prime concern.
+We proposed that roles evolve for two main reasons: organizational growth and dysfunction.
+Our observations provide support for both reasons, but add a qualification to the second: it is not just dysfunction that drives role evolution, but any mismatches, positive or negative, between expectations and behaviours.
 
-% On growth
+Regarding organizational growth, there is a long sociological tradition exploring its role in specialization and division of labour, and our data are far from surprising.
+In fact, it should be clear from pure argument that organizational growth drives role evolution.
+The smallest organization is formed by a single person, and that person should perform all the activities required for the functioning of the organization, from sales to debugging.
+As the organization gets more members, labour begins to be divided, not by amount, but by kind, so that upon reaching even a handful of members it is rare for a software group to perform undifferentiated labour.
+Furthermore, growth carries a coordination weight: larger organizations need to assign some of its members the role of coordinating the work of others, or of interfacing with other of its groups.
+That is, larger organizations have needs that are caused by their size, and that smaller organizations do not have.
+With organizations in the range we observed, some degree of specialization is well underway.
+We expect that, should we perform this study in a much larger organization, we would discover further differentiation between the roles of its members.
 
-% As an organization grows, labour is divided further, roles become more specialized.
-Some expectations that took a fraction of someone's time now are more demanding, hence the tendency to formalize them as an important part of a new role.
-But note that organization members can fight the specialization trend.
+Growth-driven role specialization can be resisted and overturned, as some Agile proposals suggest.
+In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation in the organization.
+The silos that group members wanted to break consisted of division of labour by project.
+They discovered that assigning one or a few persons per project introduced vulnerabilities (when something went wrong in a project, only one or a few persons could intervene) and rework (as many projects consisted of similar modules that had to be developed independently).
+To break these silos, Cobalt assigned five of its senior developers the position of Subsystem Leads, so that each of them would be the first point of contact for matters concerning their subsystem.
+The end result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a perceived decrease on the level of isolation, as projects begun to cross-pollinate and Subsystem Leads took care to help their teammates understand the particularities of their subsystems.
 
-% Some expectations arise or are solidified due to the weight of the organization or to deal with some sort of dysfunction or structural situation.
+The other important factor driving role evolution in our findings was a mismatch between expectations and behaviour.
+The fact that expectations are placed does not mean that they are satisfied.
+There are many potential reasons for their lack of satisfaction: for instance, the person upon whom they are placed may not know or understand the expectation, or it could be beyond her means, or it could turn out to be in conflict with other expectations.
+Therefore, an assignment of expectations that work in theory may not work in practice.
+Similarly, an assignment that \emph{does not} work in theory may work in practice, due to the initiative or motivation of some people who contribute beyond what the rest expects of them.
+However, in both cases (under- and over-performance), over time organization members grow to expect the typical performance of others, and a new understanding (a new expectation, and hence, a new role) takes hold.
+
+Several examples from our data above illustrate this phenomenon; we offer one more here.
+The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced, and in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being one of the main contacts for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction and next steps.
+In many ways, his role begun to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization.
+As a result, he now plays an essential role in the development of the group's products and in the maintenance of its relations with its users, a role that the rest of the organization depends on for its good functioning.
+
+\subsection{Roles and Communication Media (RQ6)}
 
@@
 
-% On the expectations that occur over and over (move to earlier?)
-
-% In a software organization, some expectations arise over and over.
-We can consider them a necessity for these organizations.
-A list will be arbitrary perhaps (we will almost surely miss some important ones?) but listing some key expectations is still useful.
-% - Develop production code
-% - Explore / scout development
-% - Develop configuration code
-% - Develop low level (unit) tests
-% - Develop high level (integration) tests
-% - Assign tasks
-% - Review progress
-% - Set direction, vision goals
-% - Resolve interpersonal conflicts
-% - Represent users
-% - Take care of what falls between the cracks
-% - and so on...
 % I think it will be worth listing all the important ones.

### 2011-09-22 soulgem.tex
First draft of Findings complete.

@@
 
+% I think it will be worth listing all the important ones.
+Perhaps more than a list?
+The key ideas here are that (a) these expectations *need* to be satisfied, or the performance of the organization will suffer, (b) some of them are conceptually closer to each other, and having them be carried out by the same person or persons should be more effective from a coordination perspective (see Proportionality, Maturity), and (c) understanding these patterns of responsibility pairing should yield efficiency gains more generalized.
+
 \subsection{Expectation Clustering (RQ3)}
@@
 
-% On media
+We expected to find that people enacting similar roles would use similar communication media, and that they would communicate with people enacting interfacing roles through other media.
+Although we found support for this claim, as we explain below, a biasing factor was the prevalence of face-to-face interactions in both of our field sites.
+Face-to-face was widely used in communication between people with similar and with different roles.
+Stand-up meetings, walks across the hall, and group meetings were commonplace.
 
-% People enacting different (but connected) roles communicate through boundary objects.
-Within roles, however, the communication media could be different.
-Since (a) communication media vary across roles, and (b) most or many roles are essential to the functioning of the organization and the satisfaction of its goals, studying/mining a single medium will miss important people and interactions.
+Regarding other media, we found role-based differences in medium selection and use.
+The broadest medium---the one that had the greatest use among the most kinds of roles---was the issue tracking system.
+People with expectations regarding code production, quality assurance, customer service, field implementation, product management, and performance monitoring all used the issue tracking system to some extent.
+Its use was not uniform: some group members used it to emit messages, others were mostly listeners, and others were more interested in the aggregate than in the contents of particular electronic traces.
 
-% I think it will be worth listing all the important ones.
-Perhaps more than a list?
-The key ideas here are that (a) these expectations *need* to be satisfied, or the performance of the organization will suffer, (b) some of them are conceptually closer to each other, and having them be carried out by the same person or persons should be more effective from a coordination perspective (see Proportionality, Maturity), and (c) understanding these patterns of responsibility pairing should yield efficiency gains more generalized.
+% TODO: Check if this is matches deSouza and Redmiles' observations.
+
+The use of other media was more localized.
+In Gallium, IRC use was common among some developer teams, while Customer Support used Google Documents to keep track of some of its cases.
+Email and phone were more heavily used in the management side of the organization than on its technical core.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \section{Discussion}
+
+\subsection{Summary}
 
@@
 % Second, state that the main theoretical foundations held, and were improved upon
+
+\subsection{Implications}
 
@@
 
+\subsection{Threats to Validity}
+
 \section{Conclusion}
 
-% Brief summary of findings and implications.
+% Very brief summary of findings and implications.
 

### 2011-09-22 soulgem.tex
First draft of Discussion - Summary.

@@
 
+% TODO: Need to add *bolded* sentences to point out all of the findings in each section.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-% First a brief summary of the findings
+We explored the sociological and organizational literature to generate a theoretical framework of roles and expectations in software organizations.
+Using that framework we studied several research questions regarding work in software groups.
 
-% Second, state that the main theoretical foundations held, and were improved upon
+We found that roles, defined as sets of expectations, were inconsistent within and between the organizations we studied.
+Roles are socially constructed; part of the construction occurs at the cultural or institutional level, and part occurs at the local, team level, and is caused by historical, structural, and personal factors.
+Therefore, although we can make some reasonable assumptions about the work of a software professional based on his role label or official position, these assumptions will only take us so far: professionals vary considerably in their responsibilities, and treating them as archetypical professionals (``developers,'' ``testers,'' and so on) dismisses a large part of what they bring or fail to bring to their job.
+
+We formulated a list of expectations that were common or significant in the two sites we observed.
+We do not think that our list is exhaustive, but we do think that it represents expectations of concern in most software organizations, and that formulating such lists and analyzing how expectations map to organization members gives us a good understanding of role definitions and interactions for software development.
+
+We also explored role evolution and found that it is driven by organizational growth (which leads to specialization) and by mismatches between expectations and behaviour (which lead to a readjustment of expectations).
+Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites in terms of inter-role communications, with the qualification that issue-tracking systems were used differently by people enacting different roles.
 

### 2011-09-23 expectationPeopleTable.tex
table with data

@@
     }\hfil}}
-\def\rrr#1#2#3#4{\rrrr{#1}{#2}{#3}{0.19cm}{#4}{1.5ex}}
+\def\rrr#1#2#3#4{\rrrr{#1}{#2}{#3}{0.15cm}{#4}{1.5ex}}
 
@@
 \def\b{\rrr{0}{1}{1}{\color{black}}}
+\def\0{\w}
+\def\1{\l}
+\def\2{\b}
+\def\angle{90}
 
@@
 %\scriptsize
-\begin{tabular}{l@{\hspace{-3cm}}r@{\hspace{2pt}}c}
+\begin{tabular}{l@{\hspace{-3cm}}rc@{\hspace{5pt}}c@{\hspace{5pt}}c@{\hspace{5pt}}c@{\hspace{5pt}}c@{\hspace{5pt}}c@{\hspace{5pt}}c@{\hspace{5pt}}c@{\hspace{5pt}}c@{\hspace{5pt}}c}
 \toprule
-& Expectations &People\\
+\vspace{1.35cm}\\
+& Expectations 
+&\multicolumn{1}{c}{\begin{rotate}{\angle}SW Directors\end{rotate} }
+&\multicolumn{1}{c}{\begin{rotate}{\angle}Tech Leads\end{rotate} }
+&\multicolumn{1}{c}{\begin{rotate}{\angle}Team Leads\end{rotate} }
+&\multicolumn{1}{c}{\begin{rotate}{\angle}Developers\end{rotate} }
+&\multicolumn{1}{c}{\begin{rotate}{\angle}Qa \& Ops\end{rotate} }
+&\multicolumn{1}{c}{\begin{rotate}{\angle}Prod Mgmt\end{rotate} }
+&\multicolumn{1}{c}{\begin{rotate}{\angle}Project Mgmt\end{rotate}} 
+&\multicolumn{1}{c}{\begin{rotate}{\angle}Support \& Field\end{rotate}} \\
+%& Expectations & \multicolumn{2}{l}{People} \\
 \midrule
@@
 \textbf{Technical - Core} 
-&Write bug-fix code&  \w\w\l\w\g\b\b\b\g\w\w\w\w\w\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l \\
-&Maintain legacy code&  \w\w\l\w\g\b\b\b\g\w\w\w\w\w\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l \\
-&Write new-feature code&\\
-&Write customization (or script) code&\\
+&Write bug-fix code& \0\0&\1\1\1\1&\1\2\0\2\2&\1\1\1\0\1\0\2\1&\0\0\0\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0 \\
+&Maintain legacy code& \0\0&\0\1\2\1&\2\0\0\2\2&\2\0\2\0\2\2\0\0&\0\0\0\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0  \\
+&Write new-feature code& \0\0&\2\2\1\1&\1\2\0\2\1&\1\2\0\0\0\1\2\2&\0\0\0\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0 \\
+&Write customization (or script) code& \0\0&\0\0\0\0&\0\0\2\0\0&\0\0\0\2\0\0\0\0&\0\0\0\0&\0\0\0\1&\0\0\0&\0\2\0\0\0\0\\
 %& \\
@@
 \textbf{Technical - Periphery}
-& Write build code&\\
-& Design user interface&\\
-& Write user documentation&\\
-& Bundle and prepare a release&\\
-& Write technical documentation&\\
-&Write low-level (or unit) test code&\\
-& Write experimental (or exploratory) code&\\
-& Write high-level (or integration) test code&\\
-& Write configuration and implementation code&\\
-& Write and maintain tools to aid in development&\\
+& Design user interface&\0\0&\2\1\0\0&\0\0\0\0\0&\0\0\0\0\0\0\0\2&\0\0\0\0&\1\2\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Write documentation&\0\0&\0\1\0\1&\0\0\1\0\1&\1\0\1\1\0\0\0\0&\1\0\1\0&\2\0\2\0&\0\0\0&\0\0\0\0\0\0\\
+&Write low-level (or unit) test code&\0\0&\1\1\0\1&\1\1\1\1\1&\1\1\0\1\1\1\0\0&\0\0\0\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Write Build Code and prepare a release&\0\0&\0\0\0\0&\0\0\0\1\0&\0\0\0\0\0\0\0\0&\0\0\2\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Write experimental (or exploratory) code&\0\0&\1\2\0\2&\0\0\0\0\0&\0\0\0\0\2\0\1\0&\0\0\0\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Write high-level (or integration) test code&\0\0&\0\0\1\0&\0\0\0\2\0&\0\0\0\0\0\0\0\0&\2\1\0\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Write configuration and implementation code&\0\0&\0\0\0\0&\0\0\1\0\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\0\0\0&\0\0\0&\0\2\0\0\0\0\\
+& Write and maintain tools to aid in development&\0\0&\0\0\0\1&\0\0\0\1\0&\0\0\0\0\0\0\0\0&\0\0\2\1&\0\0\0\0&\0\0\0&\0\0\0\0\0\0\\
 %& \\
@@
 \textbf{Technical - High level}
-& Elicit requirements&\\
-& Document requirements&\\
-& Design system architecture&\\
-& Determine if product meets quality goals&\\
-& Set roadmap for system (prioritize requirements)&\\
-& Provide domain context to non-domain experts (user proxy)&\\
+& Design system architecture&\0\0&\2\2\2\2&\0\0\0\1\1&\0\0\0\0\0\0\0\1&\0\0\0\0&\1\0\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Elicit and document requirements&\0\0&\0\0\0\1&\0\0\0\1\1&\0\0\0\0\0\0\0\0&\0\0\0\0&\2\2\2\0&\0\0\0&\0\0\0\2\1\2\\
+& Determine if product meets quality goals&\1\0&\1\1\0\0&\1\0\0\1\0&\0\1\0\0\0\0\1\1&\2\2\1\2&\0\2\0\1&\0\0\0&\0\0\0\0\1\1\\
+& Set roadmap for system (prioritize requirements)&\1\0&\1\0\0\1&\1\0\0\2\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\2\2\2&\0\0\0&\0\0\0\0\2\1\\
+& Provide domain context to non-domain experts (user proxy)&\0\0&\0\0\0\0&\0\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\2\0\2&\0\0\0&\0\2\0\2\2\0\\
 %& \\
@@
 \textbf{Technical - Expertise}
-& Tackle the toughest technical issues and bugs&\\
-& Provide historical perspective on design decisions&\\
-& Provide guidance (mentorship) to newcomers on technical issues&\\
+& Tackle the toughest technical issues and bugs&\0\0&\0\2\1\2&\2\0\2\0\0&\1\0\0\0\2\0\0\0&\0\0\0\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Provide historical perspective on design decisions&\0\0&\2\1\0\2&\0\0\0\0\0&\2\0\2\0\0\0\0\0&\0\1\0\0&\2\0\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Provide guidance (mentorship) to newcomers on technical issues&\0\0&\1\0\0\0&\0\0\2\2\2&\2\0\0\0\0\0\0\1&\0\0\0\0&\0\0\0\0&\0\0\0&\0\0\0\0\0\0\\
 %& \\
@@
 \textbf{Technical - Management}
-& Monitor low-level task (ticket) progress&\\
-& Monitor high-level task (project) progress&\\
-& Determine lifecycle process and practices&\\
-& Document (notify) on one's own progress&\\
-& Assign low-level tasks (tickets) to individuals&\\
-& Assign high-level tasks (projects) to individuals&\\
-& Coordinate a product release or implementation&\\
-& Provide estimates on remaining or potential tasks&\\
-& Connect an issue with the person with appropriate expertise&\\
-& Handle obstacles and interruptions for technical teams (PoC)&\\
+& Determine lifecycle process and practices&\2\1&\1\0\0\0&\1\0\1\2\1&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\0\0\0&\1\2\0&\0\0\0\0\0\0\\
+& Assign low-level tasks (tickets) to individuals&\0\1&\1\0\0\0&\2\0\1\1\0&\0\0\0\0\0\0\0\0&\0\0\0\2&\0\0\1\0&\1\1\0&\1\2\1\0\0\0\\
+& Assign high-level tasks (projects) to individuals&\2\2&\0\0\0\1&\0\0\0\0\0&\0\0\0\0\0\0\00&\0\0\0\0&\0\0\1\0&\0\0\0&\0\0\0\0\0\0\\
+& Coordinate a product release or implementation&\1\1&\1\0\1\1&\2\0\0\2\0&\0\0\0\0\0\0\0\0&\0\1\2\2&\1\2\2\0&\2\2\1&\0\1\0\0\0\0\\
+& Provide estimates on remaining or potential tasks&\1\1&\2\1\1\0&\2\0\1\2\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\1\1\0&\2\2\1&\0\0\0\0\0\0\\
+& Connect an issue with the person with appropriate expertise&\1\1&\0\0\0\1&\2\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\0\0\0&\1\1\0&\1\2\2\0\1\0\\
+& Handle obstacles and interruptions for technical teams (PoC)&\1\2&\0\0\0\0&\2\1\1\0\1&\0\0\0\0\0\0\0\0&\0\0\0\2&\0\0\0\0&\2\2\2&\0\0\0\0\0\0\\
 %& \\
@@
 \textbf{Organizational - External Interface}
-& Obtain new users&\\
-& Service current users&\\
-& Receive and respond to user feedback&\\
-& Obtain and maintain sources of funding&\\
-& Provide domain legitimacy to organization&\\
-& Provide technical legitimacy to organization&\\
-& Advocate for the resolution of user/customer problems&\\
-& Represent the organization's interests to funders/customers&\\
+& Receive and respond to user feedback&\0\2&\0\0\0\0&\0\1\0\1\1&\0\0\0\0\0\0\0\1&\0\0\0\1&\0\0\0\2&\2\2\0&\2\2\2\0\1\1\\
+& Obtain new users or sources of funding&\0\2&\0\0\0\0&\0\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\0\0\1&\0\0\0&\0\0\0\2\0\0\\
+& Provide domain legitimacy to organization&\0\0&\0\0\0\0&\0\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\2\0\2&\0\0\0&\0\0\0\0\0\0\\
+& Provide technical legitimacy to organization&\2\1&\2\0\0\2&\0\0\0\0\1&\0\0\0\0\0\0\0\0&\0\0\0\1&\0\0\0\0&\0\0\0&\0\0\0\0\0\0\\
+& Advocate for the resolution of user/customer problems&\1\2&\0\0\0\0&\1\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\1&\0\1\0\0&\2\2\0&\2\2\2\1\2\2\\
+& Represent the organization's interests to funders/customers&\0\2&\0\0\0\1&\0\1\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\1\0\2&\0\0\0&\0\0\0\1\1\1\\
 %& \\
@@
 \textbf{Organizational - Management}
-& Deal with clerical issues&\\
-& Deal with financial issues&\\
-& Deal with personnel issues&\\
-& Deal with otherwise unattended issues&\\
-& Deal with system administration issues&\\
-& Set up and modify organizational structure&\\
-& Communicate organizational status and news&\\
+& Deal with personnel issues&\2\2&\0\1\0\0&\1\1\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\1&\0\0\0\1&\0\0\2&\0\0\1\0\0\0\\
+& Deal with otherwise unattended issues&\1\1&\0\0\0\1&\1\0\0\0\0&\0\0\0\0\0\0\0\0&\0\2\0\1&\0\1\0\1&\0\0\2&\0\0\0\0\0\0\\
+& Set up and modify organizational structure&\2\2&\0\0\0\1&\0\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\1&\0\0\1\1&\0\0\0&\0\0\1\0\0\0\\
+& Communicate organizational status and news&\2\2&\0\0\0\0&\1\0\1\0\0&\0\0\0\0\0\0\0\0&\0\0\0\1&\0\0\0\0&\1\1\1&\0\0\1\0\0\0\\
+& Deal with financial, clerical, or system administration issues&\0\1&\0\0\0\0&\0\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\1&\0\0\0\0&\0\0\2&\0\0\0\0\0\0\\
 %& \\
@@
 \textbf{Organizational - Strategic}
-& Determine long-term vision for products&\\
-& Determine emphasis on new market areas&\\
+& Determine long-term vision for products&\1\2&\1\0\0\2&\0\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\2\0\2&\0\0\0&\0\0\0\0\0\0\\
+& Determine emphasis on new market areas&\0\1&\0\0\0\1&\0\0\0\0\0&\0\0\0\0\0\0\0\0&\0\0\0\0&\0\2\0\1&\0\0\0&\0\0\0\0\0\0\\
 %& \\

### 2011-09-23 soulgem.tex
table with data

--- 2011-09-22 soulgem.tex
@@
 \usepackage{fancyhdr}
+\usepackage{rotating}
 
@@
 
-Let me not to the marriage of true minds / Admit impediments, love is not love / Which alters when it alteration finds, / Or bends with the remover to remove.
+I'm leaving the abstract until later.
 
-O no, it is an ever fixed mark / That looks on tempests and is never shaken; / It is the star to every wand'ring bark, / Whose worth's unknown although his height be taken.
-
-Love's not time's fool, though rosy lips and cheeks / Within his bending sickle's compass come, / Love alters not with his brief hours and weeks, / But bears it out even to the edge of doom:
-
-If this be error and upon me proved, / I never writ, nor no man ever loved.
 \end{abstract}
@@
 
-\subsection{Dysfunction and role evolution}
+\subsection{Dysfunction and role fluctuation}
 
@@
 \item Performance that diverges from expectations, either positively or negatively, affects future expectations.
-Therefore, past performance in an organization causes \emph{evolutions} in its role definitions.
+Therefore, past performance in an organization causes \emph{fluctuations} in its role definitions.
 \end{itemize}
@@
 
-A good analysis of role evolution comes from Nicholson \cite{Nicholson1984}.
+A good analysis of role fluctuation comes from Nicholson \cite{Nicholson1984}.
 He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).
@@
 Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature.
-F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role evolution in a case in which developers attempted to rotate the job of customer support.
+F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role fluctuation in a case in which developers attempted to rotate the job of customer support.
 
@@
 
-\textbf{\emph{RQ5:}} How do roles change over time in a software organization?
+\textbf{\emph{RQ5:}} Why are some expectations consistently unsatisfied, and what consequences does such dissatisfaction bring?
 
-\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.
+\emph{Proposition:} Expectations may be consistently unsatisfied due to (a) poor performance, (b) unreasonable demands, (c) role strain, or (d) coordination problems.
 
-\textbf{\emph{RQ6:}} What media do people enacting similar roles use to communicate to each other? What media do they use to communicate with others?
+\textbf{\emph{RQ6:}} Are there purely technical roles? Can we disassociate organizational and social concerns in technical role descriptions or prescriptions?
+
+\emph{Proposition:} There are no purely technical roles. There are organizational and social concerns intertwined with the technical expectations placed on all members of the organization.
+
+\textbf{\emph{RQ7:}} How do roles change over time in a software organization?
+
+\emph{Proposition:} There are two main drivers of role fluctuation: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.
+
+\textbf{\emph{RQ8:}} What media do people enacting similar roles use to communicate to each other? What media do they use to communicate with others?
 
@@
 Our empirical evidence provided support for this proposition, as detailed below.
-For space reasons we cannot summarize more than one example of each kind.
 
-\emph{Institutional factors:} Although the software industry's culture still leaves plenty of room for role variability, it contributes to shape role definitions.
-We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital, it quickly expanded and began distributing its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them up with newcomers with the goal of taking care of that aspect of the business.
-Before this influx, the division of labour was more flexible and organic: most organization members did a bit of everything.
-After the influx, role definitions shifted towards a more stereotypical differentiation of roles and groups.
+\emph{Institutional factors:} The larger culture plays an important part in shaping the expectations of members of software organizations.
+For example, when Gallium hired one of its Product Managers, he had significant domain expertise, but no software development experience.
+He initially struggled to satisfy the expectations placed on him.
+(...)
 
-\emph{Historical factors:} Previous events in the history of the organization shape its current clustering of expectations, in the same way that alterations early in the life of other organic systems shape their later structure.
-Some developers that have been part of the organization for a long time are not just developers: they are also mentors, historians, and political players, despite what the organizational chart says about them.
-This affects the expectations that people place on them.
-During the Gallium downsizing, some people were demoted to keep the company afloat.
-One of them was the previous head of the QA department (which was dispersed) and, later, Program Manager.
-At the time of our observations he was a QA Analyst, yet he was still involved (and expected to be involved) in many strategic activities that do not match the institutional view of what a QA Analyst is supposed to do.
+\emph{Historical factors:}
 
-\emph{Structural factors:} A person's position in the organizational structure gives her a certain power and awareness that others may not have.
-Her structural position leads others to expect from her behaviours relating to its characteristics: she may be expected to play a bridging, directing, or even confrontational role.
-Many of the expectations placed on the head of the Operations group at Cobalt spawn from his structural position.
-He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from the System Administration group of the Institute that Cobalt is a part of, and he is expected to control the gateway that determines whether a product is ready to be released or not, that is, to assess the operational quality of the product.
-These expectations largely spawn from his position in the organization; it would be hard for someone without his position to satisfy them.
+\emph{Structural factors:}
 
-\emph{Personal factors:} Finally, a person's background, preferences, and personality shapes the expectations she receives from others.
-Perhaps the person's background could be deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from him.
-Alternatively, a person's traits (curiosity, proactivity, a talent) may lead him to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors.
-For example, the Project Administrator at Cobalt was hired to perform mainly clerical tasks, but his own preference led him to gradually take more and greater responsibilities.
-The organization learned to be dependent on such behaviour.
-When the Project Administrator temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement, and its projects suffered quickly and significantly, as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him.
-We should also note that at the time of our observations the Project Administrator position was undergoing an official redefinition in order to match real expectations better.
-In this way, personal factors were leading to historical and structural changes in the organization, so that future enactors of the role played by the Project Administrator would be affected by them.
+\emph{Personal factors:}
 
@@
 
-As Table X shows, we found a considerable divergence between people's positions and the sets of expectations placed on them---their roles.
-However, there is an overlap: one can reasonably expect a Developer to write code of some kind, a QA Analyst to be concerned with software quality in one way or another, and a Team Lead to monitor and report on the team's progress.
-Table X provides an abstracted view of this finding, which also holds throughout our more detailed qualitative data: while there was no full agreement on expectations between people holding similar positions, some patterns and tendencies become apparent.
-We believe that this partial agreement comes from the fact that a person's position is, first, a statement about her placement in the organizational structure, and second, a partial statement of the institutional expectations placed on her.
-That is, those in the organization with the requisite authority give her a label that approximates their expectations (though perhaps not those of the rest of the organization) and place her in a position in the structure that should help her satisfy them.
+% There is a confusion between roles and positions: We have positions that attempt to approximate roles, but they're not the same thing.
+A role is a set of responsibilities that people place on a person by virtue of his or her background, group membership, experience, position in the organization, cultural expectations, and stated commitments.
 
-However, a person's position has limited power in defining the person's role.
-As we have seen above, the expectations placed on a person also come from historical and personal factors.
-An example out of several in our observations comes from one of the Gallium Product Managers.
-He was hired as a Product Manager to steer and coordinate the development of the company's products, but he had no experience in software development, thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization he struggled to understand the level of control and direction he was expected to have.
-Over time he learned to fill the void of control that he was expected to fill, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of his peers with the same or similar positions.
+% The fact that expectations are placed does not mean that they are satisfied.
+(The expectee may not know or understand the expectation; the expectation may be beyond her means; the expectation may cause a role stress, etc.)
 
-\subsection{Role Evolution (RQ5)}
+% Therefore a dynamic of expectations that works in theory may not work in practice.
+Similarly, a dynamic that *doesn't* work in theory may work in practice, due to the initiative of some people who contribute beyond what the rest expects of them.
+However, in both cases, over time the organization grows to expect the typical performance of others (of the under/over performers), and a new understanding (a new expectation,  a new norm) takes hold.
 
-We proposed that roles evolve for two main reasons: organizational growth and dysfunction.
-Our observations provide support for both reasons, but add a qualification to the second: it is not just dysfunction that drives role evolution, but any mismatches, positive or negative, between expectations and behaviours.
+% Roles, even technical roles, include a number of organizational and social expectations that its enactors should satisfy.
+No role is purely technical, and often the organizational / social / political is a prime concern.
 
-Regarding organizational growth, there is a long sociological tradition exploring its role in specialization and division of labour, and our data are far from surprising.
-In fact, it should be clear from pure argument that organizational growth drives role evolution.
-The smallest organization is formed by a single person, and that person should perform all the activities required for the functioning of the organization, from sales to debugging.
-As the organization gets more members, labour begins to be divided, not by amount, but by kind, so that upon reaching even a handful of members it is rare for a software group to perform undifferentiated labour.
-Furthermore, growth carries a coordination weight: larger organizations need to assign some of its members the role of coordinating the work of others, or of interfacing with other of its groups.
-That is, larger organizations have needs that are caused by their size, and that smaller organizations do not have.
-With organizations in the range we observed, some degree of specialization is well underway.
-We expect that, should we perform this study in a much larger organization, we would discover further differentiation between the roles of its members.
+% On growth
 
-Growth-driven role specialization can be resisted and overturned, as some Agile proposals suggest.
-In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation in the organization.
-The silos that group members wanted to break consisted of division of labour by project.
-They discovered that assigning one or a few persons per project introduced vulnerabilities (when something went wrong in a project, only one or a few persons could intervene) and rework (as many projects consisted of similar modules that had to be developed independently).
-To break these silos, Cobalt assigned five of its senior developers the position of Subsystem Leads, so that each of them would be the first point of contact for matters concerning their subsystem.
-The end result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a perceived decrease on the level of isolation, as projects begun to cross-pollinate and Subsystem Leads took care to help their teammates understand the particularities of their subsystems.
+% As an organization grows, labour is divided further, roles become more specialized.
+Some expectations that took a fraction of someone's time now are more demanding, hence the tendency to formalize them as an important part of a new role.
+But note that organization members can fight the specialization trend.
 
-The other important factor driving role evolution in our findings was a mismatch between expectations and behaviour.
-The fact that expectations are placed does not mean that they are satisfied.
-There are many potential reasons for their lack of satisfaction: for instance, the person upon whom they are placed may not know or understand the expectation, or it could be beyond her means, or it could turn out to be in conflict with other expectations.
-Therefore, an assignment of expectations that work in theory may not work in practice.
-Similarly, an assignment that \emph{does not} work in theory may work in practice, due to the initiative or motivation of some people who contribute beyond what the rest expects of them.
-However, in both cases (under- and over-performance), over time organization members grow to expect the typical performance of others, and a new understanding (a new expectation, and hence, a new role) takes hold.
-
-Several examples from our data above illustrate this phenomenon; we offer one more here.
-The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced, and in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being one of the main contacts for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction and next steps.
-In many ways, his role begun to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization.
-As a result, he now plays an essential role in the development of the group's products and in the maintenance of its relations with its users, a role that the rest of the organization depends on for its good functioning.
-
-\subsection{Roles and Communication Media (RQ6)}
+% Some expectations arise or are solidified due to the weight of the organization or to deal with some sort of dysfunction or structural situation.
 
@@
 
+% On the expectations that occur over and over (move to earlier?)
+
+% In a software organization, some expectations arise over and over.
+We can consider them a necessity for these organizations.
+A list will be arbitrary perhaps (we will almost surely miss some important ones?) but listing some key expectations is still useful.
+% - Develop production code
+% - Explore / scout development
+% - Develop configuration code
+% - Develop low level (unit) tests
+% - Develop high level (integration) tests
+% - Assign tasks
+% - Review progress
+% - Set direction, vision goals
+% - Resolve interpersonal conflicts
+% - Represent users
+% - Take care of what falls between the cracks
+% - and so on...
 % I think it will be worth listing all the important ones.

### 2011-09-23 expectationPeopleTable.tex
some table voodoo

--- 2011-09-21 expectationPeopleTable.tex
@@
 %\scriptsize
-\begin{tabular}{l@{\hspace{-3cm}}r@{\hspace{2pt}}c}
+\begin{tabular}{@{}l@{\hspace{-1.5cm}}r@{\hspace{5pt}}
+c@{\hspace{2pt}}c@{\hspace{7pt}}
+c@{\hspace{2pt}}c@{\hspace{7pt}}
+c@{\hspace{2pt}}c@{\hspace{7pt}}
+c@{\hspace{2pt}}c@{\hspace{7pt}}
+c@{\hspace{2pt}}c@{\hspace{7pt}}
+c@{\hspace{2pt}}c@{\hspace{7pt}}
+c@{\hspace{2pt}}c@{\hspace{7pt}}
+c@{\hspace{2pt}}c@{}}
 \toprule
-& Expectations &People\\
+\vspace{1.4cm}\\
+& Expectations 
+&\multicolumn{2}{c}{\begin{rotate}{\angle}SW Directors\end{rotate} }
+&\multicolumn{2}{c}{\begin{rotate}{\angle}Tech Leads\end{rotate} }
+&\multicolumn{2}{c}{\begin{rotate}{\angle}Team Leads\end{rotate} }
+&\multicolumn{2}{c}{\begin{rotate}{\angle}Developers\end{rotate} }
+&\multicolumn{2}{c}{\begin{rotate}{\angle}Qa \& Ops\end{rotate} }
+&\multicolumn{2}{c}{\begin{rotate}{\angle}Prod Mgmt\end{rotate} }
+&\multicolumn{2}{c}{\begin{rotate}{\angle}Project Mgmt\end{rotate}} 
+&\multicolumn{2}{c}{\begin{rotate}{\angle}Support \& Field\end{rotate}} \\
+%& Expectations & \multicolumn{2}{l}{People} \\
 \midrule
@@
 %\midrule
-\textbf{Technical - Core} 
-&Write bug-fix code&  \w\w\l\w\g\b\b\b\g\w\w\w\w\w\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l \\
-&Maintain legacy code&  \w\w\l\w\g\b\b\b\g\w\w\w\w\w\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l\l \\
-&Write new-feature code&\\
-&Write customization (or script) code&\\
+\textbf{Technical} 
+&Write bug-fix code& \0&\0&\1\1&\1\1&\1\2\0&\2\2&\1\1\1\0\1\0&\2\1&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0 \\
+\textbf{Core}
+&Maintain legacy code& \0&\0&\0\1&\2\1&\2\0\0&\2\2&\2\0\2\0\2\2&\0\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0  \\
+&Write new-feature code& \0&\0&\2\2&\1\1&\1\2\0&\2\1&\1\2\0\0\0\1&\2\2&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0 \\
+&Write customization (or script) code& \0&\0&\0\0&\0\0&\0\0\2&\0\0&\0\0\0\2\0\0&\0\0&\0\0\0&\0&\0\0\0&\1&\0\0&\0&\0\2\0\0&\0\0\\
 %& \\
@@
 \midrule
-\textbf{Technical - Periphery}
-& Write build code&\\
-& Design user interface&\\
-& Write user documentation&\\
-& Bundle and prepare a release&\\
-& Write technical documentation&\\
-&Write low-level (or unit) test code&\\
-& Write experimental (or exploratory) code&\\
-& Write high-level (or integration) test code&\\
-& Write configuration and implementation code&\\
-& Write and maintain tools to aid in development&\\
+\textbf{Technical}
+& Design user interface&\0&\0&\2\1&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\2&\0\0\0&\0&\1\2\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+\textbf{Periphery}
+& Write documentation&\0&\0&\0\1&\0\1&\0\0\1&\0\1&\1\0\1\1\0\0&\0\0&\1\0\1&\0&\2\0\2&\0&\0\0&\0&\0\0\0\0&\0\0\\
+&Write low-level (or unit) test code&\0&\0&\1\1&\0\1&\1\1\1&\1\1&\1\1\0\1\1\1&\0\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Write Build Code and prepare a release&\0&\0&\0\0&\0\0&\0\0\0&\1\0&\0\0\0\0\0\0&\0\0&\0\0\2&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Write experimental (or exploratory) code&\0&\0&\1\2&\0\2&\0\0\0&\0\0&\0\0\0\0\2\0&\1\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Write high-level (or integration) test code&\0&\0&\0\0&\1\0&\0\0\0&\2\0&\0\0\0\0\0\0&\0\0&\2\1\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Write configuration and implementation code&\0&\0&\0\0&\0\0&\0\0\1&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\2\0\0&\0\0\\
+& Write and maintain tools to aid in development&\0&\0&\0\0&\0\1&\0\0\0&\1\0&\0\0\0\0\0\0&\0\0&\0\0\2&\1&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
 %& \\
@@
 \midrule
-\textbf{Technical - High level}
-& Elicit requirements&\\
-& Document requirements&\\
-& Design system architecture&\\
-& Determine if product meets quality goals&\\
-& Set roadmap for system (prioritize requirements)&\\
-& Provide domain context to non-domain experts (user proxy)&\\
+\textbf{Technical}
+& Design system architecture&\0&\0&\2\2&\2\2&\0\0\0&\1\1&\0\0\0\0\0\0&\0\1&\0\0\0&\0&\1\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+\textbf{High level}
+& Elicit and document requirements&\0&\0&\0\0&\0\1&\0\0\0&\1\1&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\2\2\2&\0&\0\0&\0&\0\0\0\2&\1\2\\
+& Determine if product meets quality goals&\1&\0&\1\1&\0\0&\1\0\0&\1\0&\0\1\0\0\0\0&\1\1&\2\2\1&\2&\0\2\0&\1&\0\0&\0&\0\0\0\0&\1\1\\
+& Set roadmap for system (prioritize requirements)&\1&\0&\1\0&\0\1&\1\0\0&\2\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\2&\2&\0\0&\0&\0\0\0\0&\2\1\\
+& Provide domain context to non-domain experts (user proxy)&\0&\0&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\0&\2&\0\0&\0&\0\2\0\2&\2\0\\
 %& \\
@@
 \midrule
-\textbf{Technical - Expertise}
-& Tackle the toughest technical issues and bugs&\\
-& Provide historical perspective on design decisions&\\
-& Provide guidance (mentorship) to newcomers on technical issues&\\
+\textbf{Technical}
+& Tackle the toughest technical issues and bugs&\0&\0&\0\2&\1\2&\2\0\2&\0\0&\1\0\0\0\2\0&\0\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+\textbf{Expertise}
+& Provide historical perspective on design decisions&\0&\0&\2\1&\0\2&\0\0\0&\0\0&\2\0\2\0\0\0&\0\0&\0\1\0&\0&\2\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Provide guidance (mentorship) to newcomers on technical issues&\0&\0&\1\0&\0\0&\0\0\2&\2\2&\2\0\0\0\0\0&\0\1&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
 %& \\
@@
 \midrule
-\textbf{Technical - Management}
-& Monitor low-level task (ticket) progress&\\
-& Monitor high-level task (project) progress&\\
-& Determine lifecycle process and practices&\\
-& Document (notify) on one's own progress&\\
-& Assign low-level tasks (tickets) to individuals&\\
-& Assign high-level tasks (projects) to individuals&\\
-& Coordinate a product release or implementation&\\
-& Provide estimates on remaining or potential tasks&\\
-& Connect an issue with the person with appropriate expertise&\\
-& Handle obstacles and interruptions for technical teams (PoC)&\\
+\textbf{Technical}
+& Determine lifecycle process and practices&\2&\1&\1\0&\0\0&\1\0\1&\2\1&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\0&\0&\1\2&\0&\0\0\0\0&\0\0\\
+ \textbf{Management}
+& Assign low-level tasks (tickets) to individuals&\0&\1&\1\0&\0\0&\2\0\1&\1\0&\0\0\0\0\0\0&\0\0&\0\0\0&\2&\0\0\1&\0&\1\1&\0&\1\2\1\0&\0\0\\
+& Assign high-level tasks (projects) to individuals&\2&\2&\0\0&\0\1&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\1&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Coordinate a product release or implementation&\1&\1&\1\0&\1\1&\2\0\0&\2\0&\0\0\0\0\0\0&\0\0&\0\1\2&\2&\1\2\2&\0&\2\2&\1&\0\1\0\0&\0\0\\
+& Provide estimates on remaining or potential tasks&\1&\1&\2\1&\1\0&\2\0\1&\2\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\1\1&\0&\2\2&\1&\0\0\0\0&\0\0\\
+& Connect an issue with the person with appropriate expertise&\1&\1&\0\0&\0\1&\2\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\0&\0&\1\1&\0&\1\2\2\0&\1\0\\
+& Handle obstacles and interruptions for technical teams (PoC)&\1&\2&\0\0&\0\0&\2\1\1&\0\1&\0\0\0\0\0\0&\0\0&\0\0\0&\2&\0\0\0&\0&\2\2&\2&\0\0\0\0&\0\0\\
 %& \\
@@
 \midrule
-\textbf{Organizational - External Interface}
-& Obtain new users&\\
-& Service current users&\\
-& Receive and respond to user feedback&\\
-& Obtain and maintain sources of funding&\\
-& Provide domain legitimacy to organization&\\
-& Provide technical legitimacy to organization&\\
-& Advocate for the resolution of user/customer problems&\\
-& Represent the organization's interests to funders/customers&\\
+\textbf{Organizational}
+& Receive and respond to user feedback&\0&\2&\0\0&\0\0&\0\1\0&\1\1&\0\0\0\0\0\0&\0\1&\0\0\0&\1&\0\0\0&\2&\2\2&\0&\2\2\2\0&\1\1\\
+\textbf{External Interface}
+& Obtain new users or sources of funding&\0&\2&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\0&\1&\0\0&\0&\0\0\0\2&\0\0\\
+& Provide domain legitimacy to organization&\0&\0&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\0&\2&\0\0&\0&\0\0\0\0&\0\0\\
+& Provide technical legitimacy to organization&\2&\1&\2\0&\0\2&\0\0\0&\0\1&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Advocate for the resolution of user/customer problems&\1&\2&\0\0&\0\0&\1\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\1\0&\0&\2\2&\0&\2\2\2\1&\2\2\\
+& Represent the organization's interests to funders/customers&\0&\2&\0\0&\0\1&\0\1\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\1\0&\2&\0\0&\0&\0\0\0\1&\1\1\\
 %& \\
@@
 \midrule
-\textbf{Organizational - Management}
-& Deal with clerical issues&\\
-& Deal with financial issues&\\
-& Deal with personnel issues&\\
-& Deal with otherwise unattended issues&\\
-& Deal with system administration issues&\\
-& Set up and modify organizational structure&\\
-& Communicate organizational status and news&\\
+\textbf{Organizational}
+& Deal with personnel issues&\2&\2&\0\1&\0\0&\1\1\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\1&\0\0&\2&\0\0\1\0&\0\0\\
+\textbf{Management}
+& Deal with otherwise unattended issues&\1&\1&\0\0&\0\1&\1\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\2\0&\1&\0\1\0&\1&\0\0&\2&\0\0\0\0&\0\0\\
+& Set up and modify organizational structure&\2&\2&\0\0&\0\1&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\1&\1&\0\0&\0&\0\0\1\0&\0\0\\
+& Communicate organizational status and news&\2&\2&\0\0&\0\0&\1\0\1&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\1\1&\1&\0\0\1\0&\0\0\\
+& Deal with financial, clerical, or system administration issues&\0&\1&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\0\0&\2&\0\0\0\0&\0\0\\
 %& \\
@@
 \midrule
-\textbf{Organizational - Strategic}
-& Determine long-term vision for products&\\
-& Determine emphasis on new market areas&\\
+\textbf{Organizational}
+& Determine long-term vision for products&\1&\2&\1\0&\0\2&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\0&\2&\0\0&\0&\0\0\0\0&\0\0\\
+\textbf{Strategic}
+& Determine emphasis on new market areas&\0&\1&\0\0&\0\1&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\0&\1&\0\0&\0&\0\0\0\0&\0\0\\
 %& \\
+\midrule
+& Groups 
+& A & B
+& A & B
+& A & B
+& A & B 
+& A & B
+& A & B
+& A & B
+& A & B\\
 \bottomrule

### 2011-09-23 soulgem.tex

--- 2011-09-22 soulgem.tex
@@
 
-% Implications for research: on media, important one. On better role definitions.
+\emph{Research:} First, the fact that role definitions are not institutionally uniform and respond to local (historical, structural, personal) factors should help clarify common misunderstandings and assumptions made by researchers in our studies.
+An archetypical developer, team lead, or tester, is rarely found in practice.
+Instead, thinking in terms of tasks and the people who are expected to perform them will help us reattach our studies and findings to software development in the real world.
 
-% Implications for practice: an awareness that role specialization is a trend that's hard to fight, and a role that makes sense in the short term to deal with a dysfunction may become a department later on that is too hard to control.
-Also an awareness that what people nominally do does not translate into the benefit that they bring to organizations, as their formal position does not match the expectations that people have of them.
+Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation clusters (that is, the expectations that make good sense, organizationally, to be placed on the same individual), which should lead to better role definitions, and for the study of bottlenecks and deficiencies in organizational structure.
+For instance, the discussion of the proper ratio of developers to testers, which is moot in the case of teams where there is no QA and developers do all the testing, can be now explored in terms of the proper ratio of people expected to write production code to people expected to write integration tests or to determine the level of quality of the product.
 
-% Anything else?
+Third, the observation that media use varies depending on the kinds of expectations placed on the user complements and helps explain the earlier finding that electronic repositories are incomplete and often misleading \cite{Aranda2009}.
+If software repository miners focus on a single electronic repository, they are likely to miss systematically important areas of organizational coordination: they will be blind to some people and expectations.
+Depending on the research questions they address, this may be a serious concern.
+Mining issue tracking systems, judging by our two cases, should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate, and focus only on those who emit messages in the system.
+
+\emph{Practice:} Our qualitative data confirm, in the software development domain, the old sociological finding that the role specialization trend as an organization grows is hard to fight.
+This is not necessarily bad, but teams of would-be generalists (which Agile techniques often propose) need to calculate the long-term impacts of seemingly necessary short-term specializations.
+A role that makes sense in the present to deal with a dysfunction may spin off to become a compartmentalized department later on, with its own political, budgetary, and organizational goals.
+This may not be in the interest of the current members of the organization.
+
+Additionally, we found that many practitioners do not have an awareness of the full scope of their peers' activities, and that such an awareness leads them to misunderstand the benefits that they bring to the organization.
+It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that the their formal position does not match the expectations that various groups of people have of them.
+In these cases a reductionism to position labels would be harmful.
+
+% TODO: Anything else?
 
@@
 
+The two organizations we studied were small (less than fifty members at the time of our observations), geographically co-located for the most part, and North American.
+The extent to which our findings are generalizable might therefore be questioned.
+However, most of our findings are supported by a theoretical framework that is not built upon these particular characteristics, but is instead general and interdisciplinary.
+We expect that some of the particular expectation clusters will vary in other organizations, depending on their size, physical co-location, and culture, but we also expect the underlying rationale and observations to hold across them.
+
+During our observations, Gallium underwent some role transitions, as it struggled to rearrange itself after a round of layoffs.
+We kept in mind the unusual state of this organization in our data analysis, and we believe that it contributed to our understanding of role evolution and mismatches in expectation assignment.
+
+Throughout this paper, we defined role as a ``set of expectations.'' However, an ``expectation'' is a subjective concept: we expect many things from many people, and an objective articulation of these expectations may never be quite right, missing some attributes or formulating an expectation at the wrong level of detail.
+We acknowledge this threat, but we see it as a necessary evil in dealing with this subject.
+We hope that with more research in the area a consensus will emerge on the right level of analysis for expectations in software development work.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Conclusion}
 
-% Very brief summary of findings and implications.
+Very brief summary of findings and implications.
 
-% Future work.
+Future work.
 

### 2011-09-23 soulgem.tex
First full draft complete (still no abstract).

@@
 
-Very brief summary of findings and implications.
+In this paper we presented a theoretical framework to make sense of roles and expectations in software development organizations, and we reported on a case study that used this framework to produce insights into coordination and the division of labour for software professionals.
 
-Future work.
+The framework and our findings generate several further questions, which we plan to address in future work.
+First, there is the issue of the right level of granularity to articulate expectations in software development.
+Second, an analysis of commonly-found and commonly-useful patterns in expectation clustering.
+Third, an analysis of  appropriate ratios of expectation assignment in software groups.
+And fourth, an exploration of the differences in specialization and expectation clustering in organizations that differ significantly from those we observed (that is, in larger organizations, geographically distributed, and imbued in different cultures).
+We hope others in the research community join us in addressing these questions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \section*{Acknowledgments}
-Thanks for all the fish.
+
+We thank the participants of our study, who generously opened their doors to us and gave us their time. We also thank Whoever Helps Us With Comments and Proofreading.
 

### 2011-09-24 soulgem.tex
Second draft of introduction.

@@
 
-People develop software, but the activities that each of them perform to do so differ from those of their colleagues: they divide their labour.
-For instance, some may write production code, others test it, and still others try to ensure that the product satisfies a real need.
-In order to talk about this division of labour, we abstract their responsibilities into a number of \emph{roles} (such as ``developer,'' ``tester,'' and ``product manager'') and we use these abstractions as labels to refer to the set of activities that we expect them to carry out.
-These labels appear everywhere in our academic literature, as well as in books for practitioners, human resources postings, blog posts, and everyday conversation.
+Poeple divide their labour in order to produce software systems.
+Some write production code, others test it, and still others try to ensure that the product satisfies a real need.
+Researchers and practitioners usually abstract these responsibilities into a number of role labels (such as ``developer,'' ``tester,'' and ``product manager''); we use these abstractions to refer to the set of activities that we expect them to carry out.
+These labels are common everywhere in our academic literature, as well as in textbooks, books for practitioners, human resources postings, blog posts, and everyday conversation.
 
-Despite this commonplace use, we have no clear indication that the people using these labels use them to refer to the same meanings.
+Despite this commonplace use, we have no clear indication that the people using these labels refer to the same meanings.
 For instance, what does a ``developer'' actually need to do in order to fulfill her responsibilities?
-What should she avoid doing?
-When should we judge that she has failed to fulfill her responsibilities, and would the answer be generalizable throughout the many domains in which the software industry functions?
+What should we expect her to be unable to do?
+And would the answer be generalizable throughout the many areas in which the software industry operates?
 
-In more mature domains, especially in those of critical importance, the division of labour is fairly formal and clear.
+In some domains, especially the more mature or critical ones, the division of labour is fairly formal and clear.
 Surgeons, anesthesiologists, nurses, and patients generally know what to expect from each other, in and out of the operating room.
-Knowing that someone is ``a nurse'' will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board.
-In our own community we use role labels just as directly and unthinkingly as in the medical domain, but without realizing (or such is the premise of our paper) that we may be talking about very different things, based on our varying assumptions, experiences, and cultures.
+Knowing that someone is a nurse will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board.
+But although in our own community we assign labels just as directly and unthinkingly as in the medical domain, we do not realize (or such is the premise of our paper) that we may be talking about very different things, based on our varying assumptions, experiences, and cultures.
 
 This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has unexplored repercussions for researchers and practitioners alike.
-Researchers may assume coordination dynamics between, for instance, analysts, designers, and developers that have no connection with the interactions between those groups in reality.
+Researchers may prescribe practices to be carried out by one role without realizing that the role is usually shared and spread out among a group of people.
 Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him.
@@
 
-In this paper we did not attempt to provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success.
+In this paper we do not attempt to provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success.
 Instead, (a) we examine the concepts of roles, positions, and expectations in software development groups, providing a theoretical framework based on sociological and organizational research, (b) we report the findings from a case study at two software organizations in which we used this theoretical framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.

### 2011-09-24 soulgem.tex
Second draft of Related Work (which now comes before Theoretical Framework).

@@
 Instead, (a) we examine the concepts of roles, positions, and expectations in software development groups, providing a theoretical framework based on sociological and organizational research, (b) we report the findings from a case study at two software organizations in which we used this theoretical framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Related Work}
+
+Sociologists and organizational scientists have explored the topic of professional and social roles for a long time.
+Biddle \cite{Biddle1986} provides a good point of entry into the theoretical work in this area, as it teases out the confusion and the overloading of the concept of roles.
+His overview covers five main variations of theoretical thinking: \emph{functional} role theory (roles are functions in a stable social system), \emph{symbolic interactionist} role theory (roles evolve through social interaction, norms only provide broad imperatives), \emph{structural} role theory (roles are patterns of behaviour shared among people in similar positions in the social structure), \emph{organizational} role theory (roles are mainly set by social positions in preplanned, rational settings), and \emph{cognitive} role theory (which focuses on determining which conditions give rise to expectations, how to measure expectations, and what impact do they have on social conduct).
+Although there are significant philosophical and methodological differences between role theories, they are compatible on many essential points.
+As Biddle points out, ``most versions of role theory presume that expectations are the major generators of roles, that expectations are learned through experience, and that persons are aware of the expectations they hold.''
+
+A similar conceptualization comes from Turner \cite{Turner1956}, who states that roles are sets of norms, where a norm is an expected or appropriate behaviour that should be consistent.
+A role ``is made up of all those norms which are thought to apply to a person occupying a given position.''
+
+% Could take this para out if we need space.
+Although there are roles that everyone in an organization is expected to enact, the roles that are useful in most analyses are those that allow us to differentiate between people.
+As Ashforth and Mael \cite{Ashforth1989} point out, we define ourselves partly in opposition to what others do.
+These distinctions generate both conflicts and opportunities for the members of a social group.
+
+% And if we need *more* space, this is only tangentially relevant.
+Goode \cite{Goode1960} talks about the stresses related with role strain, argues that in general an individual's role obligations are over-demanding, and discusses coping mechanisms commonly used.
+In contrast, Sieber \cite{Sieber1974} claims that the additional demands caused by accumulating roles also bring rewards, and that these rewards may well compensate for the corresponding strain in many cases.
+
+A good analysis of role evolution comes from Nicholson \cite{Nicholson1984}.
+He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).
+
+% TODO: Check if Indi's papers, Fernando's paper, or those of Andy Begel's, are appropriate in this discussion.
+
+In the software engineering literature, researchers discuss roles frequently, in both prescriptive and descriptive efforts.
+Chief Programmer Teams \cite{Baker1972} are an early example of a prescriptive approach.
+Process definitions tend to include role prescriptions: Scrum \cite{Schwaber2001}, for instance, calls for a Scrum Master, a Product Owner, and a number of developers.
+Practices also prescribe roles, but in a more minute detail and, usually, for a shorter term.
+Thus in a Fagan inspection \cite{Fagan1976} we have a Moderator, a Reader, a Coder, and several Reviewers.
+
+Recent work regularly discusses roles in interesting ways, but without a solid or consistent theoretical framework.
+Hoda \emph{et al.}\ \cite{Hoda2010} identified six ``informal roles'' that team members adopt in self-organizing teams.
+The authors distinguish between the roles and the people and positions that enact them, which is a distinction we advocate for in this paper.
+But several of the roles they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of expectations from our perspective, expectations that are bundled with others to conform a richer, multi-faceted role, as we will discuss in the next section.
+
+Moe \emph{et al.} \cite{Moe2010} use Dickinson and McIntyre's \cite{Dickinson1997} teamwork model to provide such a set of expectations for members of self-managing teams.
+This is certainly useful, although it cannot be used as a basis for our work, as it does not allow us to study how or why differences in expectations and specializations within teams arise.
+
+Other researchers have attempted to study roles structurally, as they arise by social network interactions.
+For instance, Toral \emph{et al.}\ \cite{Toral2010} studied ``middle-men'' or ``brokers,'' and Ricca and Marchetto \cite{Ricca2010} examine the role of ``heroes,'' in open source projects.
+Their structural analysis allows them to scale to large projects and electronic repositories, at the expense of detail.
+To our knowledge, more sophisticated analyses of roles from a structural role theory perspective, such as the sociological blockmodelling work of Breiger and Pattison \cite{Breiger1986}, has not been carried out in our domain.
+
+We can also find rich descriptions of some roles in software development.
+Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature.
+F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a comprehensive report of the consequences of role evolution in a case in which developers attempted to rotate the responsibilities of customer support.
+
+Finally, Litecky \emph{et al.}\ \cite{Litecky2010} report on the titles offered and the skills required in the job market in computing.
+Their analysis is useful for a study of roles in software organizations, as these titles and skills give us an indication of the roles that managers and human resources professionals perceive as important for their organizations.
+
+% TODO: Add a mention to Sabrina's work
 
@@
 % TODO: Add something about specialization with growth in this section?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Related Work}
-
-Sociologists and organizational scientists have explored the topic of professional and social roles for many years.
-Biddle \cite{Biddle1986} provides a good point of entry into the theoretical work in this area, as it teases out the confusion and the overloading of the concept of roles.
-His overview covers five main variations of role theory: \emph{functional} role theory (roles are functions in a stable social system), \emph{symbolic interactionist} role theory (roles evolve through social interaction, norms only provide broad imperatives), \emph{structural} role theory (roles are patterns of behaviour shared among people in similar positions in the social structure), \emph{organizational} role theory (which assumes that roles are mainly set by social positions in preplanned, rational settings), and \emph{cognitive} role theory (which focuses on determining which conditions give rise to expectations, how to measure expectations, and what impact do they have on social conduct).
-Our theoretical framework above draws mainly from the symbolic interactionist theory as described by Biddle, but although there are significant philosophical and methodological differences between role theories, they are compatible on many essential points.
-As Biddle points out, ``most versions of role theory presume that expectations are the major generators of roles, that expectations are learned through experience, and that persons are aware of the expectations they hold.''
-
-A similar conceptualization comes from Turner \cite{Turner1956}, who states that roles are sets of norms, where a norm is an expected or appropriate behaviour that should be consistent.
-A role ``is made up of all those norms which are thought to apply to a person occupying a given position.''
-
-Although there are roles that everyone in an organization is expected to enact (such as being a good team member), the roles that are useful in most analyses are those that allow us to differentiate between people.
-As Ashforth and Mael \cite{Ashforth1989} point out, we define ourselves partly in opposition to what others do.
-These distinctions generate both conflicts and opportunities for the members of a social group.
-Goode \cite{Goode1960} talks about the stresses related with role strain, argues that in general an individual's role obligations are overdemanding, and discusses coping mechanisms commonly used.
-In contrast, Sieber \cite{Sieber1974} claims that the additional demands caused by accumulating roles also bring rewards, and that these rewards may well compensate for the corresponding strain in many cases.
-
-A good analysis of role evolution comes from Nicholson \cite{Nicholson1984}.
-He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).
-
-% TODO: Check if Indi's papers, Fernando's paper, or those of Andy Begel's, are appropriate in this discussion.
-
-Researchers discuss roles frequently, in both prescriptive and descriptive efforts, although prescriptions appear to outnumber descriptions.
-Chief Programmer Teams \cite{Baker1972} are an early example in our literature; they are modelled after surgical teams, with a Chief Programmer assisted by a group of professionals.
-Process definitions tend to include role prescriptions: Scrum \cite{Schwaber2001}, for instance, calls for a Scrum Master, a Product Owner, and a number of Developers.
-Practices also prescribe roles, but in a more minute detail and, usually, for a shorter term.
-Thus in a Fagan inspection \cite{Fagan1976} we have a Moderator, a Reader, a Coder, and several Reviewers.
-
-Recent work in this community regularly discusses roles in interesting ways, but without a solid or consistent theoretical framework.
-Hoda \emph{et al.}\ \cite{Hoda2010} identified six ``informal roles'' that team members adopt in self-organizing teams.
-These roles (Mentor, Co-ordinator, Translator, Champion, Promoter, and Terminator) are separate from the people and positions that enact them (Agile Coaches, Developers, Business Analysts), a distinction we advocate for in this paper.
-But several of the ``roles'' they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of ``expectations'' from our perspective; expectations that are bundled with others to conform a richer, multi-faceted role.
-Moe \emph{et al.} \cite{Moe2010} use Dickinson and McIntyre's \cite{Dickinson1997} teamwork model to provide such a set of expectations for members of self-managing teams.
-While useful, this model is overly generic for our purposes, as it does not allow us to study how or why differences in expectations and specializations within teams arise.
-
-Other researchers have attempted to study roles structurally, as they arise by social network interactions.
-For instance, Toral \emph{et al.}\ \cite{Toral2010} studied ``middle-men'' or ``brokers,'' and Ricca and Marchetto \cite{Ricca2010} examine the role of ``heroes,'' in open source projects.
-Their structural analysis allows them to scale to large projects and electronic repositories at the expense of detail.
-To our knowledge, more sophisticated analyses of roles in social networks, such as the blockmodelling work of Breiger and Pattison \cite{Breiger1986}, has not been carried out in our domain.
-
-We can also find rich descriptions of some roles in software development.
-Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature.
-F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a rich report of the consequences of an attempted role evolution in a case in which developers attempted to rotate the job of customer support.
-
-Finally, Litecky \emph{et al.}\ \cite{Litecky2010} report on the titles offered and the skills required in the job market in computing.
-Their analysis is useful for a study of roles in software organizations, as these titles and skills give us an indication of the roles that managers and human resources professionals perceive as important for their organizations.
-
-% TODO: Add a mention to Sabrina's work
 

### 2011-09-24 soulgem.tex
Second draft of Theoretical Framework section.

@@
 
-Sociologists and organizational scientists have studied the concept of roles in social groups over decades.
-We take their findings as the starting points for the discussion below, acknowledging them where appropriate.
+The following framework is distilled from the work of sociologists and organizational scientists.
+To our knowledge this is the first attempt at analyzing these concepts in the context of software development work.
 
@@
 A \textbf{\emph{role}} is a set of expectations of behaviour placed on a person given their position, status, history, or preference \cite{Turner1956,Gordon1976,Rizzo1970}.
-This entails that roles are \emph{social constructions}: in a real sense, a ``developer'' is whatever the relevant social group determines that a developer should be.
-There is no archetypal developer.
+This entails that roles are \emph{social constructions} \cite{Berger1967}: in a real sense, a ``developer'' is whatever the relevant social group determines that a developer should be.
+There is no prototypical developer.
 Upon learning that a colleague is a developer, all the members of the organization place a number of expectations on said colleague.
@@
 Some colleagues may have no expectations other than that the developer should strive for the well-being of the organization.
-Other colleagues, especially if they interact with the developer on a regular basis, may have many more expectations, and these may be far more concrete.
+Other colleagues, especially if they interact with the developer on a regular basis, may have many more expectations, and these may be far more concrete: to adhere to the specified API or to put some test hooks in place.
 People outside the formal organization, but with connections or interactions to it, may hold other expectations on her behaviour.
@@
 A person's title plays a part in determining what she will feel compelled to do, and what she believes she is expected to do, but it is only one element of her role.
-Roles are enacted, positions are occupied (several people could fulfill the same role, despite their different positions; a person can fulfill several roles while holding a single position).
-For instance, one could carry an architectural role without holding an Software Architect position, and one could hold a Software Architect position without performing any of the activities that others would expect from an architect.
+For instance, one could enact an architectural role without holding an Software Architect position, and one could hold a Software Architect position without performing any of the activities that others would expect from an architect.
 
-We can see the positions in an organization, and their accompanying lists of responsibilities, as the explicit manifestations of what the fraction of the members of the organization that have the power to set those positions and responsibilities expect from their colleagues.
-In that sense they are a part of the total set of expectations placed upon organizational roles; a powerful part (especially if a position's responsibilities are tied to incentives and job evaluations), but still only a part.
+Positions and their responsibilities are often the explicit articulation of what a powerful fraction of the organization (the managers) expect from their colleagues.
+In that sense they are a part of the total set of expectations placed upon organizational roles; an important part (especially if tied to incentives and job evaluations), but still only a part.
 
@@
 A fitting label helps us convey the essence of the role to our interlocutors.
-But as we will see some roles have ill-fitting labels, and others have no labels or easy ways to refer to them in concrete terms (in many organizations there are people who are expected to ``keep the place together,'' for example).
-We must then remember that neither the label nor the formal position make the role.
+But some roles have ill-fitting labels, and others have no labels or easy ways to refer to them in concrete terms.
+We must remember that neither the label nor the formal position completely define the role.
 
-\subsection{Roles and culture}
+\subsection{Institutionalized roles}
 
-Of course, the larger culture plays an important part in setting and steering our expectations.
-In some domains the culture is sufficiently strong to establish an interpretation of a role throughout the population.
-As we mention in the Introduction, in the medical domain, a combination of intensive training, codes of ethics, professional societies, and familiarization in person and through popular culture leads us to expect certain behaviours and activities from a doctor, and it leads the doctor to become attuned to the expectations placed upon her.
-We find it shocking if a doctor deviates from these expectations significantly.
-In our own domain, a similar combination of forces (education, past experiences, popular culture) allows us to make some general assumptions about software development roles that we can reasonably expect will hold in most cases.
-For example, we would be surprised if a tester did not spend at least part of his time preparing or running tests.
-And so, thanks to this larger culture, a newcomer to the test department will likely assume that he is expected to prepare and run tests, and his peers and supervisors will indeed have these expectations placed upon him, and this is what he will spend his time doing, thereby reinforcing the general idea of what testers do.
+Culture plays an important part in setting and steering our expectations.
+In some domains the culture is sufficiently strong to establish an interpretation of a role throughout the population: the role has been institutionalized \cite{DiMaggio1991}.
+In some domains, a combination of intensive training, codes of ethics, professional societies, and familiarization lead us to expect certain behaviours and activities from a role.
+In our domain, these institutional forces allow us to make some general assumptions about software development roles that we can reasonably expect will hold in most cases.
+For example, we might be surprised if a tester did not spend at least part of his time preparing or running tests.
+And so, thanks to this process of institutionalization, a newcomer to the test department will likely assume that he is expected to prepare and run tests, his peers and supervisors will indeed place these expectations upon him, and this is what he will spend his time doing, thereby reinforcing the general idea of what testers do.
 
-However, in our domain, the forces from the larger culture have not yet enforced a singular view of what the different roles in a software organization consist of in any useful detail, or even which roles exist.
-As the evidence of our case study shows below, significant disagreements exist even within the same software organizations.
+However, the forces from the larger culture have not yet enforced a singular view of what the different roles in a software organization consist of in any useful detail, or even which roles exist.
+As the evidence of our case study shows, significant disagreements exist even within the same software organizations.
 Determining the reasons for this disagreement is beyond the scope of this paper; we speculate that the relative immaturity of the software industry, the wide variety of fields in which it functions, and the low barriers of entry to the profession all contribute to the phenomenon.
-But whatever the reason, the lack of conceptual uniformity in software development roles has implications for both research and practice, which we discuss in the coming pages.
+Whatever the reason, we make no judgment regarding the convenience of this lack of uniformity in role definitions, but we believe it should be explored in greater detail.
 
-\subsection{Dysfunction and role evolution}
+\subsection{Role evolution}
 
-In theory, we can use the definition of roles as sets of expectations of behaviour to build a conceptual map of roles in an organization: we find what every member of the organization expects from everyone else, and the aggregate of these expectations determines the roles that all members enact.
-But the members of the organization do not always act according to the expectations placed on them.
-The disagreement between expectation and actual behaviour is a major source of dysfunction in the organization.
+The members of an organization may begin setting their expectations based on institutional roles, but they do not always act according to the expectations placed on them.
+Several reasons explain this disagreement.
+In the case of performance that fail to meet expectations, it could be that the expectation is unrealistic (``testers should find all the bugs in our code''), that the performer lacks skill or care, that the performer is unaware of the expectation (a failure to coordinate or communicate), or that the expectation is in conflict with other expectations (role strain).
 
-Several reasons explain these disagreements between expectation and performance.
-The expectation can be unrealistic (``testers should find all the bugs in our code''), the performance may fall short of reasonable demands due to poor skill or carelessness, the recipient of the expectation may be unaware of it (a failure to coordinate or communicate), or the expectation could be in conflict with other expectations placed on the same role or on the same person.
+It is easy to find examples of poor performance, but so is it to find performance that goes beyond the original expectation.
+In any case, if disagreements recur over time, organization members undergo a shift in their expectations: they learn to expect the ``new normal.'' They adapt their expectations, and since roles are conformed by these expectations, role definitions evolve in any social group based on past performance, despite their institutional status.
 
-Poor performance in the satisfaction of expectations is commonplace, but so is a performance that goes beyond the original expectation.
-If these disagreements recur over time, we undergo a shift in our expectations: we learn to expect the performance that is common to our experience, our ``new normal.'' And therefore, since roles are conformed by sets of expectations, and these expectations change over time, role definitions fluctuate in any social group based on past performance.
+Another reason for role evolution arises from changes in the expectations that are needed to meet the organization's goals.
+An involvement in a new market niche may bring the need to coordinate the old products with the new one, to transfer expertise, and so on.
+One generally accepted cause of shifts in expectations is organizational growth \cite{Blau1971}.
+With growth comes specialization, and specialization consists of narrower role definitions.
+Therefore, other things being equal, we expect larger organizations to have a greater degree of role specialization.
 
@@
 
-\item Therefore, roles are \emph{social constructs}. There is no official definition of what a certain role is.
-
-\item A set of expectations (a role) does not need to be satisfiable. When it is not, the individual enacting the role experiences \emph{role strain}.
+\item Therefore, roles are \emph{social constructs}. There is no prototypical definition of what a certain role is.
 
@@
 
-\item Expectations are partly set by the larger \emph{culture}. However, in our domain, the culture has not enforced a singular view of the roles in software organizations.
+\item Expectations are partly set by our \emph{institutions}, that is, by the rules, norms, and culture prevalent in our society.
 
-\item Disagreements between expectation and performance are a source of organizational \emph{dysfunction}.
+\item Performance that diverges from expectations, either positively or negatively, affects future expectations. Therefore, past performance in an organization causes a role \emph{evolution}.
 
-\item Performance that diverges from expectations, either positively or negatively, affects future expectations.
-Therefore, past performance in an organization causes \emph{evolutions} in its role definitions.
+\item Organizational growth brings new requirements into the organization, which also leads to role evolution. In particular, growth tends to cause role \emph{specialization}.
 \end{itemize}
-
-Although most of these points are well established in the sociological literature, to our knowledge they had not been previously examined under the particular characteristics of software organizations.
-We provide empirical evidence to support them in the following pages.
-
-% TODO: Add something about specialization with growth in this section?
 

### 2011-09-24 expectationPeopleTable.tex
space savers included in the table
Second draft of expectations table. Now with a respectable caption :-P.

@@
 \def\2{\b}
-\def\angle{90}
+\def\angle{60}
 
-\begin{table*}[tb]
+\begin{table*}[tb!]
 \centering
@@
 & Expectations 
-&\multicolumn{2}{c}{\begin{rotate}{\angle}SW Directors\end{rotate} }
-&\multicolumn{2}{c}{\begin{rotate}{\angle}Tech Leads\end{rotate} }
-&\multicolumn{2}{c}{\begin{rotate}{\angle}Team Leads\end{rotate} }
-&\multicolumn{2}{c}{\begin{rotate}{\angle}Developers\end{rotate} }
-&\multicolumn{2}{c}{\begin{rotate}{\angle}Qa \& Ops\end{rotate} }
-&\multicolumn{2}{c}{\begin{rotate}{\angle}Prod Mgmt\end{rotate} }
-&\multicolumn{2}{c}{\begin{rotate}{\angle}Project Mgmt\end{rotate}} 
-&\multicolumn{2}{c}{\begin{rotate}{\angle}Support \& Field\end{rotate}} \\
+&\multicolumn{2}{l}{\begin{rotate}{\angle}SW Directors\end{rotate} }
+&\multicolumn{2}{l}{\begin{rotate}{\angle}Tech Leads\end{rotate} }
+&\multicolumn{2}{l}{\begin{rotate}{\angle}Team Leads\end{rotate} }
+&\multicolumn{2}{l}{\begin{rotate}{\angle}Developers\end{rotate} }
+&\multicolumn{2}{l}{\begin{rotate}{\angle}QA \& Ops\end{rotate} }
+&\multicolumn{2}{l}{\begin{rotate}{\angle}Product Mgmt\end{rotate} }
+&\multicolumn{2}{l}{\begin{rotate}{\angle}Project Mgmt\end{rotate}} 
+&\multicolumn{2}{l}{\begin{rotate}{\angle}Support \& Field\end{rotate}} \\
 %& Expectations & \multicolumn{2}{l}{People} \\
@@
 \midrule
-& Groups 
-& A & B
-& A & B
-& A & B
-& A & B 
-& A & B
-& A & B
-& A & B
-& A & B\\
-\bottomrule
+& Organizations 
+& G & C
+& G & C
+& G & C
+& G & C 
+& G & C
+& G & C
+& G & C
+& G & C\\
+\midrule
+\multicolumn{18}{c}{\2: Highly Expected \1: Somewhat Expected \0: Not Expected}\\
 \end{tabular}
+\caption{Assignment of expectations to people in our field sites.
+This matrix only includes our interviewees in both organizations; about half of the total number of organization members.
+Interviewees are clustered approximately by labels, despite label variations.
+Within each label cluster there is a further subdivision by organization (``G'' stands for Gallium, ``C'' for Cobalt).
+The colouring of each cell is necessarily subjective, but it corresponds to a rough consensus view of the expectations placed on each individual.}
 \end{table*}

### 2011-09-24 soulgem.tex
Second draft of expectations table. Now with a respectable caption :-P.
Second draft of Method and Study Execution section.

@@
 An involvement in a new market niche may bring the need to coordinate the old products with the new one, to transfer expertise, and so on.
-One generally accepted cause of shifts in expectations is organizational growth \cite{Blau1971}.
+One generally accepted cause of shifts in expectations is organizational growth \cite{Blau1971,Haveman1993}.
 With growth comes specialization, and specialization consists of narrower role definitions.
@@
 
-\input{expectationPeopleTable}
-
 \section{Method and Study Execution}
@@
 We performed a qualitative empirical study based on the theoretical framework presented above to improve our understanding of roles in software organizations.
-Since there is already a significant amount of theoretical work on the concept of roles in organizations, we did not need a grounded theory approach.
-We therefore chose to follow Yin's case study methodology \cite{Yin2003}.
+We chose to follow Yin's case study methodology \cite{Yin2003}.
 We executed an explanatory two-case case study with two embedded units of analysis (software organizations and roles).
@@
 
-\textbf{\emph{RQ3:}} Why are expectations clustered in roles in the way they are?
-That is, what is the logic, if any, behind organizing expectations in the particular ways in which they have been, and not in others?
+\textbf{\emph{RQ3:}} What is the logic (what are the factors) behind the particular clusters of expectations found in software organizations?
 
-\emph{Proposition:} Expectations are clustered in roles based on a combination of institutional, historical, structural, and personal factors.
-Institutional factors refer to the larger culture: what we generally expect a person fulfilling a certain role should be doing.
-Historical factors consist of past experiences in the organization: what people fulfilling this role or similar roles have typically done in the past.
-Structural factors refer to the power and affordances available to a person by virtue of their position in the social structure.
-Personal factors include the preferences and background of the agents enacting the role.
+\emph{Proposition:} Expectations are clustered by institutional and organizational factors, but we do not know the specific reasons that drive this clustering.
 
-\textbf{\emph{RQ4:}} Is there a significant divergence between roles and positions in software organizations? If so, why, and what are its consequences?
+\textbf{\emph{RQ4:}} Is there a significant divergence between roles and positions in software organizations? Why?
 
-\emph{Proposition:} There is a significant divergence between roles and positions, and the reason is that position is a structural factor, as specified above, and there are many other factors that contribute to the shaping of a role.
+\emph{Proposition:} There is a significant divergence between roles and positions, and the reason is that position is an institutional factor, as specified above, and there are many other factors that contribute to the shaping of a role.
 
@@
 
-\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization \cite{Blau1971,Haveman1993}, and organizational dysfunction, which leads to rearrangements of expectations.
+\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization, and organizational dysfunction, which leads to rearrangements of expectations.
 
@@
 
-To answer our research questions, we collected field data from two software organizations in the area of Victoria, Canada.
+We collected field data from two software organizations in the area of Victoria, Canada.
 Our case selection was opportunistic: we wanted groups that develop software as their main activity, and whose headquarters were located in our vicinity.
@@
 We carried out our case study during the Winter and Spring of 2011.
-The first author spent about six weeks performing observations, three weeks at each of the two sites, sitting at a desk in their development areas, and attending to stand-up meetings, product meetings, all-hands meetings, and several other gatherings and social events.
-He also gained user access to the issue tracking and project estimation systems at both sites, and manually examined several issues, bugs, and milestones.
+The first author spent five weeks performing observations, three weeks at the first and two at the second site, sitting at a desk in their development areas, and attending to several kinds of meetings (stand-up, product, all-hands, and others).
+He also gained access to the issue tracking and project estimation systems at both sites, and manually examined several issues, bugs, and milestones.
 
@@
 Whenever possible we triangulated our evidence.
-We were also particularly interested in evidence that did not fit with our theoretical framework, in order to refine the latter for future studies.
+We were also particularly interested in evidence that did not fit with our theoretical framework, in order to refine it for future studies.
 
@@
 We present our findings organized by the research question they address.
+
+\input{expectationPeopleTable}
 

### 2011-09-24 soulgem.tex
Second draft of Details section.

@@
 At the time of our observations it employed about 45 people.
-Most of them worked at the Victoria office, while several members of the management team had their offices in the American West Coast, and the company also had a few agents, mostly in charge of sales, in Europe.
+Most of them worked at the Victoria office, while several managers had their offices in the American West Coast, and the company also had a few agents, mostly in sales, in Europe.
 
+% We might not need all this detail...
 The development group was formed by 17 people, including the Director of Software Development,\footnote{All of the labels uppercased in this discussion refer to the positions that these individuals hold, not to the roles they enact.} two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster.
 The development group was divided in four teams, although during our observations one of the teams was being phased out and its members were transitioning into another team.
-The largest team had eight people (a Team Lead, five Developers, and two QA), and was in charge of maintaining and iterating on the main product of the organization.
-The second largest had five people, counting the transitioning employees (a Team Lead, the two Tech Leads, one Developer, and one QA), and was in charge of recreating and streamlining the workflow and user interface of the main product, without being dragged by the burden of the product's legacy code.
-The third team (consisting of a Team Lead and a Developer) focused on generating custom scripts and solutions for particular customers, using their own product's API.
-The fourth team (its members already accounted for above) was wrapping up on its customer obligations for a secondary product.
+The largest team, Alpha, had eight people (a Team Lead, five Developers, and two QA), and was in charge of maintaining and iterating on the main product of the organization.
+The second largest, Beta, had five people, including the incoming employees (a Team Lead, the two Tech Leads, one Developer, and one QA), and was in charge of recreating and streamlining the workflow and user interface of the main product, without being dragged by the burden of its legacy code.
+The third team, Gamma, (consisting of a Team Lead and a Developer) focused on generating custom scripts and solutions for particular customers, using their own product's API.
+The fourth team, Delta, (its members already accounted for above) was wrapping up on its customer obligations for a secondary product.
 Organizationally, the Buildmaster was not associated with any one team, although he interacted mostly with the first and largest team.
@@
 Several groups interacted with or supported the development group.
-There was a Product Definition group formed by three people (two Product Managers and a Product Design Lead), a Program Manager, a Project Management group (one Project Management Lead and two Project Managers), a Customer Support group (one Lead and two to three Customer Support Analysts), several Field Application and Field Implementation Specialists (numbers varied during our observations, but were never more than four), a Systems Administrator, several Sales Executives, a Graphics Designer, an Office Administrator, and eight other high-ranking staff, including the CEO and four VPs.
+There was a Product Definition group formed by three people (two Product Managers and a Product Design Lead), a Program Manager, a Project Management group (one Project Management Lead and two Project Managers), a Customer Support group (one Lead and two to three Customer Support Analysts), several Field Specialists (numbers varied during our observations, but were never more than four), a Systems Administrator, several Sales Executives, a Graphics Designer, an Office Administrator, and eight other high-ranking staff, including the CEO and four VPs.
 Among these people most of those who worked in the Victoria headquarters did so in personal offices along the edges of the development group's cubicle area, or in a second cubicle area connected to the first.
-An exception was one of the Product Managers, who sat with a development team to be able to provide feedback more rapidly.
+An exception was one of the Product Managers, who sat with the Beta Team to be able to provide feedback more rapidly.
 
@@
 The development group followed a lifecycle process loosely based on Scrum \cite{Schwaber2001}.
-The largest development team has a six-week release cycle; the second largest structures its work on shorter (two-week) cycles, but for internal purposes only, as it had not yet made any releases of its software by the time our observations ended.
+The Alpha Team has a six-week release cycle, while Beta structured its work on shorter (two-week) cycles, but for internal purposes only, as it had not yet made any releases of its software by the time our observations ended.
 All teams had a daily stand-up meeting, and the Team and Tech Leads, the Buildmaster, and the Director had a second daily stand-up meeting.
 The Team Leads functioned as the Scrum Masters, but the Product Owner was not defined consistently.
-The group also deviated from the simple backlogs and story cards practices (requirements were often specified in greater detail, and they were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support), but it paid attention to Agile metrics such as burndown and velocity.
+The group also deviated from the simple backlogs and story cards practices (requirements were often specified in greater detail, and they were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support), but it paid attention to Agile metrics such as burndown charts and velocity.
 
@@
 Cobalt was the second organization we studied.
-In contrast to Gallium, the group is publicly funded.
-It is a computing and data processing and storage unit servicing Canadian and international scientists in one particular field.
+In contrast to Gallium, it is publicly funded.
+It is a computing and data processing and storage unit servicing Canadian and international scientists in one particular research field.
 It is part of a research institute, but except for the sharing of support and hardware staff, it has a limited interaction with the rest of the institute.
-At the time of our observations the group consisted of about 22 people, but the figure cannot be exact as membership is fluid.
-Our figure includes visiting scientists that collaborate in some sense with the group's software development activities.
+At the time of our observations the group consisted of about 22 people, including visiting scientists that collaborate in some sense with the group's software development activities.
 All of Cobalt's employees worked at their Victoria offices.
@@
 Again in contrast with Gallium, the developers in Cobalt are not divided in teams.
-Nominally at least, all may work and request assistance for any of the projects of the group (as we will see, there was an explicit push to achieve this organizational structure in response to concerns with the more compartmentalized structure of the past).
-However, every project has a Team Lead, and the group has five Subsystem Leads---a subsystem being an area of interest and expertise for the group, such as Web Services or User Interface---that coordinate the work of those subsystems for each project.
+Nominally at least, all may work and request assistance for any of the projects of the group (as we will see, there was an explicit push to achieve this organizational structure in response to concerns with the more ``siloed'' structure of the past).
+However, every project has a Team Lead, and the group as a whole has five Subsystem Leads---a subsystem being an area of interest and expertise for the group, such as Web Services or User Interface---that coordinate the work of those subsystems for each project.
 There are no QA specialists in this group; developers are expected to test their work and that of their peers.
@@
 The Operations group supports Software in coordinating, configuring, and ensuring the appropriate execution of its projects and processes.
-It includes the head of the group, a database administrator, and two software specialists.
-The Science group is formed by researchers: government scientists, visiting scientists, and postdoctoral fellows.
-There were seven at the time of our observations.
+It includes the head of the group, a Database Administrator, and two Software Specialists.
+The Science group is formed by researchers.
+There were seven scientists at the time of our observations.
 As the name indicates, they are tasked with doing scientific research, not with writing production code.
-However, they often depend on the work of the Software group, and they have a significant say in the direction that the Software group takes.
-Although officially the Science group, like the other two, has a head, his functions are more those of a coordinator than a leader.
+However, they often depend on the work of the Software group, they have a significant say in the direction that the Software group takes, and some of them do develop script code.
+Although officially the Science group, like the other two, has a head, his functions are more akin to those of a coordinator than a leader.
 The members of the Science group have considerably more autonomy in determining their activities.
-There are two staff employees in the group that do not form part of any team in particular: the Group Leader, and a Project Administrator.
+Finally, there are two staff employees in the group that do not form part of any team in particular: the Group Leader, and a Project Administrator.
 
@@
 Our observations at Cobalt happened during a period of few changes or upsets.
-There had been two major restructures of the group (first, when the Software, Operations, and Science divisions were created; later, when the Software group broke its compartmentalization), but they had both taken place years before our visit.
+There had been two major restructures of the group (first, when the Software, Operations, and Science divisions were created; later, when the Software group created the Team Lead and Subsystem Lead structure), but they had both taken place years before our visit.
 
-For a few months since and including the period of our observations, some group members had been experimenting with some Agile practices.
+For a few months since and including the period of our observations, some group members had been experimenting with a few Agile practices.
 They introduced daily stand-up meetings in their projects, as well as continuous integration testing, but have had only partial success in getting test-driven development or code reviews adopted in the group.

### 2011-09-26 soulgem.tex
Some revisions to Findings section.

@@
 
-\textbf{\emph{RQ1:}} Do roles vary considerably between and within software organizations?
+\textbf{\emph{RQ1:}} What common roles are there in software organizations? Of what expectations are they built from? Can we provide a simple classification of roles?
+
+\emph{Proposition:} We do not expect to be able to provide a simple classification of roles.
+However, we hypothesize that some expectations often come together, and that these patterns will form a partial picture of roles in software development.
+
+\textbf{\emph{RQ2:}} Do roles vary considerably between and within software organizations?
 
 \emph{Proposition:} Roles vary considerably both between software organizations and within them. The sets of expectations placed on their members are different in significant ways.
-
-\textbf{\emph{RQ2:}} What common roles are there in software organizations? Of what expectations are they built from? Can we provide a simple classification of roles?
-
-\emph{Proposition:} We do not expect to be able to provide a simple classification of roles.
-However, we hypothesize that some expectations often come together, and that these patterns will allow us to form a partial picture of roles in software development.
 
@@
 
-\subsection{Role Variation (RQ1)}
+\subsection{Role Commonalities and Expectations (RQ1)}
 
-We found overwhelming evidence of variation in roles between and within organizations.
-In fact, it was difficult to find people in any site with roughly the same role.
+There is no simple classification of roles in the organizations we studied.
+However, we believe that, should we continue studying more groups, some patterns in role definitions would begin to emerge.
+We can see the beginnings of such patterns in Table \ref{tableExpectations}, which shows the expectations placed on the people we interviewed (about half of the total number of organization members).
+The columns in the table represent our interviewees; the rows represent common or significant expectations.
+The reader should note that our list of expectations is necessarily incomplete for the general case.
+It is also somewhat subjective, as expectations in our interviews and observations were not always clearly articulated and our participants did not always agree on their expectation assignments.
+
+For instance, working on the technical core of the organizations' products is reserved almost exclusively for developers and their more senior differentiations, technical and team leads.
+However, the leads' expectations are more spread out: technical leads focus more on architecture and design issues, whereas team leads emphasize coordination activities.
+There are also purely organizational roles; their enactors do not need any domain or technical expertise.
+But in both organizations there were key people with more than one kind of expertise (technical, domain, and organizational), and these people seemed vital for the functioning of the organization, as they transitioned between boundaries with ease.
+A representative example is a Gallium Field Specialist (the second column in the Support \& Field cluster), who straddles all three kinds of expertise.
+Several people in different areas pointed out his relevance; in the words of one informant, ``the company depends on [him] so much he should have armed guards behind him taking him to and from work.''
+
+There are several other patterns in expectation assignment in our data.
+For space reasons we only point out three more.
+First, all the system architecture expectations were placed on people still working on the technical core, writing new feature and bug fix code.
+This is in contrast to our anecdotal experiences in larger organizations.
+Second, group members that provide domain legitimacy or credibility were critical at both organizations, and they are expected to influence product development at a high level, but not to tinker with implementation details.
+And third, some expectations, such as writing unit tests and documentation, were weakly spread out through the technical side of the organizations.
+Everyone expects developers to ``play nice'' by testing and documenting their own code, but it is culturally acceptable to ignore these expectations if the urgency of the situation demands it, as it often does.
+
+% TODO: Check build code expectation in Cobalt.
+
+\subsection{Role Variation (RQ2)}
+
+As Table \ref{tableExpectations} suggests, we found a considerable evidence of variation in roles between and within organizations.
+In fact, it was difficult to find people in either site with the same role in significant detail.
 We offer several examples of this phenomenon, as well as the few counterexamples we found.
@@
 In most roles in the software organizations we studied, however, this is not the case, and most of the relevant details of what it actually means to test, develop, or manage is defined by the group.
-
-\subsection{Role Commonalities and Expectations (RQ2)}
-
-As expected, we could not provide a simple classification of roles. However, we believe that, should we continue studying more organizations, some patterns in role definitions would begin to emerge.
-
-We can see the beginnings of such patterns in Table X, which shows the expectations placed on the people we interviewed (but not on all the members of the organizations we studied).
-The columns in the table are arranged by positions, so that people with roughly similar positions\footnote{An exact matching of people to positions was impractical.} in the two organizations are grouped together.
-The rows in the table represent common or significant expectations.
-The reader should note that this list of expectations is necessarily incomplete for the general case and somewhat subjective, as expectations in our interviews and observations were not always clearly articulated and our participants did not always agree on their expectation assignments.
-
-In the case of experienced developers, for instance, we found a shifting emphasis towards providing guidance to newcomers, helping team members understand the rationale behind design decisions in legacy code, experimenting with novel approaches and techniques, having a greater say in architectural decisions, and providing the organization with technical authority in its meetings with current or potential customers.
-In other words, they transition towards expectations that are centred on their experience and expertise.
-
-% TODO: Revisit this. Do we have something to say about other role flavours here?
-
-% I think it will be worth listing all the important ones.
-Perhaps more than a list?
-The key ideas here are that (a) these expectations *need* to be satisfied, or the performance of the organization will suffer, (b) some of them are conceptually closer to each other, and having them be carried out by the same person or persons should be more effective from a coordination perspective (see Proportionality, Maturity), and (c) understanding these patterns of responsibility pairing should yield efficiency gains more generalized.
 

### 2011-09-26 soulgem.tex
More revisions to Findings - second draft near complete.

@@
 
-\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization, and organizational dysfunction, which leads to rearrangements of expectations.
+\emph{Proposition:} There are two main drivers of role evolution: organizational growth, which leads to specialization, and divergences between expectation and performance, which lead to rearrangements of expectations.
 
@@
 In fact, it was difficult to find people in either site with the same role in significant detail.
-We offer several examples of this phenomenon, as well as the few counterexamples we found.
+We offer several examples of this phenomenon, as well as a key counterexample.
 
-\emph{Product Managers, Gallium:} There are mainly three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers.
-These are the two Product Managers and the Product Design Lead.
-Some Gallium employees see the work of the three as interchangeable: one of them will be in charge of the high-level decisions of the product they're working on, and to some extent, they believe that these three people are in control of the product's direction.
-Others (typically those who have a closer interaction with this trio) assign different expectations to each, and a closer look reveals that there are indeed significant differences between their activities.
-The Product Design Lead is concerned with documenting the requirements of new functionality, he fulfills a ``historian'' responsibility (sharing design rationale with developers and explaining the pitfalls in some approaches), and he is active in shaping the organizational structure, having helped create some groups and practices throughout his time in the organization.
+\emph{Product Management at Gallium:} There are mainly three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers: the two Product Managers and the Product Design Lead.
+Some employees see the work of the three as interchangeable, but others (typically those who have a closer interaction with this trio) assign different expectations to each, and a closer look reveals that there are indeed significant differences between their activities.
+The Product Design Lead documents the requirements of new functionality, he fulfills a ``historian'' expectation (sharing design rationale with developers and pointing out pitfalls in some approaches), and he is active in shaping the organizational structure, having helped create some groups and practices.
 Of the two Product Managers, one has an inside-looking approach, coordinating the development of a product and determining what requirements will make it into the product's roadmap, and when.
-The other Product Manager has a more outside-looking approach.
+The other one has a more outside-looking approach.
 He has a Ph.D.\ in the organization's area of expertise, and his presence at client meetings, trade shows, and conferences legitimizes the organization.
-He also understands how the users would interact with the product, and provides feedback on how to improve the product's workflow and user interface to bring it closer to the users' mental model.
-Additionally, we could argue that other people in the organization also performed some Product Management activities unofficially: A Customer Services Representative or a Field Specialist could persuade developers to focus on the features that cause them the most pain, for example.
-In sum, although all three employees have similar goals, the detailed expectations placed on them vary considerably, and are not interchangeable.
+He also understands how the users would interact with the product and helps shape the user interface to bring it closer to the users' mental model.
+Additionally, other people in the organization also performed some Product Management activities unofficially: A Customer Services Representative or a Field Specialist often persuade developers to focus on the features that cause them the most pain, for example.
+In sum, although the three employees above have similar goals, the detailed expectations placed on them vary considerably, and are not interchangeable.
 Furthermore, people without a Product Management position also have some unofficial Product Management expectations.
 
-\emph{Team Leads, Gallium and Cobalt:} Every team in the sites we studied had a Team Lead, but this leadership role was tinged by the characteristics of the team and project in question.
-For instance, in Cobalt, team membership is fluid---every developer may be expected to work on any project, according to need---and some Team Leads performed coordination tasks, others had a dormant leadership position, waiting for new issues or concerns in their projects to arise, and yet others were more like ``first responders'' than coordinators, requesting for assistance if the work load for their projects became too much for a single person.
+\emph{Team Leading at Gallium and Cobalt:} Every team in the sites we studied had a Team Lead, but this leadership role was tinged by the characteristics of the team and project in question.
+For instance, in Cobalt, team membership is fluid---every developer may be expected to work on any project, according to need---and some Team Leads performed coordination tasks, others had a dormant leadership position, waiting for new issues or concerns in their projects to arise, and yet others were more like ``first responders'' than coordinators, requesting assistance only if the work load for their projects became too much for themselves.
 In Gallium, there were three Team Leads at the time of our observations.
@@
 But their roles were quite different.
-The Team Lead of the largest team would spend about half of his time developing software; among his lead activities he meets with customer service representatives, project and product managers, and he performs estimation tasks and analyzes the performance of his team.
+The Team Lead of Alpha would spend about half of his time developing software, and in his other half he meets with customer service representatives, project and product managers, performs estimation tasks, and analyzes the performance of his team.
 A second Team Lead is in charge of a product about to be discontinued, and that responsibility takes a small part of his time.
-He is also associated with another team, but since that team has an abundance of ``leads'' (it includes the two Tech Leads of the organization), his coordination and reporting expectations are minimal.
-The third Team Lead has one other team member, a newcomer to the organization, and his leadership expectations become mostly mentoring and training rather than coordination and performance analysis.
+He is also associated with the Beta Team, but since that team has an abundance of ``leads'' (it includes the two official Tech Leads of the organization), his coordination and reporting expectations are minimal.
+The third Team Lead has one other team member, a newcomer to the organization, and his leadership expectations consist mostly of mentoring and training rather than coordination and performance analysis.
 
-\emph{Developers, Gallium and Cobalt:} The quintessential role in the software development literature was one of the hardest to pin down.
+\emph{Developers at Gallium and Cobalt:} The quintessential role in the software development literature was one of the hardest to pin down.
 A developer, of course, is expected to write production code---but what else?
-There is a multitude of answers to that question: some developers are expected to test their own code, some are expected to perform code reviews, some are expected to provide architectural input, some are expected to carry their own requirements analysis, some are expected to socialize, some are expected to interface with customers.
+There is a multitude of answers to that question: some developers are expected to perform code reviews, some are expected to provide architectural input, some are expected to carry their own requirements analysis, some are expected to socialize, some are expected to interface with customers.
 None in our sample is expected to perform all of the above.
 
-\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization.\footnote{Note that, as opposed to the previous examples, this role has no clear label to refer to it.} But the kind of stuff that falls through the cracks was different in both sites, and therefore the expectations placed on this ``glue'' role were different.
+\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization.\footnote{Note that, as opposed to the previous examples, this role has no clear label to refer to it.} But the stuff that falls through the cracks was different in both sites, and therefore the expectations were different.
 In Cobalt, much of this referred to sheltering the organization from bureaucratic processes and constraints.
@@
 
-\emph{Project Managers, Gallium:} These project managers are an interesting counterexample to our first finding, to some extent.
-Although there are significant distinctions between them (one has a project management leader role, another creates templates and institutes new processes, and a third and more junior project manager implements them), the similarities in their roles seem to outweigh the differences.
+\emph{Project Management at Gallium:} These project managers are a counterexample to our finding, to some extent.
+Although there are distinctions between them (one has a project management leader role, another creates templates and institutes new processes, and all, including a third and more junior project manager, implement them), the similarities in their roles seem to outweigh the differences.
 The main reason seems to be that these Project Managers bring a fairly specific and standardized way of doing things with them, in the form of a Project Management Body of Knowledge and certification.
+In other words, the institutionalization of their role is strong.
 Part of their strategy appears to be to define their role with sufficient precision, and to enforce their definition in the rest of the organization.
@@
 
-In summary, as proposed by our theoretical framework, \textbf{we found inconsistent role definitions both within and between organizations}.
-Roles (sets of expectations) are socially constructed: a developer here is not the same as a developer there, since what we expect of developers here and there varies.
-Part of the social construction occurs at the societal level, so that we generally expect testers to test, developers to develop, and managers to manage.
-In one case, such as that of Project Managers at Gallium, this broader influence is strong enough to shape most aspects of the role.
-In most roles in the software organizations we studied, however, this is not the case, and most of the relevant details of what it actually means to test, develop, or manage is defined by the group.
-
 \subsection{Expectation Clustering (RQ3)}
 
-We proposed that expectations are clustered in roles based on a combination of institutional, historical, structural, and personal factors.
-Our empirical evidence provided support for this proposition, as detailed below.
-For space reasons we cannot summarize more than one example of each kind.
+We found that expectations are clustered in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors.
 
 \emph{Institutional factors:} Although the software industry's culture still leaves plenty of room for role variability, it contributes to shape role definitions.
-We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital, it quickly expanded and began distributing its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them up with newcomers with the goal of taking care of that aspect of the business.
+We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital.
+It quickly expanded and distributed its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them up with newcomers with the goal of taking care of those aspect of the business.
 Before this influx, the division of labour was more flexible and organic: most organization members did a bit of everything.
@@
 During the Gallium downsizing, some people were demoted to keep the company afloat.
-One of them was the previous head of the QA department (which was dispersed) and, later, Program Manager.
+At some point, one of them had been the head of the QA department (which was dispersed) and, later, the Program Manager.
 At the time of our observations he was a QA Analyst, yet he was still involved (and expected to be involved) in many strategic activities that do not match the institutional view of what a QA Analyst is supposed to do.
@@
 \emph{Structural factors:} A person's position in the organizational structure gives her a certain power and awareness that others may not have.
-Her structural position leads others to expect from her behaviours relating to its characteristics: she may be expected to play a bridging, directing, or even confrontational role.
+Her structural position leads others to expect from her certain behaviours: she may be expected to play a bridging, directing, or even confrontational role.
 Many of the expectations placed on the head of the Operations group at Cobalt spawn from his structural position.
-He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from the System Administration group of the Institute that Cobalt is a part of, and he is expected to control the gateway that determines whether a product is ready to be released or not, that is, to assess the operational quality of the product.
-These expectations largely spawn from his position in the organization; it would be hard for someone without his position to satisfy them.
+He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from system administration concerns, and he is expected to control the gateway that determines whether a product is ready to be released or not, that is, to assess and have final say on the operational quality of the product.
+These expectations come from his position in the organization: it would be hard for someone without his position to satisfy them.
 
-\emph{Personal factors:} Finally, a person's background, preferences, and personality shapes the expectations she receives from others.
-Perhaps the person's background could be deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from him.
-Alternatively, a person's traits (curiosity, proactivity, a talent) may lead him to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors.
-For example, the Project Administrator at Cobalt was hired to perform mainly clerical tasks, but his own preference led him to gradually take more and greater responsibilities.
+\emph{Personal factors:} A person's background, preferences, and personality shapes the expectations she receives from others.
+Perhaps his background is deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from him.
+Alternatively, his traits (curiosity, proactivity, a talent) may lead him to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors.
+For example, the Project Administrator at Cobalt was hired to perform clerical tasks, but his own preference led him to gradually take more and greater responsibilities.
 The organization learned to be dependent on such behaviour.
-When the Project Administrator temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement, and its projects suffered quickly and significantly, as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him.
+When he temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement, and its projects suffered quickly and significantly, as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him.
 We should also note that at the time of our observations the Project Administrator position was undergoing an official redefinition in order to match real expectations better.
@@
 
-As Table X shows, we found a considerable divergence between people's positions and the sets of expectations placed on them---their roles.
+We found a considerable divergence between people's positions and the expectations placed on them---their roles.
 However, there is an overlap: one can reasonably expect a Developer to write code of some kind, a QA Analyst to be concerned with software quality in one way or another, and a Team Lead to monitor and report on the team's progress.
-Table X provides an abstracted view of this finding, which also holds throughout our more detailed qualitative data: while there was no full agreement on expectations between people holding similar positions, some patterns and tendencies become apparent.
-We believe that this partial agreement comes from the fact that a person's position is, first, a statement about her placement in the organizational structure, and second, a partial statement of the institutional expectations placed on her.
+
+Table \ref{tableExpectations} provides an abstracted view of this finding, which also holds throughout our more detailed qualitative data: while there was no full agreement on expectations between people holding similar positions, some patterns and tendencies become apparent.
+We believe that this partial agreement comes from the fact that a person's position is, first, a statement about her placement in the organizational \emph{structure}, and second, a partial statement of the \emph{institutional} expectations placed on her.
 That is, those in the organization with the requisite authority give her a label that approximates their expectations (though perhaps not those of the rest of the organization) and place her in a position in the structure that should help her satisfy them.
 
-However, a person's position has limited power in defining the person's role.
+However, a person's position has limited power in defining their role.
 As we have seen above, the expectations placed on a person also come from historical and personal factors.
@@
 
-We proposed that roles evolve for two main reasons: organizational growth and dysfunction.
-Our observations provide support for both reasons, but add a qualification to the second: it is not just dysfunction that drives role evolution, but any mismatches, positive or negative, between expectations and behaviours.
+We proposed that roles evolve for two reasons: organizational growth and mismatches between expectation and performance. Our data provide support for both reasons.
 
@@
 As the organization gets more members, labour begins to be divided, not by amount, but by kind, so that upon reaching even a handful of members it is rare for a software group to perform undifferentiated labour.
-Furthermore, growth carries a coordination weight: larger organizations need to assign some of its members the role of coordinating the work of others, or of interfacing with other of its groups.
+Furthermore, growth carries a coordination weight: larger organizations need to assign some of its members the role of coordinating the work of others, or of interfacing between groups.
 That is, larger organizations have needs that are caused by their size, and that smaller organizations do not have.
 With organizations in the range we observed, some degree of specialization is well underway.
-We expect that, should we perform this study in a much larger organization, we would discover further differentiation between the roles of its members.
+We expect that, should we perform this study in a larger organization, we would discover further role specialization.
 
@@
 In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation in the organization.
-The silos that group members wanted to break consisted of division of labour by project.
-They discovered that assigning one or a few persons per project introduced vulnerabilities (when something went wrong in a project, only one or a few persons could intervene) and rework (as many projects consisted of similar modules that had to be developed independently).
+The silos that group members wanted to break were project-based silos.
+They discovered that assigning one person per project introduced vulnerabilities (when something went wrong in a project, only one person could intervene) and rework (as many projects consisted of similar modules that had to be developed independently).
 To break these silos, Cobalt assigned five of its senior developers the position of Subsystem Leads, so that each of them would be the first point of contact for matters concerning their subsystem.
@@
 The fact that expectations are placed does not mean that they are satisfied.
-There are many potential reasons for their lack of satisfaction: for instance, the person upon whom they are placed may not know or understand the expectation, or it could be beyond her means, or it could turn out to be in conflict with other expectations.
-Therefore, an assignment of expectations that work in theory may not work in practice.
-Similarly, an assignment that \emph{does not} work in theory may work in practice, due to the initiative or motivation of some people who contribute beyond what the rest expects of them.
-However, in both cases (under- and over-performance), over time organization members grow to expect the typical performance of others, and a new understanding (a new expectation, and hence, a new role) takes hold.
+There are many potential reasons for their lack of satisfaction: for instance, the person upon whom they are placed may not know or understand the expectation, it could be beyond her means, or it could be in conflict with other expectations.
+Therefore, an assignment of expectations that works in theory may not work in practice.
+Similarly, an assignment that does \emph{not} work in theory may work in practice, due to the initiative or motivation of some people who contribute beyond their expectations.
+However, in both the under- and the over-performing case, with time organization members grow to expect the typical performance of others, and a new expectation (and hence, a new role) takes hold.
 

### 2011-09-26 soulgem.tex
Second full draft (including abstract) complete.

@@
 \begin{document}
-\title{The Soulgem Paper}
+\title{Working Title: Roles and Expectations in Software Organizations}
 
@@
 
-Let me not to the marriage of true minds / Admit impediments, love is not love / Which alters when it alteration finds, / Or bends with the remover to remove.
+Software professionals divide their work by roles, but the nature of these roles has not been examined nor questioned in our field.
+In this paper we formulate a theoretical framework to address this issue.
+The framework is based on sociological and organizational studies, and defines roles as sets of expectations placed on the members of the software organization.
 
-O no, it is an ever fixed mark / That looks on tempests and is never shaken; / It is the star to every wand'ring bark, / Whose worth's unknown although his height be taken.
+The paper reports on an empirical case study of two software organizations, which was informed and inspired by the theoretical framework.
+In our study we found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors.
+Furthermore, role definitions evolve through organizational growth and through mismatches between expectations and performance, and people enacting different roles use different media to communicate.
 
-Love's not time's fool, though rosy lips and cheeks / Within his bending sickle's compass come, / Love alters not with his brief hours and weeks, / But bears it out even to the edge of doom:
+Our contributions should help researchers to frame better studies of division of labour in software groups, and to avoid the pitfalls that come from unwarranted abstractions of software professional roles.
+They also provide practitioners with a conceptual tool to reason about social structure and organizational design.
 
-If this be error and upon me proved, / I never writ, nor no man ever loved.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-pink robots;
+software organizations; social structure; roles
 \end{IEEEkeywords}
@@
 \subsection{Summary}
-
-The following points summarize the main elements of the previous discussion.
 
@@
 
-We expected to find that people enacting similar roles would use similar communication media, and that they would communicate with people enacting interfacing roles through other media.
+We expected to find that people enacting similar roles use similar communication media, and that they communicate with people enacting interfacing roles through other media.
 Although we found support for this claim, as we explain below, a biasing factor was the prevalence of face-to-face interactions in both of our field sites.
-Face-to-face was widely used in communication between people with similar and with different roles.
+People with similar or different roles communicate face-to-face regularly.
 Stand-up meetings, walks across the hall, and group meetings were commonplace.
@@
 
-% TODO: Need to add *bolded* sentences to point out all of the findings in each section.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We explored the sociological and organizational literature to generate a theoretical framework of roles and expectations in software organizations.
-Using that framework we studied several research questions regarding work in software groups.
+We explored the sociological and organizational literature to generate a theoretical framework of roles and expectations.
+We used that framework to study several research questions regarding work in software organizations.
 
@@
 Roles are socially constructed; part of the construction occurs at the cultural or institutional level, and part occurs at the local, team level, and is caused by historical, structural, and personal factors.
-Therefore, although we can make some reasonable assumptions about the work of a software professional based on his role label or official position, these assumptions will only take us so far: professionals vary considerably in their responsibilities, and treating them as archetypical professionals (``developers,'' ``testers,'' and so on) dismisses a large part of what they bring or fail to bring to their job.
+Although we can make some reasonable assumptions about the work of a software professional based on his label or official position, our assumptions will only take us so far: their responsibilities vary considerably, and treating them as archetypical professionals (``developers,'' ``testers,'' and so on) dismisses a large part of what they bring or fail to bring to their job.
 
@@
 
-We also explored role evolution and found that it is driven by organizational growth (which leads to specialization) and by mismatches between expectations and behaviour (which lead to a readjustment of expectations).
+We also explored role evolution and found that it is driven by organizational growth (which leads to specialization) and by mismatches between expectations and behaviour (which lead to an adjustment of expectations).
 Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites in terms of inter-role communications, with the qualification that issue-tracking systems were used differently by people enacting different roles.
@@
 
-\emph{Research:} First, the fact that role definitions are not institutionally uniform and respond to local (historical, structural, personal) factors should help clarify common misunderstandings and assumptions made by researchers in our studies.
-An archetypical developer, team lead, or tester, is rarely found in practice.
+\emph{Research:} First, the fact that role definitions are not institutionally uniform and respond to local (historical, structural, personal) factors should help clarify common misunderstandings and assumptions made in research.
+A stereotypical developer, team lead, or tester, is rarely found in practice.
 Instead, thinking in terms of tasks and the people who are expected to perform them will help us reattach our studies and findings to software development in the real world.
@@
 Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation clusters (that is, the expectations that make good sense, organizationally, to be placed on the same individual), which should lead to better role definitions, and for the study of bottlenecks and deficiencies in organizational structure.
-For instance, the discussion of the proper ratio of developers to testers, which is moot in the case of teams where there is no QA and developers do all the testing, can be now explored in terms of the proper ratio of people expected to write production code to people expected to write integration tests or to determine the level of quality of the product.
+For instance, the discussion of the proper ratio of developers to testers, which is moot for teams where there is no QA and developers do all the testing, can be now explored in terms of the proper ratio of people expected to write production code to people expected to write integration tests or to determine the level of quality of the product.
 
 Third, the observation that media use varies depending on the kinds of expectations placed on the user complements and helps explain the earlier finding that electronic repositories are incomplete and often misleading \cite{Aranda2009}.
-If software repository miners focus on a single electronic repository, they are likely to miss systematically important areas of organizational coordination: they will be blind to some people and expectations.
+If researchers focus on a single electronic repository, they are likely to miss systematically important areas of organizational coordination: they will be blind to some people and expectations.
 Depending on the research questions they address, this may be a serious concern.
@@
 
-Additionally, we found that many practitioners do not have an awareness of the full scope of their peers' activities, and that such an awareness leads them to misunderstand the benefits that they bring to the organization.
-It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that the their formal position does not match the expectations that various groups of people have of them.
-In these cases a reductionism to position labels would be harmful.
+Additionally, we found that many practitioners do not have an awareness of the full scope of their peers' activities, and that this lack of awareness leads them to misunderstand the benefits that they bring to the organization.
+It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that their formal position does not match the expectations placed on them.
+In these cases a reductionism to position labels is harmful.
 
@@
 
-The two organizations we studied were small (less than fifty members at the time of our observations), geographically co-located for the most part, and North American.
-The extent to which our findings are generalizable might therefore be questioned.
+The two organizations we studied were small (less than fifty members), geographically co-located for the most part, and North American.
+Therefore, the generalizability of our findings might be questioned.
 However, most of our findings are supported by a theoretical framework that is not built upon these particular characteristics, but is instead general and interdisciplinary.
-We expect that some of the particular expectation clusters will vary in other organizations, depending on their size, physical co-location, and culture, but we also expect the underlying rationale and observations to hold across them.
+We expect that some of the particular expectation clusters will vary in other groups, depending on their contextual factors, but we also expect the underlying rationale and observations to hold across them.
 
@@
 Third, an analysis of  appropriate ratios of expectation assignment in software groups.
-And fourth, an exploration of the differences in specialization and expectation clustering in organizations that differ significantly from those we observed (that is, in larger organizations, geographically distributed, and imbued in different cultures).
+And fourth, an exploration of the differences in specialization and expectation clustering in organizations that differ significantly from those we observed.
 We hope others in the research community join us in addressing these questions.

### 2011-09-26 soulgem.tex
Changes based on Dana and Peggy's feedback.

@@
 
-This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has unexplored repercussions for researchers and practitioners alike.
+This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has not been studied in depth so far, and it has unexplored repercussions for researchers and practitioners alike.
+Poorly defined roles may lead to dysfunction in a project team, and thus negatively impact the quality of the software.
 Researchers may prescribe practices to be carried out by one role without realizing that the role is usually shared and spread out among a group of people.
-Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him.
+Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him \cite{Dagenais2010}.
+Their communication tools may fail to be adopted if they satisfy the needs of some roles but not of others.
 For an abstraction as heavily used as role definitions in software organization, a closer look should prove beneficial.
@@
 
-We can also find rich descriptions of some roles in software development.
+We can also find rich descriptions of some roles and role interactions in software development.
 Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature.
 F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a comprehensive report of the consequences of role evolution in a case in which developers attempted to rotate the responsibilities of customer support.
+De Souza \emph{et al.}\ \cite{deSouza2003} describe that problem reports are used as boundary objects \cite{Bowker1999} between people with different roles.
 
@@
 We carried out our case study during the Winter and Spring of 2011.
-The first author spent five weeks performing observations, three weeks at the first and two at the second site, sitting at a desk in their development areas, and attending to several kinds of meetings (stand-up, product, all-hands, and others).
-He also gained access to the issue tracking and project estimation systems at both sites, and manually examined several issues, bugs, and milestones.
+The first author spent five weeks performing observations, three weeks at the first and two at the second site, sitting at a desk in their development areas, attending to several kinds of meetings (stand-up, product, all-hands, and others), and exploring their issue tracking and project estimation systems.
 
@@
 Following Yin's \cite{Yin2003} suggestions, we analyzed all the observation and interview data examining whether it supported or challenged our study propositions.
-Whenever possible we triangulated our evidence.
-We were also particularly interested in evidence that did not fit with our theoretical framework, in order to refine it for future studies.
+We were particularly interested in evidence that did not fit with our theoretical framework, in order to refine it for future studies.
 
@@
 % We might not need all this detail...
-The development group was formed by 17 people, including the Director of Software Development,\footnote{All of the labels uppercased in this discussion refer to the positions that these individuals hold, not to the roles they enact.} two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster.
+The development group included 17 people, including the Director of Software Development,\footnote{All of the labels uppercased in this discussion refer to the positions that these individuals hold, not to the roles they enact.} two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster.
 The development group was divided in four teams, although during our observations one of the teams was being phased out and its members were transitioning into another team.
 The largest team, Alpha, had eight people (a Team Lead, five Developers, and two QA), and was in charge of maintaining and iterating on the main product of the organization.
-The second largest, Beta, had five people, including the incoming employees (a Team Lead, the two Tech Leads, one Developer, and one QA), and was in charge of recreating and streamlining the workflow and user interface of the main product, without being dragged by the burden of its legacy code.
+The second largest, Beta, had five people, including the incoming employees (a Team Lead, the two Tech Leads, one Developer, and one QA), and was in charge of recreating and streamlining the workflow and user interface of the main product, without being burdened by its legacy code.
 The third team, Gamma, (consisting of a Team Lead and a Developer) focused on generating custom scripts and solutions for particular customers, using their own product's API.
@@
 
-\subsection{Role Commonalities and Expectations (RQ1)}
+\subsection{Role Classifications and Expectations (RQ1)}
 
@@
 \emph{Product Management at Gallium:} There are mainly three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers: the two Product Managers and the Product Design Lead.
-Some employees see the work of the three as interchangeable, but others (typically those who have a closer interaction with this trio) assign different expectations to each, and a closer look reveals that there are indeed significant differences between their activities.
-The Product Design Lead documents the requirements of new functionality, he fulfills a ``historian'' expectation (sharing design rationale with developers and pointing out pitfalls in some approaches), and he is active in shaping the organizational structure, having helped create some groups and practices.
-Of the two Product Managers, one has an inside-looking approach, coordinating the development of a product and determining what requirements will make it into the product's roadmap, and when.
-The other one has a more outside-looking approach.
-He has a Ph.D.\ in the organization's area of expertise, and his presence at client meetings, trade shows, and conferences legitimizes the organization.
+The latter documents the requirements of new functionality, he fulfills a ``historian'' expectation (sharing design rationale with developers and pointing out pitfalls in some approaches), and he is active in shaping the organizational structure, having helped create some groups and practices.
+Of the two Product Managers, one has an inside-looking approach, whereas the other one has an outside-looking approach: he has a Ph.D.\ in the organization's area of expertise, and his presence at client meetings, trade shows, and conferences legitimizes the organization.
 He also understands how the users would interact with the product and helps shape the user interface to bring it closer to the users' mental model.
-Additionally, other people in the organization also performed some Product Management activities unofficially: A Customer Services Representative or a Field Specialist often persuade developers to focus on the features that cause them the most pain, for example.
-In sum, although the three employees above have similar goals, the detailed expectations placed on them vary considerably, and are not interchangeable.
-Furthermore, people without a Product Management position also have some unofficial Product Management expectations.
+Other people in the organization also performed some Product Management activities unofficially: Customer Services Representatives and Field Specialists often persuade developers to focus on the features that cause them the most pain, for example.
+In sum, although the three employees above have similar goals, the expectations placed on them vary considerably, are not interchangeable, and are not their exclusive prerogative.
 
@@
 
-We found that expectations are clustered in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors.
+We found that expectations are clustered in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors, and that these kinds of factors can interplay and provide a feedback to each other.
 
@@
 
-We found a considerable divergence between people's positions and the expectations placed on them---their roles.
-However, there is an overlap: one can reasonably expect a Developer to write code of some kind, a QA Analyst to be concerned with software quality in one way or another, and a Team Lead to monitor and report on the team's progress.
-
-Table \ref{tableExpectations} provides an abstracted view of this finding, which also holds throughout our more detailed qualitative data: while there was no full agreement on expectations between people holding similar positions, some patterns and tendencies become apparent.
+We found a divergence between people's positions and the expectations placed on them.
+There is an overlap, however, and it is worth exploring it first.
+One can reasonably expect a Developer to write code of some kind, a QA Analyst to be concerned with software quality in one way or another, and a Team Lead to monitor and report on the team's progress.
+Table \ref{tableExpectations} provides an abstracted view of these patterns, which also hold throughout our more detailed qualitative data.
 We believe that this partial agreement comes from the fact that a person's position is, first, a statement about her placement in the organizational \emph{structure}, and second, a partial statement of the \emph{institutional} expectations placed on her.
@@
 Its use was not uniform: some group members used it to emit messages, others were mostly listeners, and others were more interested in the aggregate than in the contents of particular electronic traces.
-
-% TODO: Check if this is matches deSouza and Redmiles' observations.
+This finding mirrors the observation of De Souza \emph{et al.}\ regarding issue tracking systems as boundary objects \cite{deSouza2003}.
 
 The use of other media was more localized.
-In Gallium, IRC use was common among some developer teams, while Customer Support used Google Documents to keep track of some of its cases.
-Email and phone were more heavily used in the management side of the organization than on its technical core.
+In Gallium, IRC use was common among some developer teams: each team had one channel, and these were not frequented by any non-developers.
+Customer Support staff used Google Documents to keep track of some of its cases, and to provide an ongoing documentation of recent events and requests for support.
+Email and phone were more heavily used in the management side of the organization than on its technical core: developers and QA analysts rarely made or received phone calls, and they obtained most of the information they needed about their projects through the issue tracking system.
 

### 2011-09-27 soulgem.tex
Third draft, abstract.

@@
 
-Software professionals divide their work by roles, but the nature of these roles has not been examined nor questioned in our field.
-In this paper we formulate a theoretical framework to address this issue.
-The framework is based on sociological and organizational studies, and defines roles as sets of expectations placed on the members of the software organization.
+Software professionals divide their work by roles, but to date, researchers have not examined nor questioned the nature of these roles.
+To address this, we formulate a theoretical framework based on sociological and organizational research, in which roles are defined as sets of expectations placed on the members of the software organization.
 
-The paper reports on an empirical case study of two software organizations, which was informed and inspired by the theoretical framework.
+The paper also reports on an empirical case study of two software organizations, which was informed and inspired by this theoretical framework.
 In our study we found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors.
-Furthermore, role definitions evolve through organizational growth and through mismatches between expectations and performance, and people enacting different roles use different media to communicate.
+Furthermore, role definitions evolve through organizational growth and through mismatches between expectations and performance, and people enacting different roles vary in their selection of communication media.
 
-Our contributions should help researchers to frame better studies of division of labour in software groups, and to avoid the pitfalls that come from unwarranted abstractions of software professional roles.
+Our contributions should help researchers to frame better studies of division of labour in software groups and to avoid unwarranted abstractions of software professional roles.
 They also provide practitioners with a conceptual tool to reason about social structure and organizational design.

### 2011-09-27 soulgem.tex
Third draft, Intro.

@@
 \begin{document}
-\title{Working Title: Roles and Expectations in Software Organizations}
+\title{Knowing your role: Varying expectations in software organizations}
 
@@
 
-Poeple divide their labour in order to produce software systems.
-Some write production code, others test it, and still others try to ensure that the product satisfies a real need.
-Researchers and practitioners usually abstract these responsibilities into a number of role labels (such as ``developer,'' ``tester,'' and ``product manager''); we use these abstractions to refer to the set of activities that we expect them to carry out.
-These labels are common everywhere in our academic literature, as well as in textbooks, books for practitioners, human resources postings, blog posts, and everyday conversation.
+People divide their labour in order to produce software systems.
+Some write code, others test it, and still others ensure that the product satisfies a real need.
+Researchers and practitioners abstract these responsibilities into a number of role labels (such as ``developer,'' ``tester,'' and ``product manager''); we use these abstractions to refer to the set of activities that we expect them to carry out.
+These labels are common in our academic literature, as well as in textbooks, books for practitioners, human resources postings, blog posts, and everyday conversation.
 
-Despite this commonplace use, we have no clear indication that the people using these labels refer to the same meanings.
-For instance, what does a ``developer'' actually need to do in order to fulfill her responsibilities?
-What should we expect her to be unable to do?
+Despite their widespread use, we have no clear indication that the people using these labels refer to the same meanings.
+For instance, what does a ``developer'' actually do in order to fulfill her responsibilities?
+What bounds her activity?
 And would the answer be generalizable throughout the many areas in which the software industry operates?
@@
 In some domains, especially the more mature or critical ones, the division of labour is fairly formal and clear.
-Surgeons, anesthesiologists, nurses, and patients generally know what to expect from each other, in and out of the operating room.
+Surgeons, anesthesiologists, and nurses generally know what to expect from each other, in and out of the operating room.
 Knowing that someone is a nurse will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board.
-But although in our own community we assign labels just as directly and unthinkingly as in the medical domain, we do not realize (or such is the premise of our paper) that we may be talking about very different things, based on our varying assumptions, experiences, and cultures.
+But although in our community we assign labels just as effortlessly as in the medical domain, we do not realize (or such is the premise of our paper) that we may be talking about different things, based on our varying assumptions, experiences, and cultures.
 
@@
 Their communication tools may fail to be adopted if they satisfy the needs of some roles but not of others.
-For an abstraction as heavily used as role definitions in software organization, a closer look should prove beneficial.
+For an abstraction as heavily used as roles in software organization, a closer look should prove beneficial.
 
-In this paper we do not attempt to provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success.
-Instead, (a) we examine the concepts of roles, positions, and expectations in software development groups, providing a theoretical framework based on sociological and organizational research, (b) we report the findings from a case study at two software organizations in which we used this theoretical framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.
+In this paper we do not provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success.
+Instead, (a) we examine the concepts of roles, positions, and expectations in software groups, providing a theoretical framework based on interdisciplinary research, (b) we report the findings from a case study at two software organizations in which we used this framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.
 

### 2011-09-27 soulgem.tex
Third draft, Related Work.

@@
 Sociologists and organizational scientists have explored the topic of professional and social roles for a long time.
-Biddle \cite{Biddle1986} provides a good point of entry into the theoretical work in this area, as it teases out the confusion and the overloading of the concept of roles.
+Biddle \cite{Biddle1986} provides a good point of entry into the theoretical work in this area.
 His overview covers five main variations of theoretical thinking: \emph{functional} role theory (roles are functions in a stable social system), \emph{symbolic interactionist} role theory (roles evolve through social interaction, norms only provide broad imperatives), \emph{structural} role theory (roles are patterns of behaviour shared among people in similar positions in the social structure), \emph{organizational} role theory (roles are mainly set by social positions in preplanned, rational settings), and \emph{cognitive} role theory (which focuses on determining which conditions give rise to expectations, how to measure expectations, and what impact do they have on social conduct).
@@
 A similar conceptualization comes from Turner \cite{Turner1956}, who states that roles are sets of norms, where a norm is an expected or appropriate behaviour that should be consistent.
-A role ``is made up of all those norms which are thought to apply to a person occupying a given position.''
-
-% Could take this para out if we need space.
-Although there are roles that everyone in an organization is expected to enact, the roles that are useful in most analyses are those that allow us to differentiate between people.
-As Ashforth and Mael \cite{Ashforth1989} point out, we define ourselves partly in opposition to what others do.
-These distinctions generate both conflicts and opportunities for the members of a social group.
-
-% And if we need *more* space, this is only tangentially relevant.
-Goode \cite{Goode1960} talks about the stresses related with role strain, argues that in general an individual's role obligations are over-demanding, and discusses coping mechanisms commonly used.
-In contrast, Sieber \cite{Sieber1974} claims that the additional demands caused by accumulating roles also bring rewards, and that these rewards may well compensate for the corresponding strain in many cases.
+A role ``is made up of all those norms which are thought to apply to a person occupying a given position.'' Scott \cite{Scott2008} concurs, pointing out that roles are normative expectations of the appropriate goals and activities for particular individuals, and that roles can be formally constructed or emerge informally through interaction.
 
@@
 He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).
-
-% TODO: Check if Indi's papers, Fernando's paper, or those of Andy Begel's, are appropriate in this discussion.
 
@@
 The authors distinguish between the roles and the people and positions that enact them, which is a distinction we advocate for in this paper.
-But several of the roles they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of expectations from our perspective, expectations that are bundled with others to conform a richer, multi-faceted role, as we will discuss in the next section.
+But several of the roles they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of single expectations in our framework below, expectations that are bundled with others to conform a richer, multi-faceted role.
 
@@
 We can also find rich descriptions of some roles and role interactions in software development.
-Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization, and of the reasons for which they deviate from the testing prescriptions in the literature.
-F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a comprehensive report of the consequences of role evolution in a case in which developers attempted to rotate the responsibilities of customer support.
-De Souza \emph{et al.}\ \cite{deSouza2003} describe that problem reports are used as boundary objects \cite{Bowker1999} between people with different roles.
+Martin \emph{et al.}\ \cite{Martin2007} report a detailed picture of the testing role carried out by the developers in a software organization.
+F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a comprehensive report of the consequences of role evolution in a case in which developers attempted to rotate the responsibilities of customer support, and Jensen and Scacchi \cite{Jensen2007} describe the migrations in role responsibilities in several open source projects.
+De Souza \emph{et al.}\ \cite{deSouza2003} look at inter-role communication, and describe that problem reports are used as boundary objects \cite{Bowker1999} between people with different roles, while Marczak and Damian \cite{Marczak2011} explore the backchannel communication between people with different roles.
 
-Finally, Litecky \emph{et al.}\ \cite{Litecky2010} report on the titles offered and the skills required in the job market in computing.
-Their analysis is useful for a study of roles in software organizations, as these titles and skills give us an indication of the roles that managers and human resources professionals perceive as important for their organizations.
-
-% TODO: Add a mention to Sabrina's work
+Finally, Litecky \emph{et al.}\ \cite{Litecky2010} report on the positions offered and the skills required in the job market in computing.
+Their analysis is useful for a study of roles in software organizations, as it gives us an indication of the roles that human resources professionals consider important for their groups.
 

### 2011-09-27 soulgem.tex
Third draft, theoretical framework.

@@
 There is no prototypical developer.
-Upon learning that a colleague is a developer, all the members of the organization place a number of expectations on said colleague.
-Perhaps they expect the developer to spend as much time as possible writing production code, to be responsive to user requests for assistance, to validate the code produced by other developers, or to self-manage.
+Upon learning that a colleague is a developer, her peers place a number of expectations on her.
+Perhaps they expect her to spend as much time as possible writing production code, to be responsive to user requests for assistance, to validate the code produced by other developers, or to self-manage.
 Some colleagues may have no expectations other than that the developer should strive for the well-being of the organization.
-Other colleagues, especially if they interact with the developer on a regular basis, may have many more expectations, and these may be far more concrete: to adhere to the specified API or to put some test hooks in place.
-People outside the formal organization, but with connections or interactions to it, may hold other expectations on her behaviour.
+Other colleagues, especially if they interact with the developer on a regular basis, may have more expectations, and these may be far more concrete: to adhere to the specified API or to put certain test hooks in place.
+People outside the formal organization, but connected to it, may have further expectations.
 And the developer herself, the target of all these expectations, likely has a number of additional expectations on her own conduct: the activities she feels she should be doing to play her part in the organization.
@@
 The resulting set may be straightforward or conflictive (a phenomenon called \emph{role strain} \cite{Goode1960}); it may even be impossible to satisfy in full.
-And since the expectations arise from the members of the group, there is no guarantee that they will match those of other groups: the meaning of a ``developer'' role in one organization may differ from that in another organization, and even within an organization the semantic differences may be profound.
+Since the expectations arise from the group, there is no guarantee that they will match those of other groups: the meaning of a ``developer'' role in one organization may differ from that in another organization, and even within teams in an organization the semantic differences may be profound.
 
 A \textbf{\emph{position}} or a \textbf{\emph{title}}, in turn, is an official statement of a person's goals and responsibilities.
+Positions and their responsibilities are often the explicit articulation of what a powerful fraction of the organization (the managers) expect from their colleagues.
 A role is not a position \cite{Turner1956}, though they often carry the same labels: one can play an architectural role and hold the position of Software Architect.
@@
 For instance, one could enact an architectural role without holding an Software Architect position, and one could hold a Software Architect position without performing any of the activities that others would expect from an architect.
-
-Positions and their responsibilities are often the explicit articulation of what a powerful fraction of the organization (the managers) expect from their colleagues.
-In that sense they are a part of the total set of expectations placed upon organizational roles; an important part (especially if tied to incentives and job evaluations), but still only a part.
 
@@
 Culture plays an important part in setting and steering our expectations.
-In some domains the culture is sufficiently strong to establish an interpretation of a role throughout the population: the role has been institutionalized \cite{DiMaggio1991}.
-In some domains, a combination of intensive training, codes of ethics, professional societies, and familiarization lead us to expect certain behaviours and activities from a role.
-In our domain, these institutional forces allow us to make some general assumptions about software development roles that we can reasonably expect will hold in most cases.
+In some domains cultural norms are sufficiently strong to establish an interpretation of a role throughout the population: the role has been institutionalized \cite{DiMaggio1991}.
+In these cases, a combination of intensive training, codes of ethics, professional societies, and familiarization lead us to expect certain behaviours and activities from a role.
+In our domain, these institutional forces allow us to make some general assumptions that we can reasonably expect will hold in most cases.
 For example, we might be surprised if a tester did not spend at least part of his time preparing or running tests.
@@
 As the evidence of our case study shows, significant disagreements exist even within the same software organizations.
-Determining the reasons for this disagreement is beyond the scope of this paper; we speculate that the relative immaturity of the software industry, the wide variety of fields in which it functions, and the low barriers of entry to the profession all contribute to the phenomenon.
-Whatever the reason, we make no judgment regarding the convenience of this lack of uniformity in role definitions, but we believe it should be explored in greater detail.
+Determining the reasons for this disagreement is beyond the goals of this paper; we speculate that the relative immaturity, the wide scope, and the low barriers of entry to the software industry contribute to the phenomenon.
+We make no judgment regarding the convenience of this lack of uniformity in role definitions, but we believe it should be explored in greater detail.
 
@@
 
-The members of an organization may begin setting their expectations based on institutional roles, but they do not always act according to the expectations placed on them.
-Several reasons explain this disagreement.
-In the case of performance that fail to meet expectations, it could be that the expectation is unrealistic (``testers should find all the bugs in our code''), that the performer lacks skill or care, that the performer is unaware of the expectation (a failure to coordinate or communicate), or that the expectation is in conflict with other expectations (role strain).
-
-It is easy to find examples of poor performance, but so is it to find performance that goes beyond the original expectation.
-In any case, if disagreements recur over time, organization members undergo a shift in their expectations: they learn to expect the ``new normal.'' They adapt their expectations, and since roles are conformed by these expectations, role definitions evolve in any social group based on past performance, despite their institutional status.
+Group members do not always act according to the expectations placed on them, for several reasons.
+Their performance could fail to meet expectations because of unrealistic expectations, unskilled, careless, or unaware performers, or conflicting expectations.
+Their performance could also exceed expectations due to the performer's initiative, talent, or background.
+In any case, if disagreements recur over time, organization members undergo a shift in their expectations: they learn to expect the ``new normal.'' They adapt their expectations, and since roles are conformed by these expectations, role definitions evolve as a result of past performance, despite their institutional status.
 
@@
 With growth comes specialization, and specialization consists of narrower role definitions.
-Therefore, other things being equal, we expect larger organizations to have a greater degree of role specialization.
+Therefore, other things being equal, we expect growing organizations to trend towards a greater degree of role specialization.
 

### 2011-09-27 soulgem.tex
Third draft, Study Design and Execution.

@@
 
-\section{Method and Study Execution}
+\section{Study Design and Execution}
 
-\subsection{Study Design}
-
-We performed a qualitative empirical study based on the theoretical framework presented above to improve our understanding of roles in software organizations.
+We performed an empirical study based on our theoretical framework to improve our understanding of roles in software organizations.
 We chose to follow Yin's case study methodology \cite{Yin2003}.
@@
 
-We formulated eight research questions to guide our data collection and analysis.
-For each of them, we also formulated a corresponding proposition based on the theoretical framework we delineate above.
+We formulated six research questions to guide our data collection and analysis. For each of them, we formulated a corresponding proposition of what we expected to find.
 
@@
 \emph{Proposition:} We do not expect to be able to provide a simple classification of roles.
-However, we hypothesize that some expectations often come together, and that these patterns will form a partial picture of roles in software development.
+However, some expectations often come together, and these patterns may form a partial picture of roles in software development.
 
@@
 
-\textbf{\emph{RQ3:}} What is the logic (what are the factors) behind the particular clusters of expectations found in software organizations?
+\textbf{\emph{RQ3:}} What is the logic behind the particular clusters of expectations found in software organizations?
 
@@
 
-\emph{Proposition:} There is a significant divergence between roles and positions, and the reason is that position is an institutional factor, as specified above, and there are many other factors that contribute to the shaping of a role.
+\emph{Proposition:} There is a significant divergence between roles and positions, and the reason is that \emph{position} describes at best a subset of the expectations shaping a role.
 
-\textbf{\emph{RQ5:}} How do roles change over time in a software organization?
+\textbf{\emph{RQ5:}} How do roles change over time in software groups?
 
@@
 
-\subsection{Study Execution and Analysis}
+We collected field data from two software organizations in the area of Victoria, Canada.
+Our case selection was opportunistic: we wanted software development groups whose headquarters were in our vicinity.
+We negotiated access to both sites in return for feedback on their development practices.
 
-We collected field data from two software organizations in the area of Victoria, Canada.
-Our case selection was opportunistic: we wanted groups that develop software as their main activity, and whose headquarters were located in our vicinity.
-We contacted managers in both organizations and negotiated access to their sites in return for feedback on their development practices.
-
-We carried out our case study during the Winter and Spring of 2011.
-The first author spent five weeks performing observations, three weeks at the first and two at the second site, sitting at a desk in their development areas, attending to several kinds of meetings (stand-up, product, all-hands, and others), and exploring their issue tracking and project estimation systems.
+The first author spent five weeks performing field observations, three weeks at the first site and two at the second, sitting at a desk in their development areas, attending to several kinds of meetings, and exploring their issue tracking and project estimation systems.
 
@@
 In both sites the number of interviewed corresponded to about 50\% of the total number of organization members at the time.
-
-Following Yin's \cite{Yin2003} suggestions, we analyzed all the observation and interview data examining whether it supported or challenged our study propositions.
-We were particularly interested in evidence that did not fit with our theoretical framework, in order to refine it for future studies.
 

### 2011-09-27 soulgem.tex
Third draft, Details from the Field Sites.

@@
 
-\subsection{Gallium}
+\emph{Gallium} was the first organization we studied.\footnote{For confidentiality reasons, the names of both organizations are fictional, and some details have been obfuscated.} It is a biotechnology software company that offers products commonly used by academic groups, pharmaceutical companies, and corporations in the sector.
+At the time of our observations it employed about 45 people.
+Most of them worked at the Victoria office, though several managers and staff worked elsewhere in North America and Europe.
 
-Gallium was the first organization we studied.\footnote{For confidentiality reasons, the names of all organizations and people in this discussion are fictional, and some details not central to our findings have been obfuscated.} It is a biotechnology software organization headquartered in Victoria; it offers specialized software products commonly used by academic groups, pharmaceutical companies, and corporations in the biotechnology sector.
-At the time of our observations it employed about 45 people.
-Most of them worked at the Victoria office, while several managers had their offices in the American West Coast, and the company also had a few agents, mostly in sales, in Europe.
-
-% We might not need all this detail...
-The development group included 17 people, including the Director of Software Development,\footnote{All of the labels uppercased in this discussion refer to the positions that these individuals hold, not to the roles they enact.} two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster.
-The development group was divided in four teams, although during our observations one of the teams was being phased out and its members were transitioning into another team.
-The largest team, Alpha, had eight people (a Team Lead, five Developers, and two QA), and was in charge of maintaining and iterating on the main product of the organization.
-The second largest, Beta, had five people, including the incoming employees (a Team Lead, the two Tech Leads, one Developer, and one QA), and was in charge of recreating and streamlining the workflow and user interface of the main product, without being burdened by its legacy code.
-The third team, Gamma, (consisting of a Team Lead and a Developer) focused on generating custom scripts and solutions for particular customers, using their own product's API.
-The fourth team, Delta, (its members already accounted for above) was wrapping up on its customer obligations for a secondary product.
-Organizationally, the Buildmaster was not associated with any one team, although he interacted mostly with the first and largest team.
+The development group was formed by 17 people, including the Director of Software Development,\footnote{All of the labels uppercased in this discussion refer to the \emph{positions} that these individuals hold, not to the roles they enact.} two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster.
+It was divided in four teams, although one of them was being phased out and its members were transitioning into another team.
+The largest team, Alpha, had nine people and was in charge of maintaining and iterating on the main product of the organization.
+The second largest, Beta, had five people and was in charge of recreating and streamlining the workflow and user interface of the main product, without being burdened by its legacy code.
+The third team, Gamma, consisted of two people and focused on generating custom scripts and solutions for particular customers, using the product's API.
+The fourth team, Delta, was wrapping up on its customer obligations for a secondary product.
 Everyone in the development group, except for its Director, worked in the same shared area: an open space with cubicles arranged in such a way as to demarcate smaller areas functioning as team rooms.
@@
 Several groups interacted with or supported the development group.
-There was a Product Definition group formed by three people (two Product Managers and a Product Design Lead), a Program Manager, a Project Management group (one Project Management Lead and two Project Managers), a Customer Support group (one Lead and two to three Customer Support Analysts), several Field Specialists (numbers varied during our observations, but were never more than four), a Systems Administrator, several Sales Executives, a Graphics Designer, an Office Administrator, and eight other high-ranking staff, including the CEO and four VPs.
-Among these people most of those who worked in the Victoria headquarters did so in personal offices along the edges of the development group's cubicle area, or in a second cubicle area connected to the first.
-An exception was one of the Product Managers, who sat with the Beta Team to be able to provide feedback more rapidly.
+There were Product, Program, and Project Managers, Customer Support Analysts, Field Specialists, Sales Executives, and others.
+These people largely worked on a separate area in the company's offices.
 
 Our observations began at an unusual point in the organization's history, as it had recently gone through a round of layoffs.
-While this was an unfortunate situation for the organization, it allowed us to explore in greater detail the mechanics of shifting expectations across the organizational structure, and the employees' negotiation of the ``new normal'' at their offices.
+While this was an unfortunate situation for the organization, it allowed us to explore in greater detail the mechanics of shifting expectations across the organizational structure and the employees' negotiation of the ``new normal'' at their offices.
 
 The development group followed a lifecycle process loosely based on Scrum \cite{Schwaber2001}.
-The Alpha Team has a six-week release cycle, while Beta structured its work on shorter (two-week) cycles, but for internal purposes only, as it had not yet made any releases of its software by the time our observations ended.
-All teams had a daily stand-up meeting, and the Team and Tech Leads, the Buildmaster, and the Director had a second daily stand-up meeting.
+The Alpha Team has a six-week release cycle, while Beta structured its work on shorter (two-week) cycles, but for internal purposes only, as it had not yet made any releases of its software.
+All teams had daily stand-up meetings, and the larger teams paid attention to Agile metrics such as burndown and velocity.
 The Team Leads functioned as the Scrum Masters, but the Product Owner was not defined consistently.
-The group also deviated from the simple backlogs and story cards practices (requirements were often specified in greater detail, and they were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support), but it paid attention to Agile metrics such as burndown charts and velocity.
+The group also deviated from the simple backlogs and story card practices: their requirements were often specified in greater detail, and they were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support.
 
-\subsection{Cobalt}
-
-Cobalt was the second organization we studied.
+\emph{Cobalt} was the second organization we studied.
 In contrast to Gallium, it is publicly funded.
-It is a computing and data processing and storage unit servicing Canadian and international scientists in one particular research field.
+It is a computing and data processing and storage unit servicing Canadian and international scientists.
 It is part of a research institute, but except for the sharing of support and hardware staff, it has a limited interaction with the rest of the institute.
@@
 
-Cobalt is divided in three subgroups: Software, Operations, and Science.
+The organization is divided in three subgroups: Software, Operations, and Science.
 The Software group has 9 members, including the head of the group.
 Again in contrast with Gallium, the developers in Cobalt are not divided in teams.
-Nominally at least, all may work and request assistance for any of the projects of the group (as we will see, there was an explicit push to achieve this organizational structure in response to concerns with the more ``siloed'' structure of the past).
-However, every project has a Team Lead, and the group as a whole has five Subsystem Leads---a subsystem being an area of interest and expertise for the group, such as Web Services or User Interface---that coordinate the work of those subsystems for each project.
+Nominally at least, all may work and request assistance for any of the projects of the group.
+However, every project has a Lead, and the group as a whole has five Subsystem Leads---a subsystem being an area of expertise for the group, such as Web Services or User Interface---that coordinate the work of those aspects for each project.
 There are no QA specialists in this group; developers are expected to test their work and that of their peers.
 
-The Operations group supports Software in coordinating, configuring, and ensuring the appropriate execution of its projects and processes.
-It includes the head of the group, a Database Administrator, and two Software Specialists.
-The Science group is formed by researchers.
-There were seven scientists at the time of our observations.
-As the name indicates, they are tasked with doing scientific research, not with writing production code.
-However, they often depend on the work of the Software group, they have a significant say in the direction that the Software group takes, and some of them do develop script code.
-Although officially the Science group, like the other two, has a head, his functions are more akin to those of a coordinator than a leader.
-The members of the Science group have considerably more autonomy in determining their activities.
+The Operations group has four employees, and it supports Software in coordinating, configuring, and ensuring the appropriate execution of its projects and processes.
+The Science group is formed by researchers; seven at the time of our observations.
+As the name indicates, they are tasked with doing scientific research, not with writing code.
+However, they have a significant say in the direction that the Software group takes, and some of them do develop script code.
 Finally, there are two staff employees in the group that do not form part of any team in particular: the Group Leader, and a Project Administrator.
@@
 
-For a few months since and including the period of our observations, some group members had been experimenting with a few Agile practices.
+For months since and including the period of our observations, some group members had been experimenting with a few Agile practices.
 They introduced daily stand-up meetings in their projects, as well as continuous integration testing, but have had only partial success in getting test-driven development or code reviews adopted in the group.
-And although they work in one-week iterations, there is no push for a release at the end of each cycle; the iterations are merely used as an opportunity to set the priorities for the next week.
+And although they work in one-week iterations, there is no release at the end of each cycle; the iterations are merely used as an opportunity to set the priorities for the next week.
 

### 2011-09-28 expectationPeopleTable.tex
Some revisions to Findings section.
Third draft, expectations table.

@@
 \textbf{Technical} 
+&Write new-feature code& \0&\0&\2\2&\1\1&\1\2\0&\2\1&\1\2\0\0\0\1&\2\2&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0 \\
+\textbf{Core}
 &Write bug-fix code& \0&\0&\1\1&\1\1&\1\2\0&\2\2&\1\1\1\0\1\0&\2\1&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0 \\
-\textbf{Core}
 &Maintain legacy code& \0&\0&\0\1&\2\1&\2\0\0&\2\2&\2\0\2\0\2\2&\0\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0  \\
-&Write new-feature code& \0&\0&\2\2&\1\1&\1\2\0&\2\1&\1\2\0\0\0\1&\2\2&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0 \\
-&Write customization (or script) code& \0&\0&\0\0&\0\0&\0\0\2&\0\0&\0\0\0\2\0\0&\0\0&\0\0\0&\0&\0\0\0&\1&\0\0&\0&\0\2\0\0&\0\0\\
+&Write customization or script code& \0&\0&\0\0&\0\0&\0\0\2&\0\0&\0\0\0\2\0\0&\0\0&\0\0\0&\0&\0\0\0&\1&\0\0&\0&\0\2\0\0&\0\0\\
 %& \\
@@
 \textbf{Technical}
-& Design user interface&\0&\0&\2\1&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\2&\0\0\0&\0&\1\2\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+&Write low-level (or unit) test code&\0&\0&\1\1&\0\1&\1\1\1&\1\1&\1\1\0\1\1\1&\0\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
 \textbf{Periphery}
-& Write documentation&\0&\0&\0\1&\0\1&\0\0\1&\0\1&\1\0\1\1\0\0&\0\0&\1\0\1&\0&\2\0\2&\0&\0\0&\0&\0\0\0\0&\0\0\\
-&Write low-level (or unit) test code&\0&\0&\1\1&\0\1&\1\1\1&\1\1&\1\1\0\1\1\1&\0\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
-& Write Build Code and prepare a release&\0&\0&\0\0&\0\0&\0\0\0&\1\0&\0\0\0\0\0\0&\0\0&\0\0\2&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
-& Write experimental (or exploratory) code&\0&\0&\1\2&\0\2&\0\0\0&\0\0&\0\0\0\0\2\0&\1\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
 & Write high-level (or integration) test code&\0&\0&\0\0&\1\0&\0\0\0&\2\0&\0\0\0\0\0\0&\0\0&\2\1\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Write build code and prepare a release&\0&\0&\0\0&\0\0&\0\0\0&\1\0&\0\0\0\0\0\0&\0\0&\0\0\2&\2&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Write experimental or exploratory code&\0&\0&\1\2&\0\2&\0\0\0&\0\0&\0\0\0\0\2\0&\1\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
 & Write configuration and implementation code&\0&\0&\0\0&\0\0&\0\0\1&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\2\0\0&\0\0\\
 & Write and maintain tools to aid in development&\0&\0&\0\0&\0\1&\0\0\0&\1\0&\0\0\0\0\0\0&\0\0&\0\0\2&\1&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Write documentation&\0&\0&\0\1&\0\1&\0\0\1&\0\1&\1\0\1\1\0\0&\0\0&\1\0\1&\0&\2\0\2&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Design user interface&\0&\0&\2\1&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\2&\0\0\0&\0&\1\2\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
 %& \\
@@
 & Elicit and document requirements&\0&\0&\0\0&\0\1&\0\0\0&\1\1&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\2\2\2&\0&\0\0&\0&\0\0\0\2&\1\2\\
+& Prioritize requirements&\1&\0&\1\0&\0\1&\1\0\0&\2\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\2&\2&\0\0&\0&\0\0\0\0&\2\1\\
+& Provide context to non-domain experts (user proxy)&\0&\0&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\0&\2&\0\0&\0&\0\2\0\2&\2\0\\
 & Determine if product meets quality goals&\1&\0&\1\1&\0\0&\1\0\0&\1\0&\0\1\0\0\0\0&\1\1&\2\2\1&\2&\0\2\0&\1&\0\0&\0&\0\0\0\0&\1\1\\
-& Set roadmap for system (prioritize requirements)&\1&\0&\1\0&\0\1&\1\0\0&\2\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\2&\2&\0\0&\0&\0\0\0\0&\2\1\\
-& Provide domain context to non-domain experts (user proxy)&\0&\0&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\0&\2&\0\0&\0&\0\2\0\2&\2\0\\
 %& \\
@@
 & Provide historical perspective on design decisions&\0&\0&\2\1&\0\2&\0\0\0&\0\0&\2\0\2\0\0\0&\0\0&\0\1\0&\0&\2\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
-& Provide guidance (mentorship) to newcomers on technical issues&\0&\0&\1\0&\0\0&\0\0\2&\2\2&\2\0\0\0\0\0&\0\1&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
+& Provide guidance or mentorship on technical issues&\0&\0&\1\0&\0\0&\0\0\2&\2\2&\2\0\0\0\0\0&\0\1&\0\0\0&\0&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
 %& \\
@@
  \textbf{Management}
+& Coordinate a product release or implementation&\1&\1&\1\0&\1\1&\2\0\0&\2\0&\0\0\0\0\0\0&\0\0&\0\1\2&\2&\1\2\2&\0&\2\2&\1&\0\1\0\0&\0\0\\
 & Assign low-level tasks (tickets) to individuals&\0&\1&\1\0&\0\0&\2\0\1&\1\0&\0\0\0\0\0\0&\0\0&\0\0\0&\2&\0\0\1&\0&\1\1&\0&\1\2\1\0&\0\0\\
 & Assign high-level tasks (projects) to individuals&\2&\2&\0\0&\0\1&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\1&\0&\0\0&\0&\0\0\0\0&\0\0\\
-& Coordinate a product release or implementation&\1&\1&\1\0&\1\1&\2\0\0&\2\0&\0\0\0\0\0\0&\0\0&\0\1\2&\2&\1\2\2&\0&\2\2&\1&\0\1\0\0&\0\0\\
 & Provide estimates on remaining or potential tasks&\1&\1&\2\1&\1\0&\2\0\1&\2\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\1\1&\0&\2\2&\1&\0\0\0\0&\0\0\\
 & Connect an issue with the person with appropriate expertise&\1&\1&\0\0&\0\1&\2\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\0&\0&\1\1&\0&\1\2\2\0&\1\0\\
-& Handle obstacles and interruptions for technical teams (PoC)&\1&\2&\0\0&\0\0&\2\1\1&\0\1&\0\0\0\0\0\0&\0\0&\0\0\0&\2&\0\0\0&\0&\2\2&\2&\0\0\0\0&\0\0\\
+& Handle obstacles and interruptions for technical teams&\1&\2&\0\0&\0\0&\2\1\1&\0\1&\0\0\0\0\0\0&\0\0&\0\0\0&\2&\0\0\0&\0&\2\2&\2&\0\0\0\0&\0\0\\
 %& \\
@@
 \textbf{Organizational}
+& Provide domain legitimacy to organization&\0&\0&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\0&\2&\0\0&\0&\0\0\0\0&\0\0\\
+\textbf{External Interface}
+& Provide technical legitimacy to organization&\2&\1&\2\0&\0\2&\0\0\0&\0\1&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
 & Receive and respond to user feedback&\0&\2&\0\0&\0\0&\0\1\0&\1\1&\0\0\0\0\0\0&\0\1&\0\0\0&\1&\0\0\0&\2&\2\2&\0&\2\2\2\0&\1\1\\
-\textbf{External Interface}
+& Advocate for the resolution of user/customer problems&\1&\2&\0\0&\0\0&\1\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\1\0&\0&\2\2&\0&\2\2\2\1&\2\2\\
 & Obtain new users or sources of funding&\0&\2&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\0\0&\1&\0\0&\0&\0\0\0\2&\0\0\\
-& Provide domain legitimacy to organization&\0&\0&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\2\0&\2&\0\0&\0&\0\0\0\0&\0\0\\
-& Provide technical legitimacy to organization&\2&\1&\2\0&\0\2&\0\0\0&\0\1&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\0\0&\0&\0\0\0\0&\0\0\\
-& Advocate for the resolution of user/customer problems&\1&\2&\0\0&\0\0&\1\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\1\0&\0&\2\2&\0&\2\2\2\1&\2\2\\
 & Represent the organization's interests to funders/customers&\0&\2&\0\0&\0\1&\0\1\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\0&\0\1\0&\2&\0\0&\0&\0\0\0\1&\1\1\\
@@
 \textbf{Organizational}
+& Set up and modify organizational structure&\2&\2&\0\0&\0\1&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\1&\1&\0\0&\0&\0\0\1\0&\0\0\\
+\textbf{Management}
+& Communicate organizational status and news&\2&\2&\0\0&\0\0&\1\0\1&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\1\1&\1&\0\0\1\0&\0\0\\
 & Deal with personnel issues&\2&\2&\0\1&\0\0&\1\1\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\1&\0\0&\2&\0\0\1\0&\0\0\\
-\textbf{Management}
+& Deal with financial, clerical, or sysadmin issues&\0&\1&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\0\0&\2&\0\0\0\0&\0\0\\
 & Deal with otherwise unattended issues&\1&\1&\0\0&\0\1&\1\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\2\0&\1&\0\1\0&\1&\0\0&\2&\0\0\0\0&\0\0\\
-& Set up and modify organizational structure&\2&\2&\0\0&\0\1&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\1&\1&\0\0&\0&\0\0\1\0&\0\0\\
-& Communicate organizational status and news&\2&\2&\0\0&\0\0&\1\0\1&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\1\1&\1&\0\0\1\0&\0\0\\
-& Deal with financial, clerical, or system administration issues&\0&\1&\0\0&\0\0&\0\0\0&\0\0&\0\0\0\0\0\0&\0\0&\0\0\0&\1&\0\0\0&\0&\0\0&\2&\0\0\0\0&\0\0\\
 %& \\
@@
 \end{tabular}
-\caption{Assignment of expectations to people in our field sites.
-This matrix only includes our interviewees in both organizations; about half of the total number of organization members.
-Interviewees are clustered approximately by labels, despite label variations.
+\caption{{\small\label{tableExpectations}Expectation assignments.
+This table only includes our interviewees, for which we had the most information.
+They are clustered approximately by labels, despite label variations.
 Within each label cluster there is a further subdivision by organization (``G'' stands for Gallium, ``C'' for Cobalt).
-The colouring of each cell is necessarily subjective, but it corresponds to a rough consensus view of the expectations placed on each individual.}
+The shading of each cell is subjective, but it corresponds to a rough consensus among all the interviewed.}}
 \end{table*}

### 2011-09-28 soulgem.tex
Third draft, updates up to section IV from Cassie.

@@
 
-People divide their labour in order to produce software systems.
-Some write code, others test it, and still others ensure that the product satisfies a real need.
-Researchers and practitioners abstract these responsibilities into a number of role labels (such as ``developer,'' ``tester,'' and ``product manager''); we use these abstractions to refer to the set of activities that we expect them to carry out.
-These labels are common in our academic literature, as well as in textbooks, books for practitioners, human resources postings, blog posts, and everyday conversation.
+People divide their labour when producing software systems: certain people write code, others test it, and some ensure that the product satisfies a real need.
+Researchers and practitioners abstract these responsibilities into a number of role labels, such as ``developer'', ``tester'', and ``product manager'', and we use these abstractions to refer to the set of activities that we expect them to carry out.
+These labels are common in academic literature, as well as in textbooks, books for practitioners, human resources postings, blog posts, and everyday conversation.
 
-Despite their widespread use, we have no clear indication that the people using these labels refer to the same meanings.
+Despite their widespread use, we have no clear indication that the people using these labels refer to the same definitions.
 For instance, what does a ``developer'' actually do in order to fulfill her responsibilities?
 What bounds her activity?
-And would the answer be generalizable throughout the many areas in which the software industry operates?
+And would the answer be generalizable throughout the software industry?
 
-In some domains, especially the more mature or critical ones, the division of labour is fairly formal and clear.
-Surgeons, anesthesiologists, and nurses generally know what to expect from each other, in and out of the operating room.
-Knowing that someone is a nurse will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board.
-But although in our community we assign labels just as effortlessly as in the medical domain, we do not realize (or such is the premise of our paper) that we may be talking about different things, based on our varying assumptions, experiences, and cultures.
+In some domains, especially ones that are more mature or critical, the division of labour is fairly formal and clear.
+In the medical field, surgeons, anesthesiologists, and nurses generally know what to expect from each other, in and out of the operating room.
+Knowing that someone is a nurse will bring out a set of behavioural expectations that have a good probability of being shared across the board.
+But although we assign labels just as effortlessly as in the medical field, we do not realize that we may be talking about different things, based on our varying assumptions, experiences, and cultures.
 
-This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has not been studied in depth so far, and it has unexplored repercussions for researchers and practitioners alike.
+This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has yet to be studied in depth, and it has unexplored repercussions for researchers and practitioners alike.
 Poorly defined roles may lead to dysfunction in a project team, and thus negatively impact the quality of the software.
-Researchers may prescribe practices to be carried out by one role without realizing that the role is usually shared and spread out among a group of people.
+Researchers may prescribe that practices be carried out by one role without realizing that the role is usually shared and spread out among a group of people.
 Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him \cite{Dagenais2010}.
 Their communication tools may fail to be adopted if they satisfy the needs of some roles but not of others.
-For an abstraction as heavily used as roles in software organization, a closer look should prove beneficial.
+For an abstraction as heavily used as roles in software organizations, a closer look should prove beneficial.
 
-In this paper we do not provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success.
+In this paper, we do not provide a rigid definition of roles in software organizations; in fact, we argue that such a task has a poor prospect of success.
 Instead, (a) we examine the concepts of roles, positions, and expectations in software groups, providing a theoretical framework based on interdisciplinary research, (b) we report the findings from a case study at two software organizations in which we used this framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.
@@
 His overview covers five main variations of theoretical thinking: \emph{functional} role theory (roles are functions in a stable social system), \emph{symbolic interactionist} role theory (roles evolve through social interaction, norms only provide broad imperatives), \emph{structural} role theory (roles are patterns of behaviour shared among people in similar positions in the social structure), \emph{organizational} role theory (roles are mainly set by social positions in preplanned, rational settings), and \emph{cognitive} role theory (which focuses on determining which conditions give rise to expectations, how to measure expectations, and what impact do they have on social conduct).
-Although there are significant philosophical and methodological differences between role theories, they are compatible on many essential points.
+Although there are significant philosophical and methodological differences between role theories, they are compatible on many points.
 As Biddle points out, ``most versions of role theory presume that expectations are the major generators of roles, that expectations are learned through experience, and that persons are aware of the expectations they hold.''
@@
 Practices also prescribe roles, but in a more minute detail and, usually, for a shorter term.
-Thus in a Fagan inspection \cite{Fagan1976} we have a Moderator, a Reader, a Coder, and several Reviewers.
+Thus in a Fagan inspection \cite{Fagan1976}, we have a Moderator, a Reader, a Coder, and several Reviewers.
 
-Recent work regularly discusses roles in interesting ways, but without a solid or consistent theoretical framework.
+Recent work discusses roles in interesting ways, but without a solid or consistent theoretical framework.
 Hoda \emph{et al.}\ \cite{Hoda2010} identified six ``informal roles'' that team members adopt in self-organizing teams.
-The authors distinguish between the roles and the people and positions that enact them, which is a distinction we advocate for in this paper.
-But several of the roles they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of single expectations in our framework below, expectations that are bundled with others to conform a richer, multi-faceted role.
+The authors distinguish between roles and the people that enact them, which is a distinction we advocate for in this paper.
+But several of the roles they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of single expectations in our framework below---expectations that are bundled with others to produce a richer, multi-faceted role.
 
@@
 
-Other researchers have attempted to study roles structurally, as they arise by social network interactions.
-For instance, Toral \emph{et al.}\ \cite{Toral2010} studied ``middle-men'' or ``brokers,'' and Ricca and Marchetto \cite{Ricca2010} examine the role of ``heroes,'' in open source projects.
-Their structural analysis allows them to scale to large projects and electronic repositories, at the expense of detail.
+Other researchers have attempted to study roles structurally, through social network analysis.
+For instance, Toral \emph{et al.}\ \cite{Toral2010} studied ``middle-men'' or ``brokers'', and Ricca and Marchetto \cite{Ricca2010} examine the role of ``heroes'', in open source projects.
+Their structural analysis allows them to scale to large projects and electronic repositories.
 To our knowledge, more sophisticated analyses of roles from a structural role theory perspective, such as the sociological blockmodelling work of Breiger and Pattison \cite{Breiger1986}, has not been carried out in our domain.
@@
 
-Finally, Litecky \emph{et al.}\ \cite{Litecky2010} report on the positions offered and the skills required in the job market in computing.
-Their analysis is useful for a study of roles in software organizations, as it gives us an indication of the roles that human resources professionals consider important for their groups.
+Finally, Litecky \emph{et al.}\ \cite{Litecky2010} report on the positions offered and the skills required in the computing job market.
+Their analysis is useful for a study of roles in software organizations as it provides us with an indication of the roles that human resources professionals consider important in their groups.
 
@@
 
-\subsection{Roles, positions, and labels}
+\subsection{Roles, Positions, and Labels}
 
-A \textbf{\emph{role}} is a set of expectations of behaviour placed on a person given their position, status, history, or preference \cite{Turner1956,Gordon1976,Rizzo1970}.
-This entails that roles are \emph{social constructions} \cite{Berger1967}: in a real sense, a ``developer'' is whatever the relevant social group determines that a developer should be.
+A \textbf{\emph{role}} is a set of behavioural expectations placed on a person given their position, status, history, or preference \cite{Turner1956,Gordon1976,Rizzo1970}.
+This entails that roles are \emph{social constructions} \cite{Berger1967}.
+For example, a ``developer'' is whatever the relevant social group determines that a developer should be.
 There is no prototypical developer.
@@
 Some colleagues may have no expectations other than that the developer should strive for the well-being of the organization.
-Other colleagues, especially if they interact with the developer on a regular basis, may have more expectations, and these may be far more concrete: to adhere to the specified API or to put certain test hooks in place.
+Other colleagues, especially if they interact with the developer on a regular basis, may have more expectations that may be far more concrete, such as to adhere to the specified API or to put certain test hooks in place.
 People outside the formal organization, but connected to it, may have further expectations.
-And the developer herself, the target of all these expectations, likely has a number of additional expectations on her own conduct: the activities she feels she should be doing to play her part in the organization.
+And the developer herself, the target of all these expectations, likely has a number of additional expectations of her own conduct: the activities she feels she should be doing to play her part in the organization.
 
-These expectations, taken together as a collection, conform the role.
-The resulting set may be straightforward or conflictive (a phenomenon called \emph{role strain} \cite{Goode1960}); it may even be impossible to satisfy in full.
-Since the expectations arise from the group, there is no guarantee that they will match those of other groups: the meaning of a ``developer'' role in one organization may differ from that in another organization, and even within teams in an organization the semantic differences may be profound.
+This collection of expectations define the role.
+The resulting set may be straightforward or conflictive (a phenomenon called \emph{role strain} \cite{Goode1960}), and it may be impossible to satisfy it in full.
+Since the expectations arise from the group, there is no guarantee that they will match those of other groups; the meaning of a ``developer'' role in one organization may differ from that in another organization, and even within teams in an organization the semantic differences may be profound.
 
 A \textbf{\emph{position}} or a \textbf{\emph{title}}, in turn, is an official statement of a person's goals and responsibilities.
-Positions and their responsibilities are often the explicit articulation of what a powerful fraction of the organization (the managers) expect from their colleagues.
-A role is not a position \cite{Turner1956}, though they often carry the same labels: one can play an architectural role and hold the position of Software Architect.
+Positions and their responsibilities are often the explicit articulation of what the managers expect from their employees.
+A role is not a position \cite{Turner1956}, though they often carry the same labels.
+One can play an architectural role and hold the position of Software Architect.
 A person's title plays a part in determining what she will feel compelled to do, and what she believes she is expected to do, but it is only one element of her role.
@@
 As indicated above, we often use \textbf{\emph{labels}} to refer to roles.
-A fitting label helps us convey the essence of the role to our interlocutors.
-But some roles have ill-fitting labels, and others have no labels or easy ways to refer to them in concrete terms.
+A label helps us convey the essence of the role, but some roles have ill-fitting labels, and others have no labels or easy ways to refer to them in concrete terms.
 We must remember that neither the label nor the formal position completely define the role.
 
-\subsection{Institutionalized roles}
+\subsection{Institutionalized Roles}
 
 Culture plays an important part in setting and steering our expectations.
-In some domains cultural norms are sufficiently strong to establish an interpretation of a role throughout the population: the role has been institutionalized \cite{DiMaggio1991}.
-In these cases, a combination of intensive training, codes of ethics, professional societies, and familiarization lead us to expect certain behaviours and activities from a role.
+In some domains, cultural norms are strong enough to establish an interpretation of a role throughout the population (the role has been institutionalized \cite{DiMaggio1991}).
+In these cases, a combination of intensive training, ethical codes, professional societies, and familiarization lead us to expect certain behaviours and activities from a role.
 In our domain, these institutional forces allow us to make some general assumptions that we can reasonably expect will hold in most cases.
@@
 
-However, the forces from the larger culture have not yet enforced a singular view of what the different roles in a software organization consist of in any useful detail, or even which roles exist.
+However, the forces of the larger culture have not yet enforced a singular view of what the different roles in a software organization consist of in any useful detail, or even which roles exist.
 As the evidence of our case study shows, significant disagreements exist even within the same software organizations.
@@
 
-\subsection{Role evolution}
+\subsection{Role Evolution}
 
@@
 Their performance could also exceed expectations due to the performer's initiative, talent, or background.
-In any case, if disagreements recur over time, organization members undergo a shift in their expectations: they learn to expect the ``new normal.'' They adapt their expectations, and since roles are conformed by these expectations, role definitions evolve as a result of past performance, despite their institutional status.
+In any case, if disagreements recur over time, organization members undergo a shift in their expectations: they learn to expect the ``new normal''.
+They adapt their expectations, and since roles are conformed by these expectations, role definitions evolve as a result of past performance, despite their institutional status.
 
-Another reason for role evolution arises from changes in the expectations that are needed to meet the organization's goals.
-An involvement in a new market niche may bring the need to coordinate the old products with the new one, to transfer expertise, and so on.
-One generally accepted cause of shifts in expectations is organizational growth \cite{Blau1971,Haveman1993}.
+Another reason for role evolution arises from changes in the expectations needed to meet the organization's goals.
+An involvement in a new market niche may bring the need to coordinate old products with a new one, to transfer expertise, and so on.
+One generally accepted cause of expectation shifts is organizational growth \cite{Blau1971,Haveman1993}.
 With growth comes specialization, and specialization consists of narrower role definitions.
@@
 
-We performed an empirical study based on our theoretical framework to improve our understanding of roles in software organizations.
-We chose to follow Yin's case study methodology \cite{Yin2003}.
-We executed an explanatory two-case case study with two embedded units of analysis (software organizations and roles).
+Based on the theoretical framework presented above, we performed an empirical study to improve our understanding of roles in software organizations, choosing to follow Yin's case study methodology \cite{Yin2003}.
+We executed an explanatory case study of two organizations, using two embedded units of analysis (software organizations and roles).
 
@@
 
-\emph{Proposition:} Roles vary considerably both between software organizations and within them. The sets of expectations placed on their members are different in significant ways.
+\emph{Proposition:} Roles vary considerably between software organizations and within them. The expectations placed on their members are different in significant ways.
 
@@
 
-\textbf{\emph{RQ6:}} What media do people enacting similar roles use to communicate to each other? What media do they use to communicate with others?
+\textbf{\emph{RQ6:}} What media do people enacting similar roles use to communicate with each other? What media do they use to communicate with others?
 
 \emph{Proposition:} People enacting similar roles use similar communication media.
-They communicate with people enacting different but interfacing roles through other media, which act as a boundary object between them \cite{Bowker1999}.
+They communicate with people enacting different but interfacing roles through other media, and these media act as a boundary object between them \cite{Bowker1999}.
 
-We collected field data from two software organizations in the area of Victoria, Canada.
+We collected field data from two software organizations in the area of Victoria, BC, Canada.
 Our case selection was opportunistic: we wanted software development groups whose headquarters were in our vicinity.
@@
 
-The first author spent five weeks performing field observations, three weeks at the first site and two at the second, sitting at a desk in their development areas, attending to several kinds of meetings, and exploring their issue tracking and project estimation systems.
+The first author spent five weeks performing field observations (three weeks at the first site and two at the second), sitting at a desk in the organizations' development areas, attending several kinds of meetings, and exploring their issue tracking and project estimation systems.
 
-The first author also interviewed people spanning most areas of both organizations.
-He performed 36 interviews, most of them lasting almost one hour.
-In both sites the number of interviewed corresponded to about 50\% of the total number of organization members at the time.
+The first author also interviewed people spanning most areas of each organization.
+He performed 36 interviews, most of them lasting approximately one hour.
+At both sites, the number of employees interviewed corresponded to about 50\% of the total number of group members at the time.
 
@@
 \emph{Gallium} was the first organization we studied.\footnote{For confidentiality reasons, the names of both organizations are fictional, and some details have been obfuscated.} It is a biotechnology software company that offers products commonly used by academic groups, pharmaceutical companies, and corporations in the sector.
-At the time of our observations it employed about 45 people.
+At the time of our observations, it employed about 45 people.
 Most of them worked at the Victoria office, though several managers and staff worked elsewhere in North America and Europe.
 
-The development group was formed by 17 people, including the Director of Software Development,\footnote{All of the labels uppercased in this discussion refer to the \emph{positions} that these individuals hold, not to the roles they enact.} two Tech Leads, three Team Leads, seven Developers, three QA Analysts, and a Buildmaster.
-It was divided in four teams, although one of them was being phased out and its members were transitioning into another team.
-The largest team, Alpha, had nine people and was in charge of maintaining and iterating on the main product of the organization.
-The second largest, Beta, had five people and was in charge of recreating and streamlining the workflow and user interface of the main product, without being burdened by its legacy code.
-The third team, Gamma, consisted of two people and focused on generating custom scripts and solutions for particular customers, using the product's API.
+The development group was formed by 17 people, including the Director of Software Development,\footnote{All of the labels uppercased in this discussion refer to the \emph{positions} that these individuals hold, not to the roles they enact.} 2 Tech Leads, 3 Team Leads, 7 Developers, 3 QA Analysts, and a Buildmaster.
+It was divided into four teams, although one of them was being phased out and its members were transitioning onto another team.
+The largest team, Alpha, had nine people, and was in charge of maintaining and iterating on the organization's main product.
+The second largest, Beta, had five people, and was in charge of recreating and streamlining the workflow and user interface of the main product, without being burdened by its legacy code.
+The third team, Gamma, consisted of two people, and focused on generating custom scripts and solutions for individual customers, using the product's API.
 The fourth team, Delta, was wrapping up on its customer obligations for a secondary product.
@@
 There were Product, Program, and Project Managers, Customer Support Analysts, Field Specialists, Sales Executives, and others.
-These people largely worked on a separate area in the company's offices.
+Most of these people worked on a separate area in the company's offices.
 
@@
 The development group followed a lifecycle process loosely based on Scrum \cite{Schwaber2001}.
-The Alpha Team has a six-week release cycle, while Beta structured its work on shorter (two-week) cycles, but for internal purposes only, as it had not yet made any releases of its software.
+The Alpha Team had a six-week release cycle, while Beta structured its work on shorter (two-week) cycles, but for internal purposes only, as it had not yet made any releases of its software.
 All teams had daily stand-up meetings, and the larger teams paid attention to Agile metrics such as burndown and velocity.
 The Team Leads functioned as the Scrum Masters, but the Product Owner was not defined consistently.
-The group also deviated from the simple backlogs and story card practices: their requirements were often specified in greater detail, and they were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support.
+The group also deviated from the simple backlogs and story card practices: their requirements were often specified in greater detail and they were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support.
 
@@
 In contrast to Gallium, it is publicly funded.
-It is a computing and data processing and storage unit servicing Canadian and international scientists.
+It is a computing, data processing, and information storage unit servicing Canadian and international scientists.
 It is part of a research institute, but except for the sharing of support and hardware staff, it has a limited interaction with the rest of the institute.
-At the time of our observations the group consisted of about 22 people, including visiting scientists that collaborate in some sense with the group's software development activities.
+At the time of our observations, the group consisted of about 22 people, including visiting scientists that collaborated with the group's software development activities.
 All of Cobalt's employees worked at their Victoria offices.
 
-The organization is divided in three subgroups: Software, Operations, and Science.
-The Software group has 9 members, including the head of the group.
-Again in contrast with Gallium, the developers in Cobalt are not divided in teams.
-Nominally at least, all may work and request assistance for any of the projects of the group.
-However, every project has a Lead, and the group as a whole has five Subsystem Leads---a subsystem being an area of expertise for the group, such as Web Services or User Interface---that coordinate the work of those aspects for each project.
-There are no QA specialists in this group; developers are expected to test their work and that of their peers.
+The organization was divided into three subgroups: Software, Operations, and Science.
+The Software group had nine members, including the head of the group.
+Again, in contrast with Gallium, Cobalt's developers were not divided into teams; all may work for and request assistance from any of the projects of the group.
+However, every project had a Lead, and the group as a whole had five Subsystem Leads---a subsystem being an area of expertise for the group, such as Web Services or User Interface---that coordinate the work of those aspects for each project.
+There were no QA specialists in this group; developers were expected to test their work and that of their peers.
 
-The Operations group has four employees, and it supports Software in coordinating, configuring, and ensuring the appropriate execution of its projects and processes.
-The Science group is formed by researchers; seven at the time of our observations.
-As the name indicates, they are tasked with doing scientific research, not with writing code.
-However, they have a significant say in the direction that the Software group takes, and some of them do develop script code.
+The Operations group had four employees, and it supported Software in coordinating, configuring, and ensuring the appropriate execution of its projects and processes.
+The Science group was formed by researchers; seven at the time of our observations.
+As the name indicates, they were tasked with doing scientific research, not with writing code.
+However, they had a significant say in the direction that the Software group took, and some of them developed script code.
 Finally, there are two staff employees in the group that do not form part of any team in particular: the Group Leader, and a Project Administrator.
@@
 For months since and including the period of our observations, some group members had been experimenting with a few Agile practices.
-They introduced daily stand-up meetings in their projects, as well as continuous integration testing, but have had only partial success in getting test-driven development or code reviews adopted in the group.
-And although they work in one-week iterations, there is no release at the end of each cycle; the iterations are merely used as an opportunity to set the priorities for the next week.
+They introduced daily stand-up meetings in their projects, as well as continuous integration testing, but had only partial success in getting test-driven development or code reviews adopted in the group.
+And although they worked in one-week iterations, there was no release at the end of each cycle; the iterations were merely used as an opportunity to set the priorities for the next week.
 

### 2011-09-28 soulgem.tex
Still on third draft. Comments from Cassie on second half of paper.

@@
 \begin{document}
-\title{Knowing your role: Varying expectations in software organizations}
+\title{Setting the Stage for Understanding the Main Roles in Software Organizations}
 
@@
 The paper also reports on an empirical case study of two software organizations, which was informed and inspired by this theoretical framework.
-In our study we found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors.
+In our study, we found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors.
 Furthermore, role definitions evolve through organizational growth and through mismatches between expectations and performance, and people enacting different roles vary in their selection of communication media.
 
-Our contributions should help researchers to frame better studies of division of labour in software groups and to avoid unwarranted abstractions of software professional roles.
+Our contributions should help researchers to frame better studies of division of labour in software groups and to avoid unwarranted abstractions of professional software roles.
 They also provide practitioners with a conceptual tool to reason about social structure and organizational design.
@@
 
-There is no simple classification of roles in the organizations we studied.
-However, we believe that, should we continue studying more groups, some patterns in role definitions would begin to emerge.
+There was no simple role classification in the organizations we studied.
+However, it appears that some patterns in role definitions would emerge with further studies.
 We can see the beginnings of such patterns in Table \ref{tableExpectations}, which shows the expectations placed on the people we interviewed (about half of the total number of organization members).
-The columns in the table represent our interviewees; the rows represent common or significant expectations.
-The reader should note that our list of expectations is necessarily incomplete for the general case.
-It is also somewhat subjective, as expectations in our interviews and observations were not always clearly articulated and our participants did not always agree on their expectation assignments.
+The columns in the table represent our interviewees and the rows represent common or significant expectations.
+The reader should note that the list of expectations is necessarily incomplete for the general case.
+It is also somewhat subjective, as expectations were not always clearly articulated, and our participants did not agree on their expectation assignments in every case.
 
-For instance, working on the technical core of the organizations' products is reserved almost exclusively for developers and their more senior differentiations, technical and team leads.
-However, the leads' expectations are more spread out: technical leads focus more on architecture and design issues, whereas team leads emphasize coordination activities.
-There are also purely organizational roles; their enactors do not need any domain or technical expertise.
-But in both organizations there were key people with more than one kind of expertise (technical, domain, and organizational), and these people seemed vital for the functioning of the organization, as they transitioned between boundaries with ease.
-A representative example is a Gallium Field Specialist (the second column in the Support \& Field cluster), who straddles all three kinds of expertise.
+One pattern concerns the technical core of the organizations' products.
+Work on this area was reserved almost exclusively for developers and their technical and team leads.
+However, the leads' expectations were more spread out: technical leads focused more on architecture and design issues, whereas team leads emphasized coordination activities.
+There were also purely organizational roles that do not need any domain or technical expertise.
+But in both organizations some key people had more than one kind of expertise (technical, domain, and organizational), and these people seemed vital for the functioning of the organization as they transitioned between boundaries with ease.
+A representative example is a Gallium Field Specialist (the second column in the Support \& Field cluster), who straddled all three kinds of expertise.
 Several people in different areas pointed out his relevance; in the words of one informant, ``the company depends on [him] so much he should have armed guards behind him taking him to and from work.''
 
-There are several other patterns in expectation assignment in our data.
-For space reasons we only point out three more.
+Our data shows several other expectation assignment patterns, but for space reasons, we only point out three.
 First, all the system architecture expectations were placed on people still working on the technical core, writing new feature and bug fix code.
 This is in contrast to our anecdotal experiences in larger organizations.
-Second, group members that provide domain legitimacy or credibility were critical at both organizations, and they are expected to influence product development at a high level, but not to tinker with implementation details.
+Second, group members that provided domain legitimacy or credibility were critical at both organizations, and they were expected to influence product development at a high level, but not to tinker with implementation details.
 And third, some expectations, such as writing unit tests and documentation, were weakly spread out through the technical side of the organizations.
-Everyone expects developers to ``play nice'' by testing and documenting their own code, but it is culturally acceptable to ignore these expectations if the urgency of the situation demands it, as it often does.
-
-% TODO: Check build code expectation in Cobalt.
+Everyone expected developers to ``play nice'' by testing and documenting their own code, but it was culturally acceptable to ignore these expectations if the urgency of the situation demands it---as it often does.
 
@@
 
-\emph{Product Management at Gallium:} There are mainly three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers: the two Product Managers and the Product Design Lead.
+\emph{Product Management at Gallium:} There are three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers: the two Product Managers and the Product Design Lead.
 The latter documents the requirements of new functionality, he fulfills a ``historian'' expectation (sharing design rationale with developers and pointing out pitfalls in some approaches), and he is active in shaping the organizational structure, having helped create some groups and practices.
@@
 But their roles were quite different.
-The Team Lead of Alpha would spend about half of his time developing software, and in his other half he meets with customer service representatives, project and product managers, performs estimation tasks, and analyzes the performance of his team.
+The Team Lead of the Alpha Team would spend about half of his time developing software, and in his other half, he meets with customer service representatives, project and product managers, performs estimation tasks, and analyzes the performance of his team.
 A second Team Lead is in charge of a product about to be discontinued, and that responsibility takes a small part of his time.
@@
 \emph{Developers at Gallium and Cobalt:} The quintessential role in the software development literature was one of the hardest to pin down.
-A developer, of course, is expected to write production code---but what else?
+A developer, of course, is expected to write production code, but what else?
 There is a multitude of answers to that question: some developers are expected to perform code reviews, some are expected to provide architectural input, some are expected to carry their own requirements analysis, some are expected to socialize, some are expected to interface with customers.
-None in our sample is expected to perform all of the above.
+None of the developers in our sample were expected to perform all of the above.
 
-\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization.\footnote{Note that, as opposed to the previous examples, this role has no clear label to refer to it.} But the stuff that falls through the cracks was different in both sites, and therefore the expectations were different.
+\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization.\footnote{Note that, as opposed to the previous examples, this role has no clear label to refer to it.} But the stuff that falls through the cracks was different in both sites, and therefore, the expectations were different.
 In Cobalt, much of this referred to sheltering the organization from bureaucratic processes and constraints.
@@
 
-\emph{Project Management at Gallium:} These project managers are a counterexample to our finding, to some extent.
-Although there are distinctions between them (one has a project management leader role, another creates templates and institutes new processes, and all, including a third and more junior project manager, implement them), the similarities in their roles seem to outweigh the differences.
+\emph{Project Management at Gallium:} These Project Managers are a counterexample to our findings, to some extent.
+Although there are distinctions between them (one has a Project Management Leader role, another creates templates and institutes new processes, and all, including a third and more junior Project Manager, implement them), the similarities in their roles seem to outweigh the differences.
 The main reason seems to be that these Project Managers bring a fairly specific and standardized way of doing things with them, in the form of a Project Management Body of Knowledge and certification.
@@
 
-We found that expectations are clustered in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors, and that these kinds of factors can interplay and provide a feedback to each other.
+We found that expectations are clustered in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors, and that these kinds of factors can interplay and provide feedback to each other.
 
-\emph{Institutional factors:} Although the software industry's culture still leaves plenty of room for role variability, it contributes to shape role definitions.
+\emph{Institutional factors:} Although the software industry's culture still leaves plenty of room for role variability, it helps shape role definitions.
 We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital.
@@
 \emph{Historical factors:} Previous events in the history of the organization shape its current clustering of expectations, in the same way that alterations early in the life of other organic systems shape their later structure.
-Some developers that have been part of the organization for a long time are not just developers: they are also mentors, historians, and political players, despite what the organizational chart says about them.
+Some developers that have been part of the organization for a long time are not just developers, they are also mentors, historians, and political players, despite what the organizational chart says about them.
 This affects the expectations that people place on them.
@@
 Many of the expectations placed on the head of the Operations group at Cobalt spawn from his structural position.
-He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from system administration concerns, and he is expected to control the gateway that determines whether a product is ready to be released or not, that is, to assess and have final say on the operational quality of the product.
+He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from system administration concerns, and he is expected to control the gateway that determines whether a product is ready to be released or not---that is, to assess and have final say on the operational quality of the product.
 These expectations come from his position in the organization: it would be hard for someone without his position to satisfy them.
@@
 \emph{Personal factors:} A person's background, preferences, and personality shapes the expectations she receives from others.
-Perhaps his background is deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from him.
-Alternatively, his traits (curiosity, proactivity, a talent) may lead him to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors.
+Perhaps her background is deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from her.
+Alternatively, his traits (curiosity, proactivity, a talent) may lead her to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors.
 For example, the Project Administrator at Cobalt was hired to perform clerical tasks, but his own preference led him to gradually take more and greater responsibilities.
 The organization learned to be dependent on such behaviour.
-When he temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement, and its projects suffered quickly and significantly, as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him.
-We should also note that at the time of our observations the Project Administrator position was undergoing an official redefinition in order to match real expectations better.
+When he temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement.
+Projects suffered quickly and significantly as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him.
+We should also note that at the time of our observations, the Project Administrator position was undergoing an official redefinition in order to match real expectations.
 In this way, personal factors were leading to historical and structural changes in the organization, so that future enactors of the role played by the Project Administrator would be affected by them.
@@
 We found a divergence between people's positions and the expectations placed on them.
-There is an overlap, however, and it is worth exploring it first.
+There is an overlap, however, and it is worth exploring first.
 One can reasonably expect a Developer to write code of some kind, a QA Analyst to be concerned with software quality in one way or another, and a Team Lead to monitor and report on the team's progress.
@@
 An example out of several in our observations comes from one of the Gallium Product Managers.
-He was hired as a Product Manager to steer and coordinate the development of the company's products, but he had no experience in software development, thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization he struggled to understand the level of control and direction he was expected to have.
-Over time he learned to fill the void of control that he was expected to fill, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of his peers with the same or similar positions.
+He was hired as a Product Manager to steer and coordinate the development of the company's products, but he had no experience in software development, thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization, he struggled to understand the level of control and direction he was expected to have.
+Over time, he learned to fill the void of control that he was expected to fill, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of his peers with the same or similar positions.
 
@@
 
-We proposed that roles evolve for two reasons: organizational growth and mismatches between expectation and performance. Our data provide support for both reasons.
+We proposed that roles evolve for two reasons: organizational growth, and mismatches between expectation and performance. Our data provide support for both reasons.
 
@@
 The smallest organization is formed by a single person, and that person should perform all the activities required for the functioning of the organization, from sales to debugging.
-As the organization gets more members, labour begins to be divided, not by amount, but by kind, so that upon reaching even a handful of members it is rare for a software group to perform undifferentiated labour.
-Furthermore, growth carries a coordination weight: larger organizations need to assign some of its members the role of coordinating the work of others, or of interfacing between groups.
-That is, larger organizations have needs that are caused by their size, and that smaller organizations do not have.
-With organizations in the range we observed, some degree of specialization is well underway.
+As the organization obtains more members, labour begins to be divided, not by amount, but by kind, so that upon reaching even a handful of members, it is rare for a software group to perform undifferentiated labour.
+Furthermore, growth carries a coordination weight: larger organizations need to assign some of their members the role of coordinating the work of others, or of interfacing between groups.
+That is, larger organizations have needs that are caused by their size---needs that smaller organizations do not have.
+With organizations that are similar to those we observed, some degree of specialization is well underway.
 We expect that, should we perform this study in a larger organization, we would discover further role specialization.
@@
 Growth-driven role specialization can be resisted and overturned, as some Agile proposals suggest.
-In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation in the organization.
+In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} the role differentiation in the organization.
 The silos that group members wanted to break were project-based silos.
@@
 Similarly, an assignment that does \emph{not} work in theory may work in practice, due to the initiative or motivation of some people who contribute beyond their expectations.
-However, in both the under- and the over-performing case, with time organization members grow to expect the typical performance of others, and a new expectation (and hence, a new role) takes hold.
+However, in both the under- and the over-performing case, with time, organization members grow to expect the typical performance of others, and a new expectation (and hence, a new role) takes hold.
 
-Several examples from our data above illustrate this phenomenon; we offer one more here.
-The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced, and in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being one of the main contacts for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction and next steps.
+Several examples from our data above illustrate this phenomenon and we offer one more here.
+The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being one of the main contacts for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction and next steps.
 In many ways, his role begun to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization.
@@
 Customer Support staff used Google Documents to keep track of some of its cases, and to provide an ongoing documentation of recent events and requests for support.
-Email and phone were more heavily used in the management side of the organization than on its technical core: developers and QA analysts rarely made or received phone calls, and they obtained most of the information they needed about their projects through the issue tracking system.
+Email and phone were more heavily used in the management side of the organization than in its technical core: developers and QA analysts rarely made or received phone calls, and they obtained most of the information they needed about their projects through the issue tracking system.
 
@@
 \section{Discussion}
-
-\subsection{Summary}
 
@@
 
-We also explored role evolution and found that it is driven by organizational growth (which leads to specialization) and by mismatches between expectations and behaviour (which lead to an adjustment of expectations).
+We also explored role evolution and found that it is driven by organizational growth (which leads to specialization), and by mismatches between expectations and behaviour (which lead to an adjustment of expectations).
 Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites in terms of inter-role communications, with the qualification that issue-tracking systems were used differently by people enacting different roles.
@@
 
-Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation clusters (that is, the expectations that make good sense, organizationally, to be placed on the same individual), which should lead to better role definitions, and for the study of bottlenecks and deficiencies in organizational structure.
+Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation clusters (that is, the expectations that make good sense, organizationally, to be placed on the same individual).
+This in turn should lead to better role definitions, and to the study of bottlenecks and deficiencies in organizational structure.
 For instance, the discussion of the proper ratio of developers to testers, which is moot for teams where there is no QA and developers do all the testing, can be now explored in terms of the proper ratio of people expected to write production code to people expected to write integration tests or to determine the level of quality of the product.
@@
 Depending on the research questions they address, this may be a serious concern.
-Mining issue tracking systems, judging by our two cases, should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate, and focus only on those who emit messages in the system.
+Judging by our two cases, mining issue tracking systems should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate, and focus only on those who emit messages in the system.
 
@@
 It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that their formal position does not match the expectations placed on them.
-In these cases a reductionism to position labels is harmful.
-
-% TODO: Anything else?
+In these cases, a reductionism to position labels is harmful.
 
@@
 
-During our observations, Gallium underwent some role transitions, as it struggled to rearrange itself after a round of layoffs.
-We kept in mind the unusual state of this organization in our data analysis, and we believe that it contributed to our understanding of role evolution and mismatches in expectation assignment.
+During our observations, Gallium underwent some role transitions as it struggled to rearrange itself after a round of layoffs.
+We kept in mind the unstable state of this organization in our data analysis, and we believe that it contributed to our understanding of role evolution and mismatches in expectation assignment.
 
@@
 We acknowledge this threat, but we see it as a necessary evil in dealing with this subject.
-We hope that with more research in the area a consensus will emerge on the right level of analysis for expectations in software development work.
+We hope that with more research in the area, a consensus will emerge on the right level of analysis for expectations in software development work.
 
@@
 
-We thank the participants of our study, who generously opened their doors to us and gave us their time. We also thank Whoever Helps Us With Comments and Proofreading.
+We thank the participants of our study for their time, and Cassandra Petrachenko and Jamie Starke for their comments.
 

### 2011-09-28 soulgem.tex
Third draft complete.

@@
 However, it appears that some patterns in role definitions would emerge with further studies.
-We can see the beginnings of such patterns in Table \ref{tableExpectations}, which shows the expectations placed on the people we interviewed (about half of the total number of organization members).
+We can see the beginnings of such patterns in Table \ref{tableExpectations}, which shows the expectations placed on the people we interviewed.
 The columns in the table represent our interviewees and the rows represent common or significant expectations.
@@
 Work on this area was reserved almost exclusively for developers and their technical and team leads.
-However, the leads' expectations were more spread out: technical leads focused more on architecture and design issues, whereas team leads emphasized coordination activities.
+However, the expectations placed on leads were more spread out: technical leads focused more on architecture and design issues, whereas team leads emphasized coordination activities.
 There were also purely organizational roles that do not need any domain or technical expertise.
@@
 
-\emph{Product Management at Gallium:} There are three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers: the two Product Managers and the Product Design Lead.
-The latter documents the requirements of new functionality, he fulfills a ``historian'' expectation (sharing design rationale with developers and pointing out pitfalls in some approaches), and he is active in shaping the organizational structure, having helped create some groups and practices.
-Of the two Product Managers, one has an inside-looking approach, whereas the other one has an outside-looking approach: he has a Ph.D.\ in the organization's area of expertise, and his presence at client meetings, trade shows, and conferences legitimizes the organization.
-He also understands how the users would interact with the product and helps shape the user interface to bring it closer to the users' mental model.
+\emph{Product Management at Gallium:} There were three people with the role of steering the organization's products, prioritizing its backlog, and coordinating with customers: the two Product Managers and the Product Design Lead.
+The latter documented the requirements of new functionality, provided a historical perspective on design decisions, and helped shape the organizational structure.
+Of the two Product Managers, one had an inside-looking approach, whereas the other one had an outside-looking approach: he had a Ph.D.\ in the organization's area of expertise, and his presence at client meetings, trade shows, and conferences legitimized the organization.
+He also understood how the users would interact with the product and helped design the user interface to bring it closer to the users' mental model.
 Other people in the organization also performed some Product Management activities unofficially: Customer Services Representatives and Field Specialists often persuade developers to focus on the features that cause them the most pain, for example.
@@
 
-\emph{Team Leading at Gallium and Cobalt:} Every team in the sites we studied had a Team Lead, but this leadership role was tinged by the characteristics of the team and project in question.
-For instance, in Cobalt, team membership is fluid---every developer may be expected to work on any project, according to need---and some Team Leads performed coordination tasks, others had a dormant leadership position, waiting for new issues or concerns in their projects to arise, and yet others were more like ``first responders'' than coordinators, requesting assistance only if the work load for their projects became too much for themselves.
+\emph{Team Leading at Gallium and Cobalt:} Every team in the sites we studied had a Team Lead, but this leadership role was tinged by the characteristics of the team in question.
+In Cobalt, team membership is fluid---every developer may be expected to work on any project, according to need---and some Team Leads performed coordination tasks, others had a dormant leadership position, waiting for new issues to arise, and yet others were more like ``first responders'' than coordinators, requesting assistance only if their work load became excessive.
 In Gallium, there were three Team Leads at the time of our observations.
-They shared a few characteristics: they were expected to develop software, and they reported to their Director.
-But their roles were quite different.
-The Team Lead of the Alpha Team would spend about half of his time developing software, and in his other half, he meets with customer service representatives, project and product managers, performs estimation tasks, and analyzes the performance of his team.
-A second Team Lead is in charge of a product about to be discontinued, and that responsibility takes a small part of his time.
-He is also associated with the Beta Team, but since that team has an abundance of ``leads'' (it includes the two official Tech Leads of the organization), his coordination and reporting expectations are minimal.
-The third Team Lead has one other team member, a newcomer to the organization, and his leadership expectations consist mostly of mentoring and training rather than coordination and performance analysis.
+Their roles were quite different.
+The Team Lead of the Alpha Team would spend about half of his time developing software, and the other half in meetings, performing estimation tasks, and analyzing the performance of his team.
+A second Team Lead was in charge of a product about to be discontinued, but that responsibility took a small part of his time.
+He was mostly associated with the Beta Team, but since that team had an abundance of ``leads'' (it included the two official Tech Leads of the organization), his coordination and reporting expectations were minimal.
+The third Team Lead had one other team member, a newcomer to the organization, and his leadership expectations consisted mostly of mentoring and training rather than coordination and monitoring.
 
 \emph{Developers at Gallium and Cobalt:} The quintessential role in the software development literature was one of the hardest to pin down.
-A developer, of course, is expected to write production code, but what else?
-There is a multitude of answers to that question: some developers are expected to perform code reviews, some are expected to provide architectural input, some are expected to carry their own requirements analysis, some are expected to socialize, some are expected to interface with customers.
+A developer, of course, is expected to write code, but what else?
+There is a multitude of answers to that question: some developers are expected to perform code reviews, some to provide architectural input, some to carry their own requirements analysis, some to socialize, some to interface with customers.
 None of the developers in our sample were expected to perform all of the above.
 
-\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization.\footnote{Note that, as opposed to the previous examples, this role has no clear label to refer to it.} But the stuff that falls through the cracks was different in both sites, and therefore, the expectations were different.
+\emph{``Glue'' role, Gallium and Cobalt:} In both sites we found at least one person who others referred to, with more or less these words, as ``the glue that holds this place together.'' They generally meant a person that would do all sorts of things, little or large, in a variety of contexts, stuff that falls through the cracks of the formal organization.
+But the stuff that falls through the cracks was different in both sites, and therefore, the expectations were different.
 In Cobalt, much of this referred to sheltering the organization from bureaucratic processes and constraints.
@@
 
-\emph{Project Management at Gallium:} These Project Managers are a counterexample to our findings, to some extent.
-Although there are distinctions between them (one has a Project Management Leader role, another creates templates and institutes new processes, and all, including a third and more junior Project Manager, implement them), the similarities in their roles seem to outweigh the differences.
-The main reason seems to be that these Project Managers bring a fairly specific and standardized way of doing things with them, in the form of a Project Management Body of Knowledge and certification.
-In other words, the institutionalization of their role is strong.
-Part of their strategy appears to be to define their role with sufficient precision, and to enforce their definition in the rest of the organization.
+\emph{Project Management at Gallium:} These Project Managers were a counterexample to our findings, to some extent.
+Although there were distinctions between them (one had a Project Management Leader role, another created templates and instituted new processes, and all, including a third and more junior Project Manager, implemented them), the similarities in their roles seemed to outweigh the differences.
+The main reason seemed to be that they brought a fairly specific and standardized way of doing things with them, in the form of a Project Management Body of Knowledge and certification.
+In other words, the institutionalization of their role was strong.
+They also strived to define their role with precision, and to enforce their definition in the rest of the organization.
 No other role that we examined had the emphasis on role definition and consistency that the Gallium Project Managers had.
@@
 
-We found that expectations are clustered in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors, and that these kinds of factors can interplay and provide feedback to each other.
+We found that expectations are grouped in roles based on a combination of \emph{institutional}, \emph{historical}, \emph{structural}, and \emph{personal} factors, and that these factors interplay and provide feedback to each other.
 
@@
 We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital.
-It quickly expanded and distributed its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them up with newcomers with the goal of taking care of those aspect of the business.
-Before this influx, the division of labour was more flexible and organic: most organization members did a bit of everything.
-After the influx, role definitions shifted towards a more stereotypical differentiation of roles and groups.
+It quickly expanded and distributed its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them with newcomers.
+Before this influx, the division of labour was flexible and organic: most organization members did a bit of everything.
+After the influx, role definitions shifted towards a more stereotypical differentiation.
 
-\emph{Historical factors:} Previous events in the history of the organization shape its current clustering of expectations, in the same way that alterations early in the life of other organic systems shape their later structure.
+\emph{Historical factors:} Previous events in the history of the organization shape its current grouping of expectations, in the same way that alterations early in the life of other organic systems shape their later structure.
 Some developers that have been part of the organization for a long time are not just developers, they are also mentors, historians, and political players, despite what the organizational chart says about them.
-This affects the expectations that people place on them.
+This affects the expectations placed on them.
 During the Gallium downsizing, some people were demoted to keep the company afloat.
 At some point, one of them had been the head of the QA department (which was dispersed) and, later, the Program Manager.
-At the time of our observations he was a QA Analyst, yet he was still involved (and expected to be involved) in many strategic activities that do not match the institutional view of what a QA Analyst is supposed to do.
+At the time of our observations he was a QA Analyst, yet he was still expected to be involved in strategic activities that did not match the institutional view of what a QA Analyst is supposed to do.
 
 \emph{Structural factors:} A person's position in the organizational structure gives her a certain power and awareness that others may not have.
-Her structural position leads others to expect from her certain behaviours: she may be expected to play a bridging, directing, or even confrontational role.
-Many of the expectations placed on the head of the Operations group at Cobalt spawn from his structural position.
-He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from system administration concerns, and he is expected to control the gateway that determines whether a product is ready to be released or not---that is, to assess and have final say on the operational quality of the product.
-These expectations come from his position in the organization: it would be hard for someone without his position to satisfy them.
+Her structural position leads others to expect from her certain behaviours: to play a bridging, directing, or even confrontational role.
+Many of the expectations placed on the head of the Operations group at Cobalt spawned from his structural position.
+He was expected to provide news of the operational status of Cobalt's services, to deal with and shelter the Software group from system administration concerns, and to control the gateway that determines whether a product is ready to be released or not---that is, to have final say on the operational quality of the product.
+These expectations come from his position in the organization.
+It would be hard for someone in a different position to satisfy them.
 
 \emph{Personal factors:} A person's background, preferences, and personality shapes the expectations she receives from others.
-Perhaps her background is deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from her.
-Alternatively, his traits (curiosity, proactivity, a talent) may lead her to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors.
-For example, the Project Administrator at Cobalt was hired to perform clerical tasks, but his own preference led him to gradually take more and greater responsibilities.
+Perhaps her background is deficient for the fulfillment of her initial role, and others adapt by expecting different or lesser things from her.
+Alternatively, her traits (curiosity, proactivity, a talent) may lead her to take on responsibilities that would not otherwise be placed on her given institutional, historical, or structural factors.
+For example, the Project Administrator at Cobalt was hired to perform clerical tasks, but his own initiative led him to gradually take more and greater responsibilities.
 The organization learned to be dependent on such behaviour.
 When he temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement.
-Projects suffered quickly and significantly as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him.
+Projects suffered quickly and significantly as the replacement only focused on the expectations that were officially placed on him.
 We should also note that at the time of our observations, the Project Administrator position was undergoing an official redefinition in order to match real expectations.
-In this way, personal factors were leading to historical and structural changes in the organization, so that future enactors of the role played by the Project Administrator would be affected by them.
+In this way, personal factors were leading to historical and structural changes, so that future enactors of the role played by the Project Administrator would be affected by them.
 
@@
 As we have seen above, the expectations placed on a person also come from historical and personal factors.
-An example out of several in our observations comes from one of the Gallium Product Managers.
-He was hired as a Product Manager to steer and coordinate the development of the company's products, but he had no experience in software development, thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization, he struggled to understand the level of control and direction he was expected to have.
-Over time, he learned to fill the void of control that he was expected to fill, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of his peers with the same or similar positions.
+An example from our observations is one of the Gallium Product Managers.
+He was hired to steer and coordinate the development of the company's products, but he had no experience in software development.
+He thought that his main asset was his understanding of and his contacts in the products' domain, so that he was in the organization as the ``voice of the customer.'' In his early days in the organization, he struggled to understand the level of control he was expected to have.
+Over time, he adjusted his behaviour, but he continued to make use of his background, connections, and domain intuition to give his role a distinctly different flavour than those of others with similar positions.
 
@@
 
-Regarding organizational growth, there is a long sociological tradition exploring its role in specialization and division of labour, and our data are far from surprising.
-In fact, it should be clear from pure argument that organizational growth drives role evolution.
+Regarding organizational growth, it has been recognized as a factor in specialization for a long time, and our data are far from surprising.
+In fact, it should be clear from pure argument that growth drives role evolution.
 The smallest organization is formed by a single person, and that person should perform all the activities required for the functioning of the organization, from sales to debugging.
-As the organization obtains more members, labour begins to be divided, not by amount, but by kind, so that upon reaching even a handful of members, it is rare for a software group to perform undifferentiated labour.
+As the organization obtains more members, labour begins to be divided, so that upon reaching even a handful of members, it is rare for a software group to perform undifferentiated labour.
 Furthermore, growth carries a coordination weight: larger organizations need to assign some of their members the role of coordinating the work of others, or of interfacing between groups.
-That is, larger organizations have needs that are caused by their size---needs that smaller organizations do not have.
+That is, larger organizations have needs that are caused by their size.
 With organizations that are similar to those we observed, some degree of specialization is well underway.
@@
 Growth-driven role specialization can be resisted and overturned, as some Agile proposals suggest.
-In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} the role differentiation in the organization.
-The silos that group members wanted to break were project-based silos.
+In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation.
+The silos that the group wanted to break were project-based.
 They discovered that assigning one person per project introduced vulnerabilities (when something went wrong in a project, only one person could intervene) and rework (as many projects consisted of similar modules that had to be developed independently).
 To break these silos, Cobalt assigned five of its senior developers the position of Subsystem Leads, so that each of them would be the first point of contact for matters concerning their subsystem.
-The end result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a perceived decrease on the level of isolation, as projects begun to cross-pollinate and Subsystem Leads took care to help their teammates understand the particularities of their subsystems.
+The end result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a perceived decrease on the level of isolation, as projects begun to cross-pollinate.
 
-The other important factor driving role evolution in our findings was a mismatch between expectations and behaviour.
+The other important factor driving role evolution was a mismatch between expectations and behaviour.
 The fact that expectations are placed does not mean that they are satisfied.
@@
 Therefore, an assignment of expectations that works in theory may not work in practice.
-Similarly, an assignment that does \emph{not} work in theory may work in practice, due to the initiative or motivation of some people who contribute beyond their expectations.
-However, in both the under- and the over-performing case, with time, organization members grow to expect the typical performance of others, and a new expectation (and hence, a new role) takes hold.
+Similarly, an assignment that does \emph{not} work in theory may work in practice, due to the initiative or motivation of people who contribute beyond their expectations.
+However, in both cases, with time, organization members grow to expect the typical performance of others, and a new expectation (and hence, a new role) takes hold.
 
 Several examples from our data above illustrate this phenomenon and we offer one more here.
-The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being one of the main contacts for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction and next steps.
+The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being a contact for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction.
 In many ways, his role begun to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization.
@@
 We expected to find that people enacting similar roles use similar communication media, and that they communicate with people enacting interfacing roles through other media.
-Although we found support for this claim, as we explain below, a biasing factor was the prevalence of face-to-face interactions in both of our field sites.
-People with similar or different roles communicate face-to-face regularly.
+Although we found support for this claim, a biasing factor was the prevalence of face-to-face interactions in both of our field sites.
+Regardless of role, people communicated face-to-face regularly.
 Stand-up meetings, walks across the hall, and group meetings were commonplace.
@@
 People with expectations regarding code production, quality assurance, customer service, field implementation, product management, and performance monitoring all used the issue tracking system to some extent.
-Its use was not uniform: some group members used it to emit messages, others were mostly listeners, and others were more interested in the aggregate than in the contents of particular electronic traces.
+Its use was not uniform: some group members used it to emit messages, others were mostly listeners, or they were more interested in the aggregate activity of the team.
 This finding mirrors the observation of De Souza \emph{et al.}\ regarding issue tracking systems as boundary objects \cite{deSouza2003}.
@@
 In Gallium, IRC use was common among some developer teams: each team had one channel, and these were not frequented by any non-developers.
-Customer Support staff used Google Documents to keep track of some of its cases, and to provide an ongoing documentation of recent events and requests for support.
-Email and phone were more heavily used in the management side of the organization than in its technical core: developers and QA analysts rarely made or received phone calls, and they obtained most of the information they needed about their projects through the issue tracking system.
+Customer Support staff used Google Documents to keep track of its cases, and to provide an ongoing documentation of recent events and requests for support.
+Email and phone were more heavily used in the management side of the organization: developers and QA analysts rarely made or received phone calls, and they obtained most of the information they needed about their projects through the issue tracking system.
 
@@
 
-We also explored role evolution and found that it is driven by organizational growth (which leads to specialization), and by mismatches between expectations and behaviour (which lead to an adjustment of expectations).
-Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites in terms of inter-role communications, with the qualification that issue-tracking systems were used differently by people enacting different roles.
+We also explored role evolution and found that it is driven by organizational growth, and by mismatches between expectations and behaviour.
+Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites across roles.
 
@@
 A stereotypical developer, team lead, or tester, is rarely found in practice.
-Instead, thinking in terms of tasks and the people who are expected to perform them will help us reattach our studies and findings to software development in the real world.
+Instead, thinking in terms of tasks and expectation groupings will help us ground our studies to software development in the real world.
 
-Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation clusters (that is, the expectations that make good sense, organizationally, to be placed on the same individual).
+Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation pairings (expectations that make good sense, organizationally, to be placed on the same individual).
 This in turn should lead to better role definitions, and to the study of bottlenecks and deficiencies in organizational structure.
-For instance, the discussion of the proper ratio of developers to testers, which is moot for teams where there is no QA and developers do all the testing, can be now explored in terms of the proper ratio of people expected to write production code to people expected to write integration tests or to determine the level of quality of the product.
+For instance, the discussion of the proper ratio of developers to testers, which is moot for teams where there is no QA and developers do all the testing, can be explored in terms of the proper ratio of people expected to write code to people expected to write integration tests or to determine the level of quality of the product.
 
-Third, the observation that media use varies depending on the kinds of expectations placed on the user complements and helps explain the earlier finding that electronic repositories are incomplete and often misleading \cite{Aranda2009}.
-If researchers focus on a single electronic repository, they are likely to miss systematically important areas of organizational coordination: they will be blind to some people and expectations.
-Depending on the research questions they address, this may be a serious concern.
-Judging by our two cases, mining issue tracking systems should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate, and focus only on those who emit messages in the system.
+Third, the observation that media use varies depending on the kinds of expectations placed on the user complements the earlier finding that repositories are incomplete and often misleading \cite{Aranda2009}.
+If researchers focus on a single repository, they are likely to systematically miss important areas of coordination: they will be blind to some people and expectations.
+Depending on their research questions, this may be a serious concern.
+Judging by our two cases, mining issue tracking systems should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate activity.
 
-\emph{Practice:} Our qualitative data confirm, in the software development domain, the old sociological finding that the role specialization trend as an organization grows is hard to fight.
+\emph{Practice:} Our data confirm, in the software development domain, the old sociological finding that the role specialization trend as an organization grows is hard to fight.
 This is not necessarily bad, but teams of would-be generalists (which Agile techniques often propose) need to calculate the long-term impacts of seemingly necessary short-term specializations.
@@
 
-Additionally, we found that many practitioners do not have an awareness of the full scope of their peers' activities, and that this lack of awareness leads them to misunderstand the benefits that they bring to the organization.
+Additionally, we found that practitioners may not have an awareness of the full scope of their peers' activities, and that this lack of awareness leads them to misunderstand the benefits that they bring to the organization if they make judgments based on labels.
 It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that their formal position does not match the expectations placed on them.
@@
 However, most of our findings are supported by a theoretical framework that is not built upon these particular characteristics, but is instead general and interdisciplinary.
-We expect that some of the particular expectation clusters will vary in other groups, depending on their contextual factors, but we also expect the underlying rationale and observations to hold across them.
+We expect that some of the particular expectation pairings will vary in other groups, but we also expect the underlying rationale and observations to hold across them.
 
@@
 Throughout this paper, we defined role as a ``set of expectations.'' However, an ``expectation'' is a subjective concept: we expect many things from many people, and an objective articulation of these expectations may never be quite right, missing some attributes or formulating an expectation at the wrong level of detail.
-We acknowledge this threat, but we see it as a necessary evil in dealing with this subject.
-We hope that with more research in the area, a consensus will emerge on the right level of analysis for expectations in software development work.
+We acknowledge this threat, but it is a necessary evil.
+We hope that with more research in the area, a consensus will emerge on the right level of analysis for expectations in software organizations.
 

### 2011-09-29 soulgem.tex
Fourth draft complete. Ready for submission.

@@
 \begin{document}
-\title{Setting the Stage for Understanding the Main Roles in Software Organizations}
+\title{Setting the Stage to Understand the Roles in Software Organizations:\\
+A Framework and a Case Study}
 
@@
 
-Software professionals divide their work by roles, but to date, researchers have not examined nor questioned the nature of these roles.
+Software professionals divide their work by roles, but to date, researchers have not examined nor questioned the nature of these roles, which is a potential source of misunderstandings and confusion.
 To address this, we formulate a theoretical framework based on sociological and organizational research, in which roles are defined as sets of expectations placed on the members of the software organization.
 
-The paper also reports on an empirical case study of two software organizations, which was informed and inspired by this theoretical framework.
-In our study, we found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors.
+The paper also reports on an empirical case study of two software organizations, which was informed and inspired by this framework.
+We found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors.
 Furthermore, role definitions evolve through organizational growth and through mismatches between expectations and performance, and people enacting different roles vary in their selection of communication media.
@@
 In some domains, especially ones that are more mature or critical, the division of labour is fairly formal and clear.
-In the medical field, surgeons, anesthesiologists, and nurses generally know what to expect from each other, in and out of the operating room.
+In the medical field, surgeons, anesthesiologists, and nurses generally know what to expect from each other.
 Knowing that someone is a nurse will bring out a set of behavioural expectations that have a good probability of being shared across the board.
@@
 Researchers may prescribe that practices be carried out by one role without realizing that the role is usually shared and spread out among a group of people.
-Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him \cite{Dagenais2010}.
+Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train her \cite{Dagenais2010}.
 Their communication tools may fail to be adopted if they satisfy the needs of some roles but not of others.
-For an abstraction as heavily used as roles in software organizations, a closer look should prove beneficial.
+For an abstraction as heavily used as this, a closer look should prove beneficial.
 
@@
 A role ``is made up of all those norms which are thought to apply to a person occupying a given position.'' Scott \cite{Scott2008} concurs, pointing out that roles are normative expectations of the appropriate goals and activities for particular individuals, and that roles can be formally constructed or emerge informally through interaction.
-
 A good analysis of role evolution comes from Nicholson \cite{Nicholson1984}.
-He studied how people adapt to new work roles, and determined there are four main ways in which they do so, based on how much they change themselves and the environment: replication (neither the person nor the environment change much), absorption (personal development), determination (role development), and exploration (the person develops both herself and the meaning of her role).
+He studied how people adapt to new work roles, and determined there are four main strategies in which they do so, based on how much they change themselves and the environment: replication, absorption, determination, and exploration.
+He also discusses the factors that lead people to use one of these strategies over the others.
 
@@
 Process definitions tend to include role prescriptions: Scrum \cite{Schwaber2001}, for instance, calls for a Scrum Master, a Product Owner, and a number of developers.
-Practices also prescribe roles, but in a more minute detail and, usually, for a shorter term.
-Thus in a Fagan inspection \cite{Fagan1976}, we have a Moderator, a Reader, a Coder, and several Reviewers.
+Practices also prescribe roles, but in greater detail and, usually, for a shorter term.
+Thus in a Fagan inspection \cite{Fagan1976}, we have a moderator, a reader, a coder, and several reviewers.
 
@@
 Hoda \emph{et al.}\ \cite{Hoda2010} identified six ``informal roles'' that team members adopt in self-organizing teams.
-The authors distinguish between roles and the people that enact them, which is a distinction we advocate for in this paper.
-But several of the roles they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of single expectations in our framework below---expectations that are bundled with others to produce a richer, multi-faceted role.
+The authors rightly distinguish between roles and the people that enact them, but several of the roles they identify (such as Champion and Promoter) are too fine-grained, and they fall under the category of single expectations in our framework below---expectations that are bundled with others to produce a richer, multi-faceted role.
 
 Moe \emph{et al.} \cite{Moe2010} use Dickinson and McIntyre's \cite{Dickinson1997} teamwork model to provide such a set of expectations for members of self-managing teams.
-This is certainly useful, although it cannot be used as a basis for our work, as it does not allow us to study how or why differences in expectations and specializations within teams arise.
+This is certainly useful, although not as a basis for our work, as it does not allow us to study how or why differences in expectations and specializations within teams arise.
 
@@
 Their structural analysis allows them to scale to large projects and electronic repositories.
-To our knowledge, more sophisticated analyses of roles from a structural role theory perspective, such as the sociological blockmodelling work of Breiger and Pattison \cite{Breiger1986}, has not been carried out in our domain.
+To our knowledge, more sophisticated analyses of roles from a structural role theory perspective, such as the sociological blockmodelling work of Breiger and Pattison \cite{Breiger1986}, have not been carried out in our domain.
 
@@
 F{\ae}gri \emph{et al.} \cite{Faegri2010} provide a comprehensive report of the consequences of role evolution in a case in which developers attempted to rotate the responsibilities of customer support, and Jensen and Scacchi \cite{Jensen2007} describe the migrations in role responsibilities in several open source projects.
-De Souza \emph{et al.}\ \cite{deSouza2003} look at inter-role communication, and describe that problem reports are used as boundary objects \cite{Bowker1999} between people with different roles, while Marczak and Damian \cite{Marczak2011} explore the backchannel communication between people with different roles.
+De Souza \emph{et al.}\ \cite{deSouza2003} look at inter-role communication, and describe that problem reports are used as boundary objects \cite{Bowker1999} between people with different roles, while Marczak and Damian \cite{Marczak2011} explore the relevance of backchannel communications between people with different roles.
 
 Finally, Litecky \emph{et al.}\ \cite{Litecky2010} report on the positions offered and the skills required in the computing job market.
-Their analysis is useful for a study of roles in software organizations as it provides us with an indication of the roles that human resources professionals consider important in their groups.
+Their analysis is useful for a study of roles in software organizations as it provides us with an indication of the roles that managers consider important in their groups.
 
@@
 A label helps us convey the essence of the role, but some roles have ill-fitting labels, and others have no labels or easy ways to refer to them in concrete terms.
-We must remember that neither the label nor the formal position completely define the role.
+Neither the label nor the formal position completely define the role.
 
@@
 Culture plays an important part in setting and steering our expectations.
-In some domains, cultural norms are strong enough to establish an interpretation of a role throughout the population (the role has been institutionalized \cite{DiMaggio1991}).
+In some domains, cultural norms are strong enough to establish an interpretation of a role throughout the populatio---the role has been institutionalized \cite{DiMaggio1991}.
 In these cases, a combination of intensive training, ethical codes, professional societies, and familiarization lead us to expect certain behaviours and activities from a role.
@@
 
-Group members do not always act according to the expectations placed on them, for several reasons.
-Their performance could fail to meet expectations because of unrealistic expectations, unskilled, careless, or unaware performers, or conflicting expectations.
-Their performance could also exceed expectations due to the performer's initiative, talent, or background.
+Group members do not always act according to the expectations placed on them.
+Their performance could fail to meet expectations because of unrealistic or conflicting expectations, or unskilled, careless, or unaware performers.
+It could also exceed expectations due to the performer's initiative, talent, or background.
 In any case, if disagreements recur over time, organization members undergo a shift in their expectations: they learn to expect the ``new normal''.
@@
 \emph{Proposition:} People enacting similar roles use similar communication media.
-They communicate with people enacting different but interfacing roles through other media, and these media act as a boundary object between them \cite{Bowker1999}.
+They communicate with people enacting different but interfacing roles through other media, which act as boundary objects between them \cite{Bowker1999}.
 
@@
 The Team Leads functioned as the Scrum Masters, but the Product Owner was not defined consistently.
-The group also deviated from the simple backlogs and story card practices: their requirements were often specified in greater detail and they were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support.
+The group also deviated from the simple backlogs and story card practices: their requirements were often specified in detail and were derived from contractual obligations acquired by the sales team or from urgent issues identified by Field Specialists or Customer Support.
 
@@
 However, they had a significant say in the direction that the Software group took, and some of them developed script code.
-Finally, there are two staff employees in the group that do not form part of any team in particular: the Group Leader, and a Project Administrator.
+Finally, there were two staff employees that did not form part of any team in particular: the Group Leader, and a Project Administrator.
 
@@
 Our observations at Cobalt happened during a period of few changes or upsets.
-There had been two major restructures of the group (first, when the Software, Operations, and Science divisions were created; later, when the Software group created the Team Lead and Subsystem Lead structure), but they had both taken place years before our visit.
+There had been two major restructures of the group (first, when the Software, Operations, and Science divisions were created; later, when the Software group created the Subsystem Lead structure), but they had both taken place years before our visit.
 
@@
 We present our findings organized by the research question they address.
-
-\input{expectationPeopleTable}
 
@@
 However, the expectations placed on leads were more spread out: technical leads focused more on architecture and design issues, whereas team leads emphasized coordination activities.
-There were also purely organizational roles that do not need any domain or technical expertise.
+There were also purely organizational roles that did not need any domain or technical expertise.
 But in both organizations some key people had more than one kind of expertise (technical, domain, and organizational), and these people seemed vital for the functioning of the organization as they transitioned between boundaries with ease.
@@
 
+\input{expectationPeopleTable}
+
 \subsection{Role Variation (RQ2)}
@@
 He also understood how the users would interact with the product and helped design the user interface to bring it closer to the users' mental model.
-Other people in the organization also performed some Product Management activities unofficially: Customer Services Representatives and Field Specialists often persuade developers to focus on the features that cause them the most pain, for example.
+Other people in the organization also performed some Product Management activities unofficially: Customer Services Representatives and Field Specialists often persuaded developers to focus on the features that caused them the most pain, for example.
 In sum, although the three employees above have similar goals, the expectations placed on them vary considerably, are not interchangeable, and are not their exclusive prerogative.
@@
 A developer, of course, is expected to write code, but what else?
-There is a multitude of answers to that question: some developers are expected to perform code reviews, some to provide architectural input, some to carry their own requirements analysis, some to socialize, some to interface with customers.
+There is a multitude of answers to that question: some developers are expected to perform code reviews, some to provide architectural input, some to socialize, some to explain legacy code.
 None of the developers in our sample were expected to perform all of the above.
@@
 He was expected to provide news of the operational status of Cobalt's services, to deal with and shelter the Software group from system administration concerns, and to control the gateway that determines whether a product is ready to be released or not---that is, to have final say on the operational quality of the product.
-These expectations come from his position in the organization.
-It would be hard for someone in a different position to satisfy them.
+These expectations came from his position in the organization.
+It would not be sensible to place them on someone in a different position.
 
@@
 We found a divergence between people's positions and the expectations placed on them.
-There is an overlap, however, and it is worth exploring first.
+There is an overlap, however, and it is worth exploring it first.
 One can reasonably expect a Developer to write code of some kind, a QA Analyst to be concerned with software quality in one way or another, and a Team Lead to monitor and report on the team's progress.
@@
 With organizations that are similar to those we observed, some degree of specialization is well underway.
-We expect that, should we perform this study in a larger organization, we would discover further role specialization.
+Should we perform this study in a larger organization, we expect to discover further role specialization.
 
-Growth-driven role specialization can be resisted and overturned, as some Agile proposals suggest.
-In this sense, however, the most interesting event in our observations was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation.
+Growth-driven specialization can be resisted and overturned, as some Agile proposals suggest.
+In this sense, however, the most interesting event in our study was an attempt to ``break silos'' at Cobalt, which was successful for its stated goals, but \emph{increased} role differentiation.
 The silos that the group wanted to break were project-based.
@@
 To break these silos, Cobalt assigned five of its senior developers the position of Subsystem Leads, so that each of them would be the first point of contact for matters concerning their subsystem.
-The end result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a perceived decrease on the level of isolation, as projects begun to cross-pollinate.
+The result was a greater differentiation between the roles of the five Subsystem Leads (previously simply Developers), but also a decrease on the level of isolation, as projects begun to cross-pollinate.
 
@@
 Several examples from our data above illustrate this phenomenon and we offer one more here.
-The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he begun taking other responsibilities: providing feedback on how users would use some service features, being a contact for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction.
-In many ways, his role begun to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization.
+The coordinator of the Science group at Cobalt was brought in to do research with the data that Cobalt serviced in a ``consultative role.'' As time progressed, and without being contractually obliged nor pressured by his peers, he took other responsibilities: providing feedback on how users would use some service features, being a contact for scientists outside the group (which lends more domain legitimacy to the organization), and helping to steer the group's direction.
+In many ways, his role started to look more like that of the Gallium Product Manager we discussed above, without being named or being perceived as a Product Manager by anybody in the organization.
 As a result, he now plays an essential role in the development of the group's products and in the maintenance of its relations with its users, a role that the rest of the organization depends on for its good functioning.
@@
 
+Knowledge on media use and adoption is particularly important in our field, given the opportunity to provide better tools and dynamics to practitioners.
 We expected to find that people enacting similar roles use similar communication media, and that they communicate with people enacting interfacing roles through other media.
@@
 Roles are socially constructed; part of the construction occurs at the cultural or institutional level, and part occurs at the local, team level, and is caused by historical, structural, and personal factors.
-Although we can make some reasonable assumptions about the work of a software professional based on his label or official position, our assumptions will only take us so far: their responsibilities vary considerably, and treating them as archetypical professionals (``developers,'' ``testers,'' and so on) dismisses a large part of what they bring or fail to bring to their job.
+Although we can make some reasonable assumptions about the work of software professionals based on their label or official position, our assumptions will only take us so far: their responsibilities vary considerably, and treating them as archetypical professionals (``developers,'' ``testers,'' and so on) dismisses a large part of what they bring or fail to bring to their job.
 
@@
 We also explored role evolution and found that it is driven by organizational growth, and by mismatches between expectations and behaviour.
-Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites across roles.
+Finally, we found that people enacting different roles tend to use different media to communicate, that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites across roles, and that there were relevant role-based differences in the use of communication media.
 
@@
 
-Throughout this paper, we defined role as a ``set of expectations.'' However, an ``expectation'' is a subjective concept: we expect many things from many people, and an objective articulation of these expectations may never be quite right, missing some attributes or formulating an expectation at the wrong level of detail.
+Throughout this paper, we defined role as a ``set of expectations.'' However, an ``expectation'' is a subjective concept: we expect many things from many people, and an objective articulation of these expectations may never be quite right, missing some attributes or focusing at the wrong level of detail.
 We acknowledge this threat, but it is a necessary evil.
@@
 
-We thank the participants of our study for their time, and Cassandra Petrachenko and Jamie Starke for their comments.
+We thank the participants of our study for their time, and Cassandra Petrachenko, Peter Rigby, and Jamie Starke for their comments.
 

### 2011-09-29 soulgem.tex
Tiny change to abstract. Submitted version.

@@
 
-Software professionals divide their work by roles, but to date, researchers have not examined nor questioned the nature of these roles, which is a potential source of misunderstandings and confusion.
+Software professionals divide their work by roles, but to date, researchers have not examined nor questioned the nature of these roles, which is a potential source of confusion.
 To address this, we formulate a theoretical framework based on sociological and organizational research, in which roles are defined as sets of expectations placed on the members of the software organization.

### 2012-01-10 icse-rebuttal.txt
Draft of rebuttal for ICSE reviewers.

We thank the reviewers for their thoughtful comments.

All three inquired on our interview protocol and format. Our interviews were semi-structured, and our study materials will be placed online.

The  Study Design section should describe our methods better: We carried out an explanatory case study, as per Yin. In such a study, researchers do not begin with a blank slate (in contrast to Grounded Theory). They take advantage of preexisting theories which, if correct, have consequences for the domain of study. These consequences are stated as propositions. Reviewer 3 considers that 'hypotheses' is a better term; our paper uses Yin's terms.   The interviews were transcribed by Aranda, and analyzed following Yin's approach: for each proposition we formulated linking logic statements in advance (descriptions of what data would support or refute the proposition, and how to obtain it). These statements guided our data collection. We will expand on our methodological description.

Reviewer 1 says that the theoretical framework is actually a clarification of concepts. We do clarify concepts, but we also link them to produce the propositions that form the basis for the study. We chose to label it a framework as it is not concrete enough to be a theory, but interconnected and stimulating enough to build research on top of it.

Reviewer 2 asked why the word "ethnography" was not used. Ethnographies tend to be more theory-neutral at start, whereas we drew from decades of sociological research to guide our research.

Reviewer 2 asks if the table represents expectations from the self, others, or a mix. It is a mix and we will clarify this in a revised version. As a follow-up, we are working on a finer distinction of expectations, including representing conflicts between self- and other-rated expectations, and on measuring role differentiation with cluster analysis, which is akin to Reviewer 2's suggestion to use the Hamming distance.

Reviewer 3 mentions our small sample size, short research period, and small organization size. We would like the reviewer to consider the nature of qualitative research. Two organizations would be an extremely small sample for an experiment, but a single informative case may be enough in case study research: we strive for generalization to a theory, not to a population (as per Yin's suggestion). Our research period allowed us to interview half of the workforce at both sites, and later interviews added little information. We expect larger organizations to offer further insights; we aim to report on them in the near future.

We wish our findings on the meaning of roles in industry will be of benefit to the community, and we thank the reviewers for their attention.

### 2012-01-11 icse-rebuttal.txt
Minimal changes to rebuttal to fit into the space required.

@@
 These consequences are stated as propositions.
-Reviewer 3 considers that 'hypotheses' is a better term; our paper uses Yin's terms.
+R3 considers that 'hypotheses' is a better term; our paper uses Yin's terms.
   The interviews were transcribed by Aranda, and analyzed following Yin's approach: for each proposition we formulated linking logic statements in advance (descriptions of what data would support or refute the proposition, and how to obtain it).
-These statements guided our data collection.
-We will expand on our methodological description.
+These
+statements guided our data collection. We will expand on our methodological description.
 
-Reviewer 1 says that the theoretical framework is actually a clarification of concepts.
+R1 says that the theoretical framework is actually a clarification of concepts.
 We do clarify concepts, but we also link them to produce the propositions that form the basis for the study.
@@
 
-Reviewer 2 asked why the word "ethnography" was not used. Ethnographies tend to be more theory-neutral at start, whereas we drew from decades of sociological research to guide our research.
+R2 asks why the word "ethnography" was not used. Ethnographies tend to be more theory-neutral at start, whereas we drew from decades of sociological research to guide our research.
 
-Reviewer 2 asks if the table represents expectations from the self, others, or a mix.
+R2 asks if the table represents expectations from the self, others, or a mix.
 It is a mix and we will clarify this in a revised version.
-As a follow-up, we are working on a finer distinction of expectations, including representing conflicts between self- and other-rated expectations, and on measuring role differentiation with cluster analysis, which is akin to Reviewer 2's suggestion to use the Hamming distance.
+As a follow-up, we are working on a finer distinction of expectations, including representing conflicts between self- and other-rated expectations, and on measuring role differentiation with cluster analysis, which is akin to R2's suggestion to use the Hamming distance.
 
-Reviewer 3 mentions our small sample size, short research period, and small organization size.
-We would like the reviewer to consider the nature of qualitative research.
+R3 mentions our small sample size, short research period, and small organization size.
+We would like to point out the nature of qualitative research.
 Two organizations would be an extremely small sample for an experiment, but a single informative case may be enough in case study research: we strive for generalization to a theory, not to a population (as per Yin's suggestion).
 Our research period allowed us to interview half of the workforce at both sites, and later interviews added little information.
-We expect larger organizations to offer further insights; we aim to report on them in the near future.
+Finally, we expect larger organizations to offer further insights; we aim to report on them in the near future.
 
